### 2018-11-12 ms/cover_letter.tex
add manuscript tex

Dear Editor,

We are pleased to submit our manuscript "UMAP reveals deep population structure and phenotype heterogeneity in large genomic cohorts" for publication in eLife. This work applies a newly developed dimensional reduction method called UMAP to visualize population structure in genomic datasets.

By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts. For example, we identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region, who are likely descendants of the Hispano population. We also identify multiple associations between genetics, geography, and phenotypes in the UK biobank and sub-continental population structure within 1000 Genomes Project populations.

This work has broad applicability. Our approach has already been implemented as a visualization tool by leading cohorts such as GnomAD (1), and used as a preprocessing step for machine learning approaches (2). Our preprint is also in the 99th pecentile of all-time bioRxiv preprints for the Altmetric impact statistic. Because of its extreme computational tractability and its ability to efficiently summarize genomic information, we expect that our approach will become a new standard tool in data analysis for large genomic studies.

The manuscript presents new methodology and applications that reveal fine-scale genomic and phenotypic variation in large-scale cohorts.  
We believe that this will be of interest to a broad variety of disciplines including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for eLife.

Sincerely,
Simon Gravel and Alex Diaz-Papkovich
on behalf of all authors.

1. Fancioli, Laurent. “gnomAD v2.1”. https://macarthurlab.org/2018/10/17/gnomad-v2-1/ (2018-10-17).
2. Tonkin-Hill, Gerry, et al. "Fast Hierarchical Bayesian Analysis of Population Structure." bioRxiv (2018): 454355.

### 2018-11-12 ms/main.tex
add manuscript tex

\documentclass[12pt]{pnas-new}
%\documentclass[draft]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
\newcommand{\sgcomment}[1]{{\textcolor{blue}{SG: #1}}}
\newcommand{\adpcomment}[1]{{\textcolor{orange}{ADP: #1}}}
%\usepackage{subfig}
\usepackage{adjustbox}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{dblfloatfix}
\usepackage[switch]{lineno}
\usepackage{comment}
%\usepackage[section]{placeins}

\linenumbers

\templatetype{pnasresearcharticle} % Choose template
\title{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b]{Alex Diaz-Papkovich}
\author[b,c]{Luke Anderson-Trocmé}
\author[b,c,1]{Simon Gravel} 
\affil[a]{Department of Quantitative Life Sciences, McGill University, Montreal, QC, H3A 0G1 Canada}
\affil[b]{McGill University and Genome Quebec Innovation Centre, Montreal, QC, H3A 0G1, Canada}
\affil[c]{Department of Human Genetics, McGill University, Montreal, QC, H3A 0G1, Canada. \textsuperscript{1}To whom correspondence should be addressed. E-mail: simon.gravel@mcgill.ca}

% Please include corresponding author, author contribution and author declaration information
\correspondingauthor{}

% Please give the surname of the lead author for the running footer
\leadauthor{Diaz-Papkovich}

\begin{abstract}
%Max 250
Background: Population structure in genetic data depends on complex demographic processes including geographic isolation, genetic drift, migration, and admixture. Together with technical artifacts, population structure is a prominent confounder of genomic studies. Identifying such patterns is therefore central to the genomic enterprise. Whereas many methods can identify specific types of population structure, few are able to provide simple representations of genomic diversity across a range of scales. We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP).
\\
\\
Results: We demonstrate using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank that projections using PCA-UMAP effectively cluster individuals who are genetically closely related while placing them in a global continuum of genetic variation. These projections reveal non-trivial population groupings, reflect ethnicity and geography on fine-scale levels, and uncover patterns in the distributions of a variety of phenotypes.
\\
\\Conclusion: These projections succinctly illustrate population structure in large cohorts and capture relationships on local and global scales, establishing PCA-UMAP as a general-purpose approach to exploratory analysis in genomics.
\end{abstract}

\keywords{genomics $|$ genetics $|$ population structure $|$ ethnicity $|$ ancestry $|$ machine learning $|$ data visualization $|$ dimension reduction}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\subsection*{Background}

Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference}.

Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations. %Relatedness patterns are shaped by geographic isolation, founder effects, migration patterns, and admixture. Understanding such patterns is therefore useful to understand the genetic history of individuals and populations. It is also important to interpret genomic association studies, as unaccounted-for relatedness patterns can generate spurious correlations between genotypes and phenotypes. 
If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$. Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.

Principal component analysis (PCA) is often the first dimensional reduction tool used for genomic data. It identifies and ranks directions in genotype space that explain most-to-least variance among individuals. Positions of individuals along directions of highest variance can then be used to summarize individual genotypes. PCA coordinates have natural genealogical interpretations in terms of times to a most recent common ancestor (TMRCA) \cite{mcvean2009genealogical}, and are used empirically to reveal admixture \cite{brisbin2012pcadmix}, continuous isolation-by-distance \cite{novembre2008europe, nelson2008population}, as well as technical artefacts. PCA coordinates are particularly well-suited to correct for population structure in GWAS~\cite{eigen2006}.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.8\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{Four methods of dimension reduction of 1KGP genotype data with population labels (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

As sample sizes increase, the amount of information encoded in the lower-variance principal components increases, and researchers typically examine multiple two-dimensional projections to get a sense of the data. While many features of the data can be identified in this manner, other features may be hidden by the projections or hard to interpret.

To display more of the high-dimensional features of the data in a two dimensional figure,  we can use non-linear transformations that seek to preserve the local structure of the data. A popular method for visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}. Whereas PC projection is designed to capture as much variance as possible for a linear transformation, t-SNE seeks a low-dimensional representations of the data that preserves pairwise distances, using a probabilistic heuristic to weight mismatches. t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}. Using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}. However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.

Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space \cite{2018arXivUMAP}. The assumption behind UMAP is that data are uniformly distributed on local manifolds in high-dimensional space, which can be approximated as fuzzy sets that are patched together to form a topological representation. One can then construct a low-dimensional topological representation that minimizes the differences between the two representations. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.

A common practice in dimensional reduction is to first apply PCA to reduce the number of dimensions before performing nonlinear dimensional reduction. In addition to being computationally advantageous, this discards statistical noise that can confound nonlinear approaches: Population structure arising from $n$ isolated randomly-mating demes can be described by the leading $n-1$ PCs, with the following PCs describing stochastic variation in relatedness~\cite{eigen2006}. Selecting the leading PCs therefore has potential to extract meaningful population structure while filtering out stochastic noise.

We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}.  
%A common practice in dimensional reduction is to first apply PCA before performing nonlinear dimensional reduction. We then then feed the top principal components into UMAP and tSNE to generate a two-dimensional representation of the data. After the projections are generated, we create dimensional images colored by variables such as ethnicity (or a proxy for ethnicity), admixture, geography, and phenotype.

\begin{figure}
\centering
   %\includegraphics[width=0.6\linewidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
   \caption{\adpcomment{Using a new image here} UMAP on the first 10 principal components of HRS data. Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}. Individuals with intermediate admixture ratios sometimes fall between clusters of Black, White, and Hispanic individuals (alternate projection shown in figure~\ref{fig:supp_hrs_admix_alt}). }
    %\caption{UMAP on the first 10 principal components of HRS data. coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. Individuals with intermediate admixture ratios sometimes fall between three clusters of Black, White, and Hispanic individuals. This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}}
    \label{fig:umap_hrs_admix}
\end{figure}

\subsection*{Results}
\begin{figure}
\centering
   %\includegraphics[width=0.6\linewidth]{images/HRS_HISP_UMAP_PC7_NC2_NN15_MD05_201844193148.pdf}
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_highlighted.pdf}
    \caption{The top $7$ principal components of the Hispanic population of the HRS projected by UMAP, colored by region of birth. The highlighted region consists almost entirely of individuals who were born in the Mountain region of the United States. This region contains New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same figure colored according to estimated continental ancestry proportion. Figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} shows that populations from the 1KGP do not map to the cluster when projected onto the UMAP embedding.}
    \label{fig:hrs_hisp}
\end{figure}

\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and PCA-UMAP (that is, UMAP with PCA pre-processing). Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either on the top principal components leads to more distinct population clusters and less defined continental structure (see figure \ref{fig:supp_megamontage_pc10_50} for PCA-tSNE). Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; further components gradually approach results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9} and~\ref{fig:supp_megamontage_pc10_50}).

Focusing on PCA-UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we also find several population clusters that reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a group reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. The East Asian super-population forms three sub-populations split by geography: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Looser geographical groupings include Colombians and Peruvians, and the Esan and Yoruba populations of Nigeria; both groupings appear as connected sub-clusters. The South Asian super-population also forms a loose grouping.

Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Only two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering is robust to, e.g., the choice of the number of PCs considered (see figure ~\ref{fig:supp_megamontage_pc10_50}).

Finally, contrasting UMAP and t-SNE, we find that UMAP preserves more of the global structure of the data than t-SNE, and is more robust to choices of data pre-processing (figure~\ref{fig:supp_megamontage_pc10_50}).

\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct population groups across the world, which makes the data particularly easy to cluster. Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas.

For example, the HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
We have scaled these three proportions to values between 0 and 255, to color individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry. Using the first 10 principal components and UMAP, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

\adpcomment{Talking about using HRS with 1KGP here. Changed "four" clusters to "several"}
The HRS forms several large groupings and clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are clearly visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). The clarity of the interconnected clusters varies by parameterization, but they consistently form a large, roughly connected group.

To investigate possible ancestries related to populations in the 1KGP we took two approaches. In the first we generated PC axes and a UMAP embedding using UMAP and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}). In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}). In the latter case, we again see a gradient in the Hispanic cluster, though this time with ethnicity rather than admixture. In both cases we see groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population. One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying PCA-UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region. One separate cluster, highlighted in figure~\ref{fig:hrs_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States. This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6. Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster. Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding (figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}).\adpcomment{Added the 1KGP note here}
The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations. A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} The UKBB provides genotype data on 488,377 individuals along with self-identified ethnic background in a hierarchical tree-structured dictionary. Participants provided ethnic background on two occasions. We used the initial ethnicity after finding minimal differences between the two. The dataset is majority White (88.3\% British, 2.6\% Irish, 3.4\% other), with large populations identifying as Black (1.6\% either African, Caribbean, or other), Asian (1.9\% either Indian, Pakistani, Bangladeshi, or other), Chinese (0.3\%), an other ethnic group (0.8\%), mixed ethnicity (0.6\%), or an unavailable response (0.5\%).

UMAP on the top 10 principal components reveals both continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): The patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of large and multi-ethnic datasets.

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.pdf}%
\caption{Principal components 1 and 2}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{UMAP on first 10 principal components}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{The UKBB projected onto two dimensions, colored by self-reported ethnic background. (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. (b) UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
%The main body of the image is made up of British and Irish individuals. Section (i) is zoomed in on the structure of Black/Black British individuals. Section (ii) is focused on East and South Asian populations, and section (iii) is a mix of a variety of ethnicities.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

The largest body in the figure consists of the White British and Irish populations. The Irish population concentrates in a portion of this group, but many individuals are also scattered throughout the British-identified population. Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters. Chinese individuals form a cluster, within what is suspected to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well. The East and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group". The patchwork of genetic neighbourhoods is connected by trails of admixed individuals. These trails come together in a nexus of individuals with a variety of ethnicities; many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group". Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asia backgrounds. Additionally, there are many individuals whose ethnicity is White but neither British nor Irish (AWB) forming clusters distinct from the British and Irish cluster.

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the projection in figure~\ref{fig:umap_ukbb_pc10} colored in by  geographical coordinates according to the Ordnance Survey National Grid (OSGB1936) as distances north and east of the Isles of Scilly. UMAP coordinates within the "White British" cluster broadly map to geographic coordinates, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines connect to the South East corner of this cluster, corresponding to the position of the city of London and reflecting its high migrant population.

\begin{figure}
    \centering
    \includegraphics[width=0.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{The UKBB projected onto two dimensions using PCA-UMAP with each individual colored by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colors individuals by their north-south ("northing") coordinates, and the right image colors them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Individuals with missing geographic data are not shown. To prevent outlying individuals from washing out the colour scheme, northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. To protect participant privacy, data has been randomized as explained in the materials and methods section.}
    \label{fig:umap_ukbb_geo_osgb}
\end{figure}

The detailed shape of extended clusters is not stable as we vary the number of PCs included. Figure \ref{fig:supp_montage_ukbb_eth} shows a UMAP plot using the top 20 PCs from the UKBB. The shape of the "White British Cluster" is notably different, and we observe finer patterns of geographic variation, yet the qualitative observations made above are maintained. As an alternate visualization of diversity's correlation with geography, we performed a 3D UMAP projection and converted the normalized UMAP values into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}). The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.

Similarly to UMAP, t-SNE applied to the UKBB data both displays diversity within the "White British" population and identifies clusters among other groups. However, it has three drawbacks: it is much slower, requiring 2.26 hours for its first thousand iterations alone on 10 principal components against UMAP's 14 minutes; it fails to find a global optimum, which results in a scattering of individuals and groups that are not stable across independent runs; and it does not identify continuity between different continental groups resulting from admixture (figure~\ref{fig:supp_ukbb_tsne}).

\subsubsection*{Patterns in phenotype related to population structure} Our involvement with the UK biobank data is through a project on autoimmune disease and asthma. More than in geographic coordinates, we are interested in whether genetic population structure correlates with phenotypes and covariates of interest.

\begin{figure}
    \centering
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
    \caption{The UKBB projected onto two dimensions using PCA-UMAP (as in~\ref{fig:umap_ukbb_pc10}), with females colored by age-adjusted difference from mean population height (left) and leukocyte counts (right). To protect participant privacy, data has been randomized as explained in the materials and methods section.}
    \label{fig:umap_height_female}
\end{figure}

Covariates such as height (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures  (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure. Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and several unidentified ethnic groups have noticeably lower-than-average heights. More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots). Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.

 Forced expiratory volume in 1 second (FEV1) (Figure~\ref{fig:supp_box_fev_f} also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture continuum, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; while another close to European and British populations has lower values relative to the population mean. These results merit further investigation and underscore the exploratory value of PCA-UMAP --- these  populations are largely unidentified and there is no straightforward way to separate them within the data otherwise.

\begin{figure}
    \centering
    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{A map of Great Britain colored by a 3D UMAP projection. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in color. Patterns in genetic similarity are visible in Scotland, South England, the East and West Midlands, and major urban centres. A flattened 2D view of of the 3D projection used for coloring is presented in the top right. To protect participant privacy, data has been randomized as explained in the materials and methods.}
    \label{fig:umap_ukbb_geo_map}
\end{figure}

\subsection*{Discussion}
Understanding population structure is important to identify confounders in medical genomics and studies of anthropology and human evolution. PCA of genomic data reflects genealogical and geographic data, but visualization in large datasets still requires scanning through a large number of pairwise plots. UMAP condenses these components and comprehensively illustrates information --- phenotypic, geographic, and ancestral --- contained within genotypes on fine-scale levels and within the context of a global population structure. In large datasets where the number of significant PCs is large, the resulting representation has important advantages over
PCA alone and provides a superior visualization to t-SNE.

%PCA-UMAP reveals numerous clusters that are based on shared ancestries, admixture proportion, and geography. Examples include the closeness of the Chinese Dai and Kinh of Vietnam and the diversity and admixture of the American Hispanic population.  Using the 1KGP, we illustrate in figures~\ref{fig:supp_megamontage_pc2_9} and~\ref{fig:supp_megamontage_pc10_50} that using progressively more principal components can create clusters on finer scales than previously demonstrated.

Examinations of clustering in the three datasets provided many intriguing clusters that would otherwise have been difficult to identify. In particular, several areas from figure~\ref{fig:umap_ukbb_pc10}, highlighted in figure~\ref{fig:supp_ukbb_zoom}, show multiple unidentified groups related to each of the East Asian and South Asian super-populations, as well as to either or both of African or admixed populations. Additionally, the Hispanic population of the HRS contains a geographically-restricted cluster that could not have been identified from pairwise examinations of principal components. The 1KGP --- frequently used in medical and population studies --- contained splits in the Gujarati and Punjabi population samples that were not visible PCA or Admixture analysis alone (although a split among Gujarati is arguably visible in the Admixture analysis with K=12 in \cite{10002015global}).

Application to the UKBB underscores the strength of PCA-UMAP in large cohorts. We see clear, fine relationships between genotype and phenotype and geography, and this is presented in a visualization that accounts for natural genetic clustering. Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with phenotypes such as height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.

Importantly, using UMAP is straightforward and fast. Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the a "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results). Given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.

There are downsides to using nonlinear approaches to visualize the data. Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for populations with larger sample sizes compared to PCA. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult. Nonlinearity also complicates the interpretation of results. Distances in UMAP or tSNE space should not be used as a proxy for genetic distance. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artifact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

\subsection*{Conclusion}
With these caveats in mind, a priori data visualization plays a central role in quality control, hypothesis generation, and confounder identification for a wide range of genomic applications. Nonlinear approaches, despite their limitations, become increasingly useful as the size of datasets increases: We have shown that UMAP, in particular, reveals a wide range of features that would not be apparent using linear maps. Given its ease of use, breadth of results, and low computational cost, we propose that UMAP should become a default companion to PCA in large genomic cohorts.

\matmethods{We used genotype data from 12,454 individuals from the Health and Retirement Study (HRS), genotyped on the Illumina Human Omni 2.5M platform\cite{juster1995overview}. Principal components were computed in PLINK v1.90b5.2 64-bit\cite{purcell2007plink} using variants with a minor allele frequency greater than 0.05, Hardy-Weinberg p-value of more than $1\times 10^{-6}$, and genotype missing rate of less than $0.1$, and sample with genotype missing rate of less than $0.1$. We used the principal components of genotype data from 488,377 individuals in the UK BioBank (UKBB) as computed by the cohort \cite{sudlow2015uk}. We used genotype data from 3,450 individuals from the 1KGP project using Affy 6.0 genotyping\cite{10002015global}.

Scripts for all tests and plotting functions can be found on \url{https://github.com/diazale/gt-dimred}. A demo version using freely available 1KGP data is available at \url{https://github.com/diazale/1KGP_dimred}. PCA and standard t-SNE were done with Scikit-learn\cite{scikit-learn}. UMAP was performed using a Python implementation\cite{2018arXivUMAP}. Statistical testing was done in SciPy\cite{scipy} and StatsModels\cite{seabold2010statsmodels}.

Both UMAP and t-SNE feature a number of adjustable parameters. Among the parameters that we varied, the number of PCs used in pre-processing of the data has the largest effect for both methods (see figures~\ref{fig:supp_megamontage_pc2_9} and \ref{fig:supp_megamontage_pc10_50}).

We tested different choices for perplexity in t-SNE. The default value of 30 provided comparable performance to other parameter choices.  Similarly, we tested different parameter choices for UMAP, with the clearest results generated by specifying 15 nearest neighbours (the default value) and a "minimum distance" between points in low dimensions of 0.5. UMAP developers described "sensible" values for nearest neighbours as between 5 and 50 and minimum distance between 0.5 and 0.001.

UMAP and t-SNE projections were carried out on an iMac with a 3.5GhZ Intel Core i7 processor, 32 GB 1600 MHz DDR3 of RAM, and an NVIDIA GeForce GTX 775M 2048 MB graphics card.

To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figures. We rounded each attribute to an attribute-specific number of bins, and then permuted the data in the following way: 
For each point (i.e. each individual) in UMAP visualizations, and each attribute, we identified the 9 nearest neighbouring points, and copied the attribute from a randomly selected neighbor (thus allowing for the possibility of one value being printed twice).  Because this process is done independently for each visualization, a given point shown on the figure will copy values from different randomly selected individuals. Additionally, spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km.

For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}

\showmatmethods{} % Display the Materials and Methods section

%\acknow{}

%\showacknow{} % Display the acknowledgments section

\subsection*{Declarations}
\subsubsection*{Ethics approval and consent to participate}
HRS data was under IRB Study No. A11-E91-13B - The apportionment of genetic diversity within the United States. UKBB data was accessed under accession number 6728.

\subsubsection*{Consent for publication} Not applicable.

\subsubsection*{Availability of data and material} All data is publicly available to researchers.

\subsubsection*{Competing interests}
We have no competing interests to declare.

\subsubsection*{Funding}
This research was undertaken, in part, thanks to funding from the Canada Research Chairs program and CIHR grant MOP-136855.

\subsubsection*{Authors' contributions}
A.D.P. and S.G. designed the research. A.D.P. carried out dimension reduction and analysis. S.G. provided analysis. A.D.P and S.G. wrote the paper. L.A.T. provided the UK map visualization and analysis.

\subsubsection*{Acknowledgements}
We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the dataset. We also thank Chief Ben-Eghan, Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access, and Selin Jessa for introducing us to UMAP. This research has been conducted using the UK Biobank Resource under Application Number 6728.

% Bibliography
\bibliography{pca_umap_genotype}

\newpage

\beginsupplement
\subsection*{Supporting Information (SI)}
\subsubsection*{SI Figures}
\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/megamontage_PC2_9.pdf}
    \end{subfigure}
    \caption{UMAP (left two columns) and t-SNE (right two columns) applied to the top principal components of the 1KGP labelled by the number of components used. Adding more components results in progressively finer population clusters using both methods.}
    \label{fig:supp_megamontage_pc2_9}  
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/megamontage_PC10_50.pdf}
    \end{subfigure}
    \caption{UMAP (left two columns) and t-SNE (right two columns) applied to the top principal components of the 1KGP labelled by the number of components used. Results are similar until approximately 11 components, where t-SNE breaks apart clusters of South Asian (in green) and Central and South American populations (in pink) while UMAP preserves them. At approximately 30 components populations begin to drift together with UMAP and disperse with t-SNE.}
    \label{fig:supp_megamontage_pc10_50}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/megamontage_eth_ukbb.pdf}
    \end{subfigure}
    \caption{PCA-UMAP on UKBB data, colored by self-identified ethnic background. Images are labelled by the number of components included.}
    \label{fig:supp_montage_ukbb_eth}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/montage_ukbb_ns_pm.pdf}
    \end{subfigure}
    \caption{PCA-UMAP on UKBB data, colored by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates. Images are labelled by the number of components included. Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_montage_ukbb_ns}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/montage_ukbb_ew_pm.pdf}
    \end{subfigure}
    \caption{PCA-UMAP on UKBB data, colored by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates. Images are labelled by the number of components included. Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_montage_ukbb_ew}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.8\textwidth}
    \includegraphics[width=\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_combined_resized.pdf}
    \end{subfigure}
    \caption{Zoomed in areas of figure~\ref{fig:umap_ukbb_pc10}. Sections (i) and (ii) respectively focus on the African and Asian superpopulations, and section (iii) focuses on an area with individuals from many ethnic backgrounds. Noticeable clusters of unidentified ethnic backgrounds appear and are labelled "OEG" "(Other Ethnic Group)".}
    \label{fig:supp_ukbb_zoom}
\end{figure*}

%%%%% UKBB phenotype plots
\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417039_basophill_count_pct5_f.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by basophil count (female). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_basophill_f}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417039_basophill_count_pct5_m.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by basophil count (male). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_basophill_m}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417720_eosinophill_count_pct5_f.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_eosinophill_f}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417720_eosinophill_count_pct5_m.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_eosinophill_m}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416305_3063_0_0_pct1_f.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_fev_f}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416305_3063_0_0_pct1_m.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_fev_m}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by height (female). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_height_f}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_m.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by height (male). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_height_m}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_leukocyte_f}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_m.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_leukocyte_m}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714165614_neutrophill_count_pct5_f.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_neutrophill_f}
\end{figure}

\begin{figure}
    \centering
    %\begin{subfigure}{0.8\textwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714165614_neutrophill_count_pct5_m.pdf}
    %\end{subfigure}
    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_ukbb_neutrophill_m}
\end{figure}

%%%%% UKBB boxplots below

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_basophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_basophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_basophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_basophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_eosinophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_eosinophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_eosinophill_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_eosinophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_fev_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_fev_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_height_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_height_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_leukocyte_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_leukocyte_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_leukocyte_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_leukocyte_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/female_neutrophil_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_neutrophill_f}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/male_neutrophil_boxplot_annotated.pdf}
    \end{subfigure}
    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
    \label{fig:supp_box_neutrophill_m}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/montage_fev1_height_afr_permuted.pdf}
    \end{subfigure}
    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) colored by self-identified ethnic background (left, from figure~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right). An arrow points to an area where the FEV1 distribution appears to change, corresponding to where the clusters contain more people with self-identified mixed backgrounds.}
    \label{fig:supp_comparison_fev_afr}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.pdf}
    \end{subfigure}
    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) colored by ethnicity (left), FEV1 (middle), and age-adjusted height (right). The OEG cluster next to the Chinese cluster is colored differently, suggesting this population may have different FEV1 characteristics. A cluster of OEG/other white individuals is more blue, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
    \label{fig:supp_comparison_fev_chi_eur}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    %\includegraphics[width=0.95\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_2018330153123_RACE_HISP_MEX.pdf}
    \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_race_hisp_mex_labels.pdf}
    \end{subfigure}
    \caption{UMAP applied to the first 10 principal components of HRS data. Points colored by self-identified race, Hispanic status, and Mexican-American status. The cluster on the left is mostly people who identify as neither Black nor White and were born outside the contiguous United States or in the Pacific census region. Clustering with the 1KGP data places them with Asian-identified populations. BNH, Black (not Hispanic); BHO, Black (Hispanic, Other); WNH, White (not Hispanic); WHM, White (Hispanic, Mexican-American); WHO, White Hispanic (Other); ONH, Other (not Hispanic); OHM, Other (Hispanic, Mexican-American); OHO, Other (Hispanic, Other).}
    \label{fig:supp_umap_hrs_eth}
\end{figure*}

%\begin{figure*}
%    \centering
%    \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_race_hisp_mex_labels.jpeg}
%    \end{subfigure}
%    \caption{}
%    \label{fig:supp_hrs_projection_2}
%\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    %\includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_2018330153123_born.pdf}
    \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_born.pdf}
    \end{subfigure}
    \caption{UMAP on the top 10 principal components of the HRS dataset, colored by Census Bureau birth region. There is no obvious pattern in the clusters of majority "White Not Hispanic" individuals.}
    \label{fig:supp_hrs_born}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_custom_label.pdf}
    \end{subfigure}
    \caption{\adpcomment{This is also new} UMAP on the top 10 principal components of the HRS data, with 1KGP data projected onto the embedding. Individuals from the HRS are coloured grey and largely transparent. British (GBR) and other European (CEU) individuals are scattered throughout the "White Not Hispanic" clusters. Finns (FIN) form clear groupings. Spanish (IBS) and Italian (TSI) individuals cluster near the Hispanic grouping. There is a clear gradient in the Hispanic cluster formed of Puerto Ricans (PUR), Colombians (CLM), Mexicans (MXL), and Peruvians (PEL). Populations with African ancestry (AFR) appear with Black individuals. East Asian (EAS) populations comprising Chinese, Kinh, and Japanese individuals cluster together with what appears in figure~\ref{fig:umap_hrs_admix} as a population of mostly Asian ancestry. South Asian (SAS) populations with Indian, Pakistani, and Sri Lankan ancestry cluster in a separate area. One "White Not Hispanic" cluster at the bottom does not cluster with any 1KGP populations.}
    \label{fig:supp_hrs_1kgp_projected}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
    \end{subfigure}
    \caption{UMAP on the top 10 principal components of the combined HRS and 1KGP datasets. The colors match those used in the separate 1KGP and HRS plots. The large blue clusters are "White Not Hispanic" individuals. Scattered brown dots within these clusters are Utah residents with Northern/Western European ancestry, and British in England and Scotland ("CEU" and "GBR", respectively, in the 1KGP datasets). A cluster of Finnish individuals consistently shows up and is labelled "FIN". Toscani individuals regularly form clusters and are labelled "TSI". However, their locations relative to the other clusters of "White Not Hispanic" are not consistent, sometimes appearing next to clusters of Spanish individuals and other times not. One cluster of "White Not Hispanic" individuals is highlighted in a box --- unlike the other clusters, CEU and GBR individuals are not projected here.}
    \label{fig:supp_hrs_1000g}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_PCGRID_8.pdf}
    \end{subfigure}
    \caption{Pairwise plots of the first 8 principal components of the Hispanic subset of the HRS. Those born in the Mountain region are colored green.}
    \label{fig:supp_hrs_hisp_grid}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    %\includegraphics[width=\textwidth]{images/HRS_HISP_UMAP_PC7_NN15_MD05_admix.pdf}
    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
    \end{subfigure}
    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, colored by estimated admixture proportions.}
    \label{fig:supp_umap_hrs_hisp_admix}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    %\includegraphics[width=\textwidth]{images/HRS_HISP_UMAP_PC7_NN15_MD05_admix.pdf}
    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \end{subfigure}
    \caption{UMAP of the first $7$ principal components of the Hispanic population of the HRS, colored by birth region, with selected 1KGP populations projected onto the UMAP embedding. The highlighted area consists almost entirely of individuals born in the Mountain region of the census (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). The six populations do not fall onto the highlighted cluster.  CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy.}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/admixture_plot_highlight_mountain_copy.pdf}
    \end{subfigure}
    \caption{Admixture plot of Hispanic individuals in the HRS. Those born in the Mountain census region fall between the white lines (indices 48 to 184)}
    \label{fig:supp_hrs_hisp_admix}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
    \end{subfigure}
    \caption{UMAP on the first $10$ principal components of HRS data, coloring individuals by estimated admixture from three ancestral populations. Alternate projection to demonstrate how individuals may fall between clusters.}
    \label{fig:supp_hrs_admix_alt}
\end{figure*}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\columnwidth]{images/UKBB_TSNE_10PCs_DefaultPerplexity_eth.pdf}
    \caption{t-SNE applied to the top 10 principal components of the UKBB, colored by ethnic background. The unbalanced populations resulted in many individuals and populations being orphaned along the periphery of the main cluster.}
    \label{fig:supp_ukbb_tsne}
\end{figure}

\end{document}

### 2018-11-13 ms/main.tex
validated alex comments

@@
 \author[a,b]{Alex Diaz-Papkovich}
-\author[b,c]{Luke Anderson-Trocmé}
+\author[b,c]{Luke Anderson-Trocm\'e}
 \author[b,c,1]{Simon Gravel} 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{images/montage_1000G_labels_axes.pdf}
+    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
     \caption{Four methods of dimension reduction of 1KGP genotype data with population labels (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry.
@@
    \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
-   \caption{\adpcomment{Using a new image here} UMAP on the first 10 principal components of HRS data.
+   \caption{UMAP on the first 10 principal components of HRS data.
 Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
 This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.
-Individuals with intermediate admixture ratios sometimes fall between clusters of Black, White, and Hispanic individuals (alternate projection shown in figure~\ref{fig:supp_hrs_admix_alt}).
-}
+Individuals with intermediate admixture ratios sometimes fall between clusters of Black, White, and Hispanic individuals \adpcomment{(alternate projection shown in figure~\ref{fig:supp_hrs_admix_alt}).} }
     %\caption{UMAP on the first 10 principal components of HRS data.
@@
 
-\adpcomment{Talking about using HRS with 1KGP here. Changed "four" clusters to "several"}
 The HRS forms several large groupings and clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}).
-Gradients in admixture proportion are clearly visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics.
+Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics.
 The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}).
@@
 In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}).
-In the latter case, we again see a gradient in the Hispanic cluster, though this time with ethnicity rather than admixture.
-In both cases we see groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population.
+Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population.
 One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.
@@
 Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster.
-Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding (figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}).\adpcomment{Added the 1KGP note here}
+Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding (figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}).
 The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations.

### 2018-11-16 ms/main.tex
removed comments, changed some HRS Hispanic image and HRS_1KGP combined projection, rewrote abstract, minor ce

@@
 Background: Population structure in genetic data depends on complex demographic processes including geographic isolation, genetic drift, migration, and admixture.
-Together with technical artifacts, population structure is a prominent confounder of genomic studies.
-Identifying such patterns is therefore central to the genomic enterprise.
-Whereas many methods can identify specific types of population structure, few are able to provide simple representations of genomic diversity across a range of scales.
-We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP).
+Existing data exploration methods struggle with unbalanced cohort ancestries, reconciliation of global versus local relationships, and the sheer size of modern datasets.
+We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
 \\
@@
 Results: We demonstrate using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank that projections using PCA-UMAP effectively cluster individuals who are genetically closely related while placing them in a global continuum of genetic variation.
-These projections reveal non-trivial population groupings, reflect ethnicity and geography on fine-scale levels, and uncover patterns in the distributions of a variety of phenotypes.
+These projections quickly reveal subpopulations within the American Hispanic population, deep relationships between geography and genotype in the UK population, and stark phenotypic variation between groups in the UK Biobank.
 \\
-\\Conclusion: These projections succinctly illustrate population structure in large cohorts and capture relationships on local and global scales, establishing PCA-UMAP as a general-purpose approach to exploratory analysis in genomics.
+\\
+Conclusion: These projections succinctly illustrate population structure in large cohorts and capture relationships on local and global scales, establishing PCA-UMAP as a general-purpose approach to exploratory analysis in genomics.
 \end{abstract}
@@
 Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations.
-%Relatedness patterns are shaped by geographic isolation, founder effects, migration patterns, and admixture.
-Understanding such patterns is therefore useful to understand the genetic history of individuals and populations.
-It is also important to interpret genomic association studies, as unaccounted-for relatedness patterns can generate spurious correlations between genotypes and phenotypes.
 If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$.
@@
 We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}.
-%A common practice in dimensional reduction is to first apply PCA before performing nonlinear dimensional reduction.
-We then then feed the top principal components into UMAP and tSNE to generate a two-dimensional representation of the data.
-After the projections are generated, we create dimensional images colored by variables such as ethnicity (or a proxy for ethnicity), admixture, geography, and phenotype.
 
@@
 Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
-This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.
-Individuals with intermediate admixture ratios sometimes fall between clusters of Black, White, and Hispanic individuals \adpcomment{(alternate projection shown in figure~\ref{fig:supp_hrs_admix_alt}).} }
-    %\caption{UMAP on the first 10 principal components of HRS data.
-coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
-Individuals with intermediate admixture ratios sometimes fall between three clusters of Black, White, and Hispanic individuals.
-This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}}
+This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
     \label{fig:umap_hrs_admix}
@@
 
-\subsection*{Results}
-\begin{figure}
-\centering
-   %\includegraphics[width=0.6\linewidth]{images/HRS_HISP_UMAP_PC7_NC2_NN15_MD05_201844193148.pdf}
-   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_highlighted.pdf}
-    \caption{The top $7$ principal components of the Hispanic population of the HRS projected by UMAP, colored by region of birth.
+%\subsection*{Results}
+%\begin{figure}
+%\centering
+%   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_highlighted.pdf}
+%    \caption{The top $7$ principal components of the Hispanic population of the HRS projected by UMAP, colored by region of birth.
 The highlighted region consists almost entirely of individuals who were born in the Mountain region of the United States.
@@
 Figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} shows that populations from the 1KGP do not map to the cluster when projected onto the UMAP embedding.}
-    \label{fig:hrs_hisp}
-\end{figure}
+%    \label{fig:hrs_hisp}
+%\end{figure}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.6\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
+    \end{subfigure}
+    \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
+Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
+When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster.
+Six relevant 1KGP populations are presented: CLM, Colombian in Medellin, Colombia;
+    IBS, Iberian in Spain;
+    MXL, Mexican in Los Angeles, California;
+    PEL, Peruvian;
+    PUR, Puerto Rican;
+    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.}
+    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
+\end{figure*}
 
@@
 \subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying PCA-UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
-One separate cluster, highlighted in figure~\ref{fig:hrs_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
+One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
 This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6.
 Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster.
-Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding (figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}).
+Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding.
 The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations.
@@
     \begin{subfigure}{\textwidth}
-    %\includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_2018330153123_born.pdf}
     \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_born.pdf}
@@
     \end{subfigure}
-    \caption{\adpcomment{This is also new} UMAP on the top 10 principal components of the HRS data, with 1KGP data projected onto the embedding.
-Individuals from the HRS are coloured grey and largely transparent.
+    \caption{UMAP on the top 10 principal components of the HRS data, with 1KGP data projected onto the embedding.
+Individuals from the HRS are grey.
 British (GBR) and other European (CEU) individuals are scattered throughout the "White Not Hispanic" clusters.
@@
 Spanish (IBS) and Italian (TSI) individuals cluster near the Hispanic grouping.
-There is a clear gradient in the Hispanic cluster formed of Puerto Ricans (PUR), Colombians (CLM), Mexicans (MXL), and Peruvians (PEL).
+There are sub-groups in the Hispanic cluster formed of Puerto Ricans (PUR), Colombians (CLM), Mexicans (MXL), and Peruvians (PEL).
 Populations with African ancestry (AFR) appear with Black individuals.
@@
     \end{subfigure}
-    \caption{UMAP on the top 10 principal components of the combined HRS and 1KGP datasets.
-The colors match those used in the separate 1KGP and HRS plots.
-The large blue clusters are "White Not Hispanic" individuals.
-Scattered brown dots within these clusters are Utah residents with Northern/Western European ancestry, and British in England and Scotland ("CEU" and "GBR", respectively, in the 1KGP datasets).
-A cluster of Finnish individuals consistently shows up and is labelled "FIN".
-Toscani individuals regularly form clusters and are labelled "TSI".
-However, their locations relative to the other clusters of "White Not Hispanic" are not consistent, sometimes appearing next to clusters of Spanish individuals and other times not.
-One cluster of "White Not Hispanic" individuals is highlighted in a box --- unlike the other clusters, CEU and GBR individuals are not projected here.}
+    \caption{UMAP projection of the top 10 principal components of the combined HRS and 1KGP datasets.
+One cluster (in the box) does not group with any of the 1KGP populations.
+A cluster of Finnish (FIN) individuals consistently appears in the "White Not Hispanic" (WNH) group.
+Groups of Central and South American populations from the 1KGP (CLM, Colombian; MXL, Mexican; PEL, Peruvian; PUR, Puerto Rican) form nearby or within the HRS Hispanic cluster (HIS).
+Iberian individuals (IBS) cluster near the Hispanic population.
+Toscani individuals (TSI) form some small clusters and sometimes appear near the Iberian and Hispanic populations.
+Individuals with British/Scottish (GBR) or Northern/Western European ancestry (CEU) are scattered throughout the WNH clusters.
+Individuals with African ancestry from the 1KGP group with Black Americans from the HRS (AFR).
+Similar population groupings occur with South Asian (SAS) and East Asian (EAS) individuals.}
     \label{fig:supp_hrs_1000g}
@@
 
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    %\includegraphics[width=\textwidth]{images/HRS_HISP_UMAP_PC7_NN15_MD05_admix.pdf}
-    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
-    \end{subfigure}
-    \caption{UMAP of the first $7$ principal components of the Hispanic population of the HRS, colored by birth region, with selected 1KGP populations projected onto the UMAP embedding.
+%\begin{figure*}
+%    \centering
+%    \begin{subfigure}{\textwidth}
+%    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
+%    \end{subfigure}
+%    \caption{UMAP of the first $7$ principal components of the Hispanic population of the HRS, colored by birth region, with selected 1KGP populations projected onto the UMAP embedding.
 The highlighted area consists almost entirely of individuals born in the Mountain region of the census (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
@@
 CLM, Colombian in Medellin, Colombia;
-    IBS, Iberian in Spain;
-    MXL, Mexican in Los Angeles, California;
-    PEL, Peruvian;
-    PUR, Puerto Rican;
-    TSI, Tuscani in Italy.}
-    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
-\end{figure*}
+%    IBS, Iberian in Spain;
+%    MXL, Mexican in Los Angeles, California;
+%    PEL, Peruvian;
+%    PUR, Puerto Rican;
+%    TSI, Tuscani in Italy.}
+%    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
+%\end{figure*}
 
@@
 
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
-    \end{subfigure}
-    \caption{UMAP on the first $10$ principal components of HRS data, coloring individuals by estimated admixture from three ancestral populations.
+%\begin{figure*}
+%    \centering
+%    \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
+%    \end{subfigure}
+%    \caption{UMAP on the first $10$ principal components of HRS data, coloring individuals by estimated admixture from three ancestral populations.
 Alternate projection to demonstrate how individuals may fall between clusters.}
-    \label{fig:supp_hrs_admix_alt}
-\end{figure*}
+%    \label{fig:supp_hrs_admix_alt}
+%\end{figure*}
 

### 2018-11-16 ms/main.tex
edited abstract

@@
 %Max 250
-Background: Population structure in genetic data depends on complex demographic processes including geographic isolation, genetic drift, migration, and admixture.
-Existing data exploration methods struggle with unbalanced cohort ancestries, reconciliation of global versus local relationships, and the sheer size of modern datasets.
+Background: Genetic structure in large cohorts can arise from technical, sampling and demographic variation.
+Data visualisation is therefore a first step in most genomic analyses.
+However, existing data exploration methods struggle with unbalanced sampling and the multiple scales of population structure in large cohorts.
 We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
@@
 \\
-Results: We demonstrate using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank that projections using PCA-UMAP effectively cluster individuals who are genetically closely related while placing them in a global continuum of genetic variation.
-These projections quickly reveal subpopulations within the American Hispanic population, deep relationships between geography and genotype in the UK population, and stark phenotypic variation between groups in the UK Biobank.
+Results: Using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank, we demonstrate that PCA-UMAP effectively clusters individuals who are genetically closely related while placing them in a global continuum of genetic variation.
+This dimension reduction approach reveals previously overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank.
 \\
 \\
-Conclusion: These projections succinctly illustrate population structure in large cohorts and capture relationships on local and global scales, establishing PCA-UMAP as a general-purpose approach to exploratory analysis in genomics.
+Conclusion: PCA-UMAP reveals genetic subgroups among study participants that had been previously overlooked, opening new lines of investigation for demographic research and statistical genetics.
+Given its small computational cost, it provides an important general-purpose approach to exploratory analysis in population-scale datasets.
 \end{abstract}

### 2018-11-16 ms/main.tex
minor tweaks

@@
 Data visualisation is therefore a first step in most genomic analyses.
-However, existing data exploration methods struggle with unbalanced sampling and the multiple scales of population structure in large cohorts.
+However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure in large cohorts.
 We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
@@
     \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
-Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
+Coloring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
 When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster.
-Six relevant 1KGP populations are presented: CLM, Colombian in Medellin, Colombia;
+Six 1KGP populations are presented:
+    CLM, Colombian in Medellin, Colombia;
     IBS, Iberian in Spain;
@@
     PUR, Puerto Rican;
-    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.}
+    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions.}
     \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
@@
 Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
-%The main body of the image is made up of British and Irish individuals.
-Section (i) is zoomed in on the structure of Black/Black British individuals.
-Section (ii) is focused on East and South Asian populations, and section (iii) is a mix of a variety of ethnicities.
 BA, Black African; 
@@
 Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters.
-Chinese individuals form a cluster, within what is suspected to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well.
-The East and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group".
+Chinese individuals form a cluster, within appears to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well.
+The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group".
 The patchwork of genetic neighbourhoods is connected by trails of admixed individuals.
 These trails come together in a nexus of individuals with a variety of ethnicities; many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group".
-Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asia backgrounds.
+Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asian backgrounds.
 Additionally, there are many individuals whose ethnicity is White but neither British nor Irish (AWB) forming clusters distinct from the British and Irish cluster.
 
-Figure~\ref{fig:umap_ukbb_geo_osgb} presents the projection in figure~\ref{fig:umap_ukbb_pc10} colored in by  geographical coordinates according to the Ordnance Survey National Grid (OSGB1936) as distances north and east of the Isles of Scilly.
+Figure~\ref{fig:umap_ukbb_geo_osgb} presents the projection in figure~\ref{fig:umap_ukbb_pc10} colored in by  geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
 UMAP coordinates within the "White British" cluster broadly map to geographic coordinates, as has been observed in Europe-wide data\cite{novembre2008europe}.
@@
 Individuals with missing geographic data are not shown.
-To prevent outlying individuals from washing out the colour scheme, northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km.
+To prevent outlying individuals from washing out the color scheme, northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km.
 To protect participant privacy, data has been randomized as explained in the materials and methods section.}
@@
 Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and several unidentified ethnic groups have noticeably lower-than-average heights.
-More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
+More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t-test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
 Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.
 
- Forced expiratory volume in 1 second (FEV1) (Figure~\ref{fig:supp_box_fev_f} also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
+Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
 Notably, there appears to be a juncture in the admixture continuum, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes.
@@
 
-%PCA-UMAP reveals numerous clusters that are based on shared ancestries, admixture proportion, and geography.
-Examples include the closeness of the Chinese Dai and Kinh of Vietnam and the diversity and admixture of the American Hispanic population.
-Using the 1KGP, we illustrate in figures~\ref{fig:supp_megamontage_pc2_9} and~\ref{fig:supp_megamontage_pc10_50} that using progressively more principal components can create clusters on finer scales than previously demonstrated.
-
 Examinations of clustering in the three datasets provided many intriguing clusters that would otherwise have been difficult to identify.
@@
 Importantly, using UMAP is straightforward and fast.
-Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the a "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results).
+Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results).
 Given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
@@
 
-For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}
+For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the color value from a randomly selected neighbour.}
 

### 2018-11-16 ms/cover_letter.tex
add letter

@@
+\documentclass{article}
+\setlength\parindent{0pt}
+
+\begin{document}
 Dear Editor,
-
-We are pleased to submit our manuscript "UMAP reveals deep population structure and phenotype heterogeneity in large genomic cohorts" for publication in eLife.
+\\\\
+We are pleased to submit our manuscript ``UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts'' for publication in eLife.
 This work applies a newly developed dimensional reduction method called UMAP to visualize population structure in genomic datasets.
-
+\\\\
 By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts.
@@
 We also identify multiple associations between genetics, geography, and phenotypes in the UK biobank and sub-continental population structure within 1000 Genomes Project populations.
-
+\\\\
 This work has broad applicability.
-Our approach has already been implemented as a visualization tool by leading cohorts such as GnomAD (1), and used as a preprocessing step for machine learning approaches (2).
-Our preprint is also in the 99th pecentile of all-time bioRxiv preprints for the Altmetric impact statistic.
+Our approach has already been implemented as a visualization tool by leading cohorts such as GnomAD\footnote{Fancioli, Laurent.
+``gnomAD v2.1''.
+https://macarthurlab.org/2018/10/17/gnomad-v2-1/ (2018-10-17).}, and used as a preprocessing step for machine learning approaches\footnote{Tonkin-Hill, Gerry, et al.
+``Fast Hierarchical Bayesian Analysis of Population Structure.'' bioRxiv (2018): 454355.}.
+Our preprint is also in the 99th percentile of all-time bioRxiv preprints for the Altmetric impact statistic.
 Because of its extreme computational tractability and its ability to efficiently summarize genomic information, we expect that our approach will become a new standard tool in data analysis for large genomic studies.
-
+\\\\
 The manuscript presents new methodology and applications that reveal fine-scale genomic and phenotypic variation in large-scale cohorts.
 We believe that this will be of interest to a broad variety of disciplines including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for eLife.
-
+\\
+\\
 Sincerely,
+\\
 Simon Gravel and Alex Diaz-Papkovich
+\\
 on behalf of all authors.
 
-1. Fancioli, Laurent. “gnomAD v2.1”. https://macarthurlab.org/2018/10/17/gnomad-v2-1/ (2018-10-17).
-2. Tonkin-Hill, Gerry, et al. "Fast Hierarchical Bayesian Analysis of Population Structure." bioRxiv (2018): 454355.
+\end{document}

### 2018-11-16 ms/main.tex
removed redundancy in abstract

@@
 Results: Using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank, we demonstrate that PCA-UMAP effectively clusters individuals who are genetically closely related while placing them in a global continuum of genetic variation.
-This dimension reduction approach reveals previously overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank.
+This dimension reduction approach reveals subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank.
 \\

### 2018-11-16 ms/main.tex
shorter abstract

@@
 %Max 250
-Background: Genetic structure in large cohorts can arise from technical, sampling and demographic variation.
-Data visualisation is therefore a first step in most genomic analyses.
-However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure in large cohorts.
-We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
-\\
-\\
-Results: Using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank, we demonstrate that PCA-UMAP effectively clusters individuals who are genetically closely related while placing them in a global continuum of genetic variation.
-This dimension reduction approach reveals subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank.
-\\
-\\
-Conclusion: PCA-UMAP reveals genetic subgroups among study participants that had been previously overlooked, opening new lines of investigation for demographic research and statistical genetics.
-Given its small computational cost, it provides an important general-purpose approach to exploratory analysis in population-scale datasets.
+Genetic structure in large cohorts results from technical, sampling and demographic variation.
+Visualisation is therefore a first step in most genomic analyses.
+However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure.
+We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
+Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation.
+This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population.
+This opens new lines of investigation for demographic research and statistical genetics.
+Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
 \end{abstract}

### 2018-12-05 scripts/tsne_seeded_by_umap.py
tweaks to code, comparisons of methods

@@ -1,0 +1,100 @@
+import numpy as np
+import logging
+import os
+import sklearn
+import sys
+import time
+
+from sklearn.manifold import TSNE
+from sklearn.manifold import TSNE_no_ee as cTSNE
+
+# directory to store intermediate steps (for animations)
+anim_dir = "/Volumes/Stockage/alex/animations"
+
+# PCA files
+ukbb_dir = "/Volumes/Stockage/alex/ukbb_projections"
+hrs_dir = "/Volumes/Stockage/alex/hrs/projections"
+tg_dir = "/Volumes/Stockage/alex/1000G/projections"
+
+ukbb_path = os.path.join(ukbb_dir,"ukbb_pca_only")
+hrs_path = os.path.join(hrs_dir,"hrs_200_pc")
+tg_path = os.path.join(tg_dir,"pca_1000g_100")
+
+print("This script uses 10 principal components.")
+
+dset = sys.argv[1] # Dataset - must be UKBB/HRS/1000G
+num_iter = int(sys.argv[2]) # Number of iterations (default 1000)
+seed_dset = sys.argv[3] # Whether to seed by UMAP or just use the PCs
+pcs = [10]
+
+tstamp = ''.join([str(t) for t in time.gmtime()[0:6]])
+
+# Directory for this particular animation
+anim_path = os.path.join(anim_dir, dset + "_" + tstamp)
+os.makedirs(anim_path)
+
+# Load the PCA data and UMAP data and set the output directory
+try:
+    if dset=='UKBB':
+        with open(ukbb_path) as pc:
+            pca_contents = pc.readlines()
+
+        pca_data = []
+
+        for pc in pca_contents[1:]:
+            pca_data.append(pc.split()[3:len(pc)])
+
+        pca_data_array = np.array(pca_data).astype(np.float)
+        del(pca_data)
+        del(pc)
+        del(pca_contents)
+        print("Successfully imported UKBB PCA data.")
+
+        umap_proj = np.loadtxt(os.path.join(ukbb_dir,"UKBB_UMAP_PC10_NN15_MD0.5_2018328174511"))
+        print("Importing reference file: UKBB_UMAP_PC10_NN15_MD0.5_2018328174511")
+
+        out_dir = ukbb_dir
+    elif dset=='HRS':
+        pca_data_array = np.loadtxt(hrs_path)
+        print("Successfully imported HRS PCA data")
+
[…]

### 2018-12-05 ms/main.tex
add some more plots and graphs, update ms with images

@@
 \begin{abstract}
-%Max 250
-Genetic structure in large cohorts results from technical, sampling and demographic variation.
-Visualisation is therefore a first step in most genomic analyses.
-However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure.
-We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
-Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation.
-This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population.
-This opens new lines of investigation for demographic research and statistical genetics.
-Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
+\adpcomment{Changed the abstract.}
+Genetic structure in cohorts results from technical, sampling and demographic variation, necessitating visualization as an early step in genomic analyses.
+Existing approaches are stymied by the massive size of modern cohorts and unbalanced sizes of populations of many disparate ancestries and struggle to place individuals of admixed ancestries.
+They also struggle to reconcile local genetic relationships in a global context, and the push for larger and more diverse genomic data will only exacerbate these issues.
+We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly reveal cryptic population structure, the flow between layers of diversity, and phenotype heterogeneity between populations.
+Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP intuitively clusters closely related individuals while placing them in a global continuum of genetic variation.
+We uncover previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population, opening new lines of investigation for demographic research and statistical genetics.
+We also demonstrate that in a massive and diverse cohort, UMAP outperforms state-of-the-art dimension reduction.
+Given its small computational cost and promising results, PCA-UMAP should become a standard approach to exploratory analysis of population-scale genetic datasets.
 \end{abstract}
@@
 \end{figure}
+
+\adpcomment{Added this subsection}
+\subsubsection*{Performance of t-SNE and UMAP}
+To compare UMAP to t-SNE, we ran Python's sklearn implementation of t-SNE on the top 10 PCs of each dataset, once using the default initialization, and once initializing it with a two-dimensional UMAP embedding and skipping the early exaggeration phase of t-SNE (typically the first 250 iterations).
+The two approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).
+Using the UKBB, it becomes clear that UMAP is far better equipped to handle massive genomic datasets with individuals from a variety of ancestral populations.
+Looking at the Kullback-Leibler divergence after 20,000 iterations, t-SNE's default initialization did not improve upon initialization with the UMAP embedding (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
+Qualitatively the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
+Given the many desirable properties of UMAP, such computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.
 
@@
     \begin{subfigure}{0.95\textwidth}
-    %\includegraphics[width=0.95\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_2018330153123_RACE_HISP_MEX.pdf}
     \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_race_hisp_mex_labels.pdf}
@@
 \end{figure*}
-
-%\begin{figure*}
-%    \centering
-%    \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_race_hisp_mex_labels.jpeg}
-%    \end{subfigure}
-%    \caption{}
-%    \label{fig:supp_hrs_projection_2}
-%\end{figure*}
 
@@
     \begin{subfigure}{\textwidth}
-    %\includegraphics[width=\textwidth]{images/HRS_HISP_UMAP_PC7_NN15_MD05_admix.pdf}
     \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
@@
 \end{figure*}
-
-%\begin{figure*}
-%    \centering
-%    \begin{subfigure}{\textwidth}
-%    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
-%    \end{subfigure}
-%    \caption{UMAP of the first $7$ principal components of the Hispanic population of the HRS, colored by birth region, with selected 1KGP populations projected onto the UMAP embedding.
-The highlighted area consists almost entirely of individuals born in the Mountain region of the census (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
-The six populations do not fall onto the highlighted cluster.
-CLM, Colombian in Medellin, Colombia;
-%    IBS, Iberian in Spain;
-%    MXL, Mexican in Los Angeles, California;
-%    PEL, Peruvian;
-%    PUR, Puerto Rican;
-%    TSI, Tuscani in Italy.}
-%    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
-%\end{figure*}
 
@@
 
-%\begin{figure*}
-%    \centering
-%    \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
-%    \end{subfigure}
-%    \caption{UMAP on the first $10$ principal components of HRS data, coloring individuals by estimated admixture from three ancestral populations.
-Alternate projection to demonstrate how individuals may fall between clusters.}
-%    \label{fig:supp_hrs_admix_alt}
-%\end{figure*}
-
 \begin{figure}[!htb]
@@
 
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/1KGP_tsne_umap.jpeg}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP.}
+    \label{fig:supp_tsne_umap_compare_1kgp}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/HRS_tsne_umap.jpeg}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the HRS.}
+    \label{fig:supp_tsne_umap_compare_hrs}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/ukbb_tsne_umap.jpeg}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the UKBB.}
+    \label{fig:supp_tsne_umap_compare_ukbb}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_1kgp.jpeg}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration. Done on the 1KGP dataset with 5000 iterations.}
+    \label{fig:supp_tsne_umap_compare_1kgp_graph}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_hrs.jpeg}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration. Done on the HRS dataset with 5000 iterations.}
+    \label{fig:supp_tsne_umap_compare_hrs_graph}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_ukbb.jpeg}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration. Done on the UKBB dataset with 20000 iterations.}
+    \label{fig:supp_tsne_umap_compare_ukbb_graph}
+\end{figure}
+
 \end{document}

### 2018-12-06 ms/cover_letter.tex
minor tweak
update cover letter

@@
 \\\\
-We are pleased to submit our manuscript ``UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts'' for publication in eLife.
-This work applies a newly developed dimensional reduction method called UMAP to visualize population structure in genomic datasets.
+We are pleased to submit our manuscript ``UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts'' for publication in PLOS Genetics.
+Using a newly developed dimensional reduction method called UMAP, this work addresses the pressing issue of quickly and succinctly identifying population structure in massive genomic datasets comprised of many layers of unbalanced and diverse sub-populations.
+Given the demand for and supply of larger and more ethnically diverse data, this approach will soon be a crucial early step in genomic analyses.
 \\\\
@@
 The manuscript presents new methodology and applications that reveal fine-scale genomic and phenotypic variation in large-scale cohorts.
-We believe that this will be of interest to a broad variety of disciplines including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for eLife.
+We believe that this will be of interest to a broad variety of disciplines including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
 \\
@@
 \\
-Simon Gravel and Alex Diaz-Papkovich
+Alex Diaz-Papkovich and Simon Gravel
 \\

### 2018-12-06 ms/main.tex
edited discussion of convergence

@@
 \adpcomment{Added this subsection}
-\subsubsection*{Performance of t-SNE and UMAP}
-To compare UMAP to t-SNE, we ran Python's sklearn implementation of t-SNE on the top 10 PCs of each dataset, once using the default initialization, and once initializing it with a two-dimensional UMAP embedding and skipping the early exaggeration phase of t-SNE (typically the first 250 iterations).
-The two approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).
-Using the UKBB, it becomes clear that UMAP is far better equipped to handle massive genomic datasets with individuals from a variety of ancestral populations.
-Looking at the Kullback-Leibler divergence after 20,000 iterations, t-SNE's default initialization did not improve upon initialization with the UMAP embedding (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
-Qualitatively the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
-Given the many desirable properties of UMAP, such computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.
+\subsubsection*{Comparing t-SNE and UMAP}
+Assessing the relative merits of data visualization methods is notoriously difficult, because different users might give different importance to different features of the data.
+Methods differ in their choice of a similarity metric between high- and low-dimensional representation, which are hard to assess,
+ but also in their strategy to optimize the metrics, which are amenable to direct comparison.
+This is especially true with UMAP and t-SNE which, despite different mathematical motivations, have computationally similar similarity matrices.
+Given that t-SNE generated a smattering of clusters for the minority populations when applied to the UKBB, we hypothesized that this was due to a convergence issue rather than a property of the global optimum of the t-SNE metric.
+
+To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a UMAP run with \sgcomment{x epochs}.
+The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).
+When applied to the UKbiobank, we find (perhaps subjectively) that UMAP provides a more compelling visualization of the data compared to both version of t-SNE.
+We also find that the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum.
+Qualitatively, the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity between clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
+Given the many desirable properties of UMAP, such as computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.
 

### 2018-12-06 ms/main.tex
edited abstract

@@
 Genetic structure in cohorts results from technical, sampling and demographic variation, necessitating visualization as an early step in genomic analyses.
-Existing approaches are stymied by the massive size of modern cohorts and unbalanced sizes of populations of many disparate ancestries and struggle to place individuals of admixed ancestries.
-They also struggle to reconcile local genetic relationships in a global context, and the push for larger and more diverse genomic data will only exacerbate these issues.
-We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly reveal cryptic population structure, the flow between layers of diversity, and phenotype heterogeneity between populations.
+Existing approaches are stymied by the massive size of modern cohorts and unbalanced sizes of populations of many disparate ancestries.
+They also struggle to reconcile local genetic relationships with the global context.
+The push for larger and more diverse genomic data will only exacerbate these issues.
+We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly reveal cryptic population structure, the flow between layers of diversity\sgcomment{What does that even mean?}, and phenotype heterogeneity between populations.
 Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP intuitively clusters closely related individuals while placing them in a global continuum of genetic variation.
 We uncover previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population, opening new lines of investigation for demographic research and statistical genetics.
-We also demonstrate that in a massive and diverse cohort, UMAP outperforms state-of-the-art dimension reduction.
-Given its small computational cost and promising results, PCA-UMAP should become a standard approach to exploratory analysis of population-scale genetic datasets.
+We also demonstrate that in a massive and diverse cohort, UMAP outperforms state-of-the-art dimension reduction methods for both global convergence and preservation of qualitative features of the data.
+Given its small computational cost and promising results, we think that UMAP should be run in tandem with PCA as a standard initial analysis of population-scale genetic datasets.
 \end{abstract}

### 2018-12-06 ms/cover_letter.tex
edited for concreteness

@@
 \\\\
-We are pleased to submit our manuscript ``UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts'' for publication in PLOS Genetics.
-Using a newly developed dimensional reduction method called UMAP, this work addresses the pressing issue of quickly and succinctly identifying population structure in massive genomic datasets comprised of many layers of unbalanced and diverse sub-populations.
-Given the demand for and supply of larger and more ethnically diverse data, this approach will soon be a crucial early step in genomic analyses.
+We are pleased to submit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics.
+Using a newly developed dimensional reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in massive and heterogeneous genomic datasets.
+In contrast to existing methods, this approach is able to represent both discrete and continuous population structure, as well as admixture.
 \\\\
 By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts.
-For example, we identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region, who are likely descendants of the Hispano population.
-We also identify multiple associations between genetics, geography, and phenotypes in the UK biobank and sub-continental population structure within 1000 Genomes Project populations.
+For example, we identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region.
+These individuals are likely descendants of the Hispano population.
+We also identify multiple associations between genetics, geography, and phenotypes in the UK biobank and population structure within 1000 Genomes Project populations.
 \\\\

### 2018-12-06 ms/main.tex
shift charts, revert abstract, add scRNA seq citation

--- 2018-12-06 ms/main.tex
@@
 \begin{abstract}
-\adpcomment{Changed the abstract.}
-Genetic structure in cohorts results from technical, sampling and demographic variation, necessitating visualization as an early step in genomic analyses.
-Existing approaches are stymied by the massive size of modern cohorts and unbalanced sizes of populations of many disparate ancestries and struggle to place individuals of admixed ancestries.
-They also struggle to reconcile local genetic relationships in a global context, and the push for larger and more diverse genomic data will only exacerbate these issues.
-We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly reveal cryptic population structure, the flow between layers of diversity, and phenotype heterogeneity between populations.
-Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP intuitively clusters closely related individuals while placing them in a global continuum of genetic variation.
-We uncover previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population, opening new lines of investigation for demographic research and statistical genetics.
-We also demonstrate that in a massive and diverse cohort, UMAP outperforms state-of-the-art dimension reduction.
-Given its small computational cost and promising results, PCA-UMAP should become a standard approach to exploratory analysis of population-scale genetic datasets.
+Genetic structure in large cohorts results from technical, sampling and demographic variation.
+Visualisation is therefore a first step in most genomic analyses.
+However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure.
+We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
+Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation.
+This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population.
+This opens new lines of investigation for demographic research and statistical genetics.
+Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
 \end{abstract}
@@
 
-Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space \cite{2018arXivUMAP}.
+Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}.
 The assumption behind UMAP is that data are uniformly distributed on local manifolds in high-dimensional space, which can be approximated as fuzzy sets that are patched together to form a topological representation.
@@
 With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds.
-The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.
+The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.\adpcomment{added this sentence.} The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.
 
@@
 \centering
-   %\includegraphics[width=0.6\linewidth]{images/HRS_UMAP_PC10_NN15_MD05_admix_crop.pdf}
    \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
@@
 \end{figure}
-
-%\subsection*{Results}
-%\begin{figure}
-%\centering
-%   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_highlighted.pdf}
-%    \caption{The top $7$ principal components of the Hispanic population of the HRS projected by UMAP, colored by region of birth.
-The highlighted region consists almost entirely of individuals who were born in the Mountain region of the United States.
-This region contains New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana.
-Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same figure colored according to estimated continental ancestry proportion.
-Figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} shows that populations from the 1KGP do not map to the cluster when projected onto the UMAP embedding.}
-%    \label{fig:hrs_hisp}
-%\end{figure}
 
@@
 
-\adpcomment{Added this subsection}
 \subsubsection*{Comparing t-SNE and UMAP}
@@
     \includegraphics[width=0.95\columnwidth]{images/1KGP_tsne_umap.jpeg}
-    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP.}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP. t-SNE used 5000 iterations.}
     \label{fig:supp_tsne_umap_compare_1kgp}
@@
     \includegraphics[width=0.95\columnwidth]{images/HRS_tsne_umap.jpeg}
-    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the HRS.}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the HRS. t-SNE used 5000 iterations.}
     \label{fig:supp_tsne_umap_compare_hrs}
@@
     \includegraphics[width=0.95\columnwidth]{images/ukbb_tsne_umap.jpeg}
-    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the UKBB.}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the UKBB. t-SNE used 20000 iterations}
     \label{fig:supp_tsne_umap_compare_ukbb}
@@
     \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_1kgp.jpeg}
-    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration. Done on the 1KGP dataset with 5000 iterations.}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
+Done on the 1KGP dataset with 5000 iterations.
+The UMAP-initialized graph has been shifted by 600 iterations to approximate the 600 epochs UMAP uses for small datasets ($n\leq10,000$).}
     \label{fig:supp_tsne_umap_compare_1kgp_graph}
@@
     \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_hrs.jpeg}
-    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration. Done on the HRS dataset with 5000 iterations.}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
+Done on the HRS dataset with 5000 iterations.
+The UMAP-initialized graph has been shifted by 230 iterations to approximate the 230 epochs UMAP uses for large datasets ($n>10,000$).}
     \label{fig:supp_tsne_umap_compare_hrs_graph}
@@
     \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_ukbb.jpeg}
-    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration. Done on the UKBB dataset with 20000 iterations.}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
+Done on the UKBB dataset with 20000 iterations.
+The UMAP-initialized graph has been shifted by 230 iterations to approximate the 230 epochs UMAP uses for large datasets ($n>10,000$).}
     \label{fig:supp_tsne_umap_compare_ukbb_graph}

### 2018-12-07 ms/main.tex
abstract touchup

@@
 However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure.
-We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
+We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly illustrate population structure in large cohorts and capture their relationships on local and global scales.
 Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation.

### 2018-12-07 ms/cover_letter.tex
cover letter edits: added tsne

@@
 \\\\  
-This work has broad applicability.
-Our approach has already been implemented as a visualization tool by leading cohorts such as GnomAD\footnote{Fancioli, Laurent.
+In other -omics fields, such as single-cell expression data, nonlinear nonparametric approaches such as t-SNE have become standard tools for data exploration.
+Principal component analysis has remained the standard for dimension reduction in human genetics, partly because nonlinear alternatives such as t-SNE don't behave well when applied to large-scale genetic data.
+We show that UMAP outperforms t-SNE computationally, qualitatively and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
+\\\\
+In the few weeks since our preprint was published, leading cohorts such as GnomAD have started using UMAP to represent their data\footnote{Fancioli, Laurent.
 ``gnomAD v2.1''.
-https://macarthurlab.org/2018/10/17/gnomad-v2-1/ (2018-10-17).}, and used as a preprocessing step for machine learning approaches\footnote{Tonkin-Hill, Gerry, et al.
+https://macarthurlab.org/2018/10/17/gnomad-v2-1/ (2018-10-17)}, and it has also been integrated as a preprocessing step in Machine Learning approaches\footnote{Tonkin-Hill, Gerry, et al.
 ``Fast Hierarchical Bayesian Analysis of Population Structure.'' bioRxiv (2018): 454355.}.
 Our preprint is also in the 99th percentile of all-time bioRxiv preprints for the Altmetric impact statistic.
-Because of its extreme computational tractability and its ability to efficiently summarize genomic information, we expect that our approach will become a new standard tool in data analysis for large genomic studies.
 \\\\
-The manuscript presents new methodology and applications that reveal fine-scale genomic and phenotypic variation in large-scale cohorts.
-We believe that this will be of interest to a broad variety of disciplines including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
+We believe that this will be useful and of interest to a broad variety of geneticists including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
 \\

### 2018-12-07 ms/main.tex
rm comments, details in umap vs tsne pgh

@@
 With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds.
-The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.\adpcomment{added this sentence.} The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.
+The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.
+The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.
 
@@
 
-To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a UMAP run with \sgcomment{x epochs}.
+To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data).
 The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).
@@
 Qualitatively, the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity between clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
+An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files.
 Given the many desirable properties of UMAP, such as computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.

### 2018-12-11 ms/main.tex
match formatting, add author summary

@@
 \begin{abstract}
+{\large Abstract.}
+
 Genetic structure in large cohorts results from technical, sampling and demographic variation.
@@
 
-\subsection*{Background}
+\subsection*{Author summary}
+%We ask that all authors of research articles include a 150?200 word non-technical summary of the work as part of the manuscript to immediately follow the abstract.
+This text is subject to editorial change, should be written in the first-person voice, and should be distinct from the scientific abstract.
+
+%Aim to highlight where your work fits within a broader context; present the significance or possible implications of your work simply and objectively; and avoid the use of acronyms and complex terminology wherever possible.
+The goal is to make your findings accessible to a wide audience that includes both scientists and non-scientists.
+
+%Authors may benefit from consulting with a science writer or press officer to ensure they effectively communicate their findings to a general audience.
+
+Population structure is a systematic difference in allele frequencies between populations.
+It is a consequence of non-random mating from factors such as geographic isolation and founder effects from migrations, and is a common confounder in genetic studies.
+We use a newly developed method (UMAP) to visualize population structure from three genomic datasets.
+Using genotype data alone, we reveal numerous patterns that relate population structure and ancestry to traits such as white blood cell count, height, and FEV1, a baseline measure for airway obstruction.
+We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure, and that when working in higher dimensions it can be used to study, for example, the spatial distribution of genetic relatedness\adpcomment{ (talking about the UK map here)}.
+We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries.
+The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions, and the images generated will pique the interest of general audiences.
+
+\subsection*{Introduction}
 
@@
 
+\subsection*{Results}
 \subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}.

### 2018-12-11 ms/main.tex
small edits to summary

@@
 %Authors may benefit from consulting with a science writer or press officer to ensure they effectively communicate their findings to a general audience.
-
-Population structure is a systematic difference in allele frequencies between populations.
-It is a consequence of non-random mating from factors such as geographic isolation and founder effects from migrations, and is a common confounder in genetic studies.
+Because of geographic isolation, individuals tend to be more genetically related to people living nearby than to people living far.
+This is an example of population structure, a situation where a large population contains subgroups that share more than the average amount of DNA.
+This structure can tell us about human history, and it can also have a large effect on medical studies.
 We use a newly developed method (UMAP) to visualize population structure from three genomic datasets.
-Using genotype data alone, we reveal numerous patterns that relate population structure and ancestry to traits such as white blood cell count, height, and FEV1, a baseline measure for airway obstruction.
-We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure, and that when working in higher dimensions it can be used to study, for example, the spatial distribution of genetic relatedness\adpcomment{ (talking about the UK map here)}.
+Using genotype data alone, we reveal numerous subgroups related to ancestry and correlated with traits such as white blood cell count, height, and FEV1, a baseline measure for airway obstruction.
+We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure.
 We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries.

### 2018-12-11 ms/main.tex
tweak sentence

@@
 We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries.
-The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions, and the images generated will pique the interest of general audiences.
+The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions, and for conveying concepts to wider audiences.
 

### 2018-12-11 ms/main.tex
quick edit in summary

@@
 We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries.
-The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions, and for conveying concepts to wider audiences.
+The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions.
 

### 2018-12-11 ms/main.tex
quick edit 2 in summary

@@
 We use a newly developed method (UMAP) to visualize population structure from three genomic datasets.
-Using genotype data alone, we reveal numerous subgroups related to ancestry and correlated with traits such as white blood cell count, height, and FEV1, a baseline measure for airway obstruction.
+Using genotype data alone, we reveal numerous subgroups related to ancestry and correlated with traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction.
 We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure.

### 2019-03-26 ms/main.tex
add disco

@@
 \templatetype{pnasresearcharticle} % Choose template
-\title{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}
+\title{Beyond PCA: UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}
 
@@
 
-Genetic structure in large cohorts results from technical, sampling and demographic variation.
-Visualisation is therefore a first step in most genomic analyses.
-However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure.
+Genetic structure in large cohorts results from technical, sampling, and demographic variation.
+Identifying structure through visualization is typically the first step in most genomic analyses.
+Existing exploration methods struggle with unbalanced sampling and the many scales of population structure.
+
 We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly illustrate population structure in large cohorts and capture their relationships on local and global scales.
@@
 UMAP condenses these components and comprehensively illustrates information --- phenotypic, geographic, and ancestral --- contained within genotypes on fine-scale levels and within the context of a global population structure.
-In large datasets where the number of significant PCs is large, the resulting representation has important advantages over
-PCA alone and provides a superior visualization to t-SNE.
+In large datasets where the number of significant PCs is large, the resulting representation has important advantages over PCA alone and provides a superior visualization to t-SNE.
 
@@
 Given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
+
+%added new stuff here on cautions and how to appropriately use UMAP
+Like most non-linear methods, UMAP lacks direct interpretability.
+The dimensions in its embeddings have no inherent meaning, unlike PCA where the axes correspond to the proportion of variance explained.
+UMAP is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are very distant from each other cannot be considered to be more different from each other.
+Since the low dimensional embedding is formed by patching together many neighbourhoods there is a semblance of a global structure -- one that is superior to methods like t-SNE; however, the developers of UMAP caution that researchers interested primarily in global structure may wish to consider other methods.
+
+% mention genealogical interpretation here?
+% making sure I have the wording right....
+% what I'd like to highlight: UMAP finds local similarities in populations. even in the HRS
+% hispanic population we see a cluster appear
+% basically you can find groups that seem to be uniquely closely related within a cluster
+% you can't really say "how different" they are from others though
+UMAP works well at highlighting groups of individuals who share more ancestry within their group, as evidenced by the Hispanic individuals in the HRS and the many 1KGP populations.
+When clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters.
+The 1KGP dataset was deliberately crafted to come from geographically separated populations that are relatively homogeneous internally.
+For this reason most of its populations form isolated clusters, with the exception of those that most closely share ancestries (Iberians and Toscani, Han Chinese and Southern Han Chinese).
+In figure~\ref{fig:supp_megamontage_pc10_50}, population clusters bounce around in each projection, again highlighting that the positions of disconnected clusters relative to one another are not meaningful.
+
+% with the UKBB, the geography colouring is only possible with UMAP
+% tSNE would just break apart into tiny islands and scatter the people everywhere else.
+UMAP excels with larger datasets containing individuals with admixed backgrounds.
+Unlike t-SNE, it patches together many local neighbourhoods, resulting in a relatively continuous flow to the data.
+t-SNE can generate seemingly arbitrary clusters, as in figures \ref{fig:supp_tsne_umap_compare_ukbb} and \ref{fig:supp_ukbb_tsne}, where small islands of ethnic groups are scattered across the projection.
+These are undesirable when looking at colourings by phenotypes, geography, or admixture, where we are interested in spotting gradients across populations.
+This continuity
+
+We suggest making several runs of UMAP and to highlight consistent qualitative results.
+Tuning parameters will generally not make a difference beyond making some patterns easier to see.
+Selecting a lower or higher number of neighbours
 

### 2019-04-02 ms/main.tex
ms tweaks, add figure with COB in UKBB

@@
 \templatetype{pnasresearcharticle} % Choose template
-\title{Beyond PCA: UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}
+\title{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}
 
@@
 \author[b,c]{Luke Anderson-Trocm\'e}
+\author[b,c]{Chief Ben-Eghan}
 \author[b,c,1]{Simon Gravel} 
@@
 \subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}.
-Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and PCA-UMAP (that is, UMAP with PCA pre-processing).
+Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing.
 Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters.
-Using either on the top principal components leads to more distinct population clusters and less defined continental structure (see figure \ref{fig:supp_megamontage_pc10_50} for PCA-tSNE).
+Using either on the top principal components leads to more distinct population clusters and less defined continental structure (see figure \ref{fig:supp_megamontage_pc10_50} for tSNE).
 Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; further components gradually approach results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9} and~\ref{fig:supp_megamontage_pc10_50}).
 
-Focusing on PCA-UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we also find several population clusters that reflect shared ancestries.
+Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we also find several population clusters that reflect shared ancestries.
 British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry.
@@
 
-\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying PCA-UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
+\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
 One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
@@
     \includegraphics[width=.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
-    \caption{The UKBB projected onto two dimensions using PCA-UMAP with each individual colored by their geographical coordinates of residence.
+    \caption{The UKBB projected onto two dimensions using UMAP with each individual colored by their geographical coordinates of residence.
 Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain.
@@
     \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
-    \caption{The UKBB projected onto two dimensions using PCA-UMAP (as in~\ref{fig:umap_ukbb_pc10}), with females colored by age-adjusted difference from mean population height (left) and leukocyte counts (right).
+    \caption{The UKBB projected onto two dimensions using UMAP (as in~\ref{fig:umap_ukbb_pc10}), with females colored by age-adjusted difference from mean population height (left) and leukocyte counts (right).
 To protect participant privacy, data has been randomized as explained in the materials and methods section.}
@@
 Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; while another close to European and British populations has lower values relative to the population mean.
-These results merit further investigation and underscore the exploratory value of PCA-UMAP --- these  populations are largely unidentified and there is no straightforward way to separate them within the data otherwise.
+Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; using available country of birth data, most of the unidentified ethnic group were born in Japan.
+These results merit further investigation and underscore the exploratory value of UMAP --- these  populations are largely unidentified and there is no straightforward way to separate them within the data otherwise.
 
@@
 
-Application to the UKBB underscores the strength of PCA-UMAP in large cohorts.
+Application to the UKBB underscores the strength of UMAP in large cohorts.
 We see clear, fine relationships between genotype and phenotype and geography, and this is presented in a visualization that accounts for natural genetic clustering.
@@
 The dimensions in its embeddings have no inherent meaning, unlike PCA where the axes correspond to the proportion of variance explained.
+Approaches like PCA will keep more distantly related populations further from each other, but in choosing which dimensions to examine, we are forced to flatten dimensions and clump together groups such as Central and South Americans with South Asians as in figure~\ref{fig:1000g_compare}.
 UMAP is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are very distant from each other cannot be considered to be more different from each other.
@@
 
-% mention genealogical interpretation here?
-% making sure I have the wording right....
 % what I'd like to highlight: UMAP finds local similarities in populations. even in the HRS
@@
 In figure~\ref{fig:supp_megamontage_pc10_50}, population clusters bounce around in each projection, again highlighting that the positions of disconnected clusters relative to one another are not meaningful.
+For this reason, UMAP coordinates should not be used, for example, as covariates in a GWAS or calculating distances between populations.
 
@@
 % tSNE would just break apart into tiny islands and scatter the people everywhere else.
-UMAP excels with larger datasets containing individuals with admixed backgrounds.
+UMAP excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
 Unlike t-SNE, it patches together many local neighbourhoods, resulting in a relatively continuous flow to the data.
@@
 These are undesirable when looking at colourings by phenotypes, geography, or admixture, where we are interested in spotting gradients across populations.
-This continuity
 
+% Note: the HRS data didn't really "converge" as nicely as 1KGP data with its subcontinental structure. PCs may serve a dual role to filter data?
 We suggest making several runs of UMAP and to highlight consistent qualitative results.
 Tuning parameters will generally not make a difference beyond making some patterns easier to see.
-Selecting a lower or higher number of neighbours
+The number of principal components used will have an effect on the visualization depending on the variable of interest.
+In each dataset, the groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
+Geographical coordinates in the UKBB continued to concentrate even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
+The total number of components to use is limited primarily by computational power.
 
@@
 provided the UK map visualization and analysis.
+C.B.E.
+prepared several UKBB datasets and principal component vectors.
 
@@
 We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the dataset.
-We also thank Chief Ben-Eghan, Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access, and Selin Jessa for introducing us to UMAP.
+We also thank Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access, and Selin Jessa for introducing us to UMAP.
 This research has been conducted using the UK Biobank Resource under Application Number 6728.
@@
     \end{subfigure}
-    \caption{PCA-UMAP on UKBB data, colored by self-identified ethnic background. Images are labelled by the number of components included.}
+    \caption{UMAP on UKBB data, colored by self-identified ethnic background. Images are labelled by the number of components included.}
     \label{fig:supp_montage_ukbb_eth}
@@
     \end{subfigure}
-    \caption{PCA-UMAP on UKBB data, colored by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates.
+    \caption{UMAP on UKBB data, colored by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates.
 Images are labelled by the number of components included.
@@
     \end{subfigure}
-    \caption{PCA-UMAP on UKBB data, colored by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates.
+    \caption{UMAP on UKBB data, colored by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates.
 Images are labelled by the number of components included.
@@
 
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_cob_resized.jpeg}
+    \end{subfigure}
+    \caption{Using country of birth data, some of the larger unidentified groups from figure~\ref{fig:supp_ukbb_zoom} were identified as being born mostly in Japan, the Philippines, North Africa, the Middle East, and Central and South America.
+The large cluster of "Any other Asian Background" were mostly born in Sri Lanka.}
+    \label{fig:supp_ukbb_cob}
+\end{figure*}
+
 %%%%% UKBB phenotype plots
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by basophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_basophill_f}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by basophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_basophill_m}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_eosinophill_f}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_eosinophill_m}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_fev_f}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_fev_m}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by height (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by height (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_height_f}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by height (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by height (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_height_m}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_leukocyte_f}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_leukocyte_m}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_neutrophill_f}
@@
     %\end{subfigure}
-    \caption{PCA-UMAP on the top 10 principal components of the UKBB colored by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_neutrophill_m}

### 2019-04-04 ms/main.tex
add 1KGP convergence image and 23andme cite

@@
 Using either on the top principal components leads to more distinct population clusters and less defined continental structure (see figure \ref{fig:supp_megamontage_pc10_50} for tSNE).
-Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; further components gradually approach results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9} and~\ref{fig:supp_megamontage_pc10_50}).
+Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; further components gradually approach results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
 
@@
 This clustering is robust to, e.g., the choice of the number of PCs considered (see figure ~\ref{fig:supp_megamontage_pc10_50}).
+Analysis by 23andMe determined the distinct Gujarati cluster mostly have the last name Patel and suggested their grouping is the result of endogamy~\cite{23andme}.
 
@@
     \label{fig:supp_megamontage_pc10_50}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.95\textwidth}
+    \includegraphics[width=0.95\textwidth]{images/montage_1KGP_umap_convergence_resize.jpeg}
+    \end{subfigure}
+    \caption{UMAP applied to the first few hundred principal components of the 1KGP data.
+As more components are added, the figure begins to resemble that of UMAP carried out on the full genotype dataset.}
+    \label{fig:supp_montage_1kgp_converge}
 \end{figure*}

### 2019-04-10 ms/main.tex
update MS with new figures and text

@@
 
+%An Analysis is a new analysis of existing data (typically large genomic, transcriptomic or proteomic data sets from arrays or other high-throughput platforms) or new data obtained in a comparative analysis of technologies that lead to novel and arresting conclusions of importance to a broad audience.
+The main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words.
+The abstract is typically 100-150 words, unreferenced.
+Analyses have no more than 6 display items (figures and/or tables).
+An introduction (without heading) is followed by sections headed Results, Discussion and online Methods.
+The Results and online Methods should be divided by topical subheadings; the Discussion does not contain subheadings.
+As a guideline, Analysis allow up to 50 references.
+
 \begin{abstract}
 {\large Abstract.}
-
-Genetic structure in large cohorts results from technical, sampling, and demographic variation.
-Identifying structure through visualization is typically the first step in most genomic analyses.
-Existing exploration methods struggle with unbalanced sampling and the many scales of population structure.
-
-We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly illustrate population structure in large cohorts and capture their relationships on local and global scales.
-Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation.
-This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population.
-This opens new lines of investigation for demographic research and statistical genetics.
-Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
+%The abstract is typically 100-150 words, unreferenced. 
+Improvements in sequencing have generated massive and diverse genomic datasets of populations.
+Dimension reduction is necessary to condense and analyze the data.
+Uniform manifold approximation and projection (UMAP) is a non-linear dimension reduction tool recently made available.
+Here we apply UMAP to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data.
+We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry population-scale datasets.
 \end{abstract}
-
-\keywords{genomics $|$ genetics $|$ population structure $|$ ethnicity $|$ ancestry $|$ machine learning $|$ data visualization $|$ dimension reduction}
 
@@
 \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
-\subsection*{Author summary}
-%We ask that all authors of research articles include a 150?200 word non-technical summary of the work as part of the manuscript to immediately follow the abstract.
-This text is subject to editorial change, should be written in the first-person voice, and should be distinct from the scientific abstract.
-
-%Aim to highlight where your work fits within a broader context; present the significance or possible implications of your work simply and objectively; and avoid the use of acronyms and complex terminology wherever possible.
-The goal is to make your findings accessible to a wide audience that includes both scientists and non-scientists.
-
-%Authors may benefit from consulting with a science writer or press officer to ensure they effectively communicate their findings to a general audience.
-Because of geographic isolation, individuals tend to be more genetically related to people living nearby than to people living far.
-This is an example of population structure, a situation where a large population contains subgroups that share more than the average amount of DNA.
-This structure can tell us about human history, and it can also have a large effect on medical studies.
-We use a newly developed method (UMAP) to visualize population structure from three genomic datasets.
-Using genotype data alone, we reveal numerous subgroups related to ancestry and correlated with traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction.
-We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure.
-We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries.
-The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions.
-
-\subsection*{Introduction}
-
 Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies.
@@
 Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations.
-If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$.
+If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$.
 Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.
@@
 Positions of individuals along directions of highest variance can then be used to summarize individual genotypes.
-PCA coordinates have natural genealogical interpretations in terms of times to a most recent common ancestor (TMRCA) \cite{mcvean2009genealogical}, and are used empirically to reveal admixture \cite{brisbin2012pcadmix}, continuous isolation-by-distance \cite{novembre2008europe, nelson2008population}, as well as technical artefacts.
+PCA coordinates have natural genealogical interpretations in terms of expected times to a most recent common ancestor (TMRCA) \cite{mcvean2009genealogical}, and are used empirically to reveal admixture \cite{brisbin2012pcadmix}, continuous isolation-by-distance \cite{novembre2008europe, nelson2008population}, as well as technical artefacts.
 PCA coordinates are particularly well-suited to correct for population structure in GWAS~\cite{eigen2006}.
@@
 Axes in UMAP and t-SNE are arbitrary.
+Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
     ACB, African Caribbean in Barbados;
@@
 
-As sample sizes increase, the amount of information encoded in the lower-variance principal components increases, and researchers typically examine multiple two-dimensional projections to get a sense of the data.
-While many features of the data can be identified in this manner, other features may be hidden by the projections or hard to interpret.
-
+The amount of information encoded in the lower-variance principal components increases with sample size, so researchers typically examine multiple two-dimensional projections to explore data.
+Some features may be hidden by the projections or hard to interpret.
 To display more of the high-dimensional features of the data in a two dimensional figure, we can use non-linear transformations that seek to preserve the local structure of the data.
 A popular method for visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}.
-Whereas PC projection is designed to capture as much variance as possible for a linear transformation, t-SNE seeks a low-dimensional representations of the data that preserves pairwise distances, using a probabilistic heuristic to weight mismatches.
-t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}.
-Using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}.
+t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}; using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}.
 However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.
 
-Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}.
-The assumption behind UMAP is that data are uniformly distributed on local manifolds in high-dimensional space, which can be approximated as fuzzy sets that are patched together to form a topological representation.
-One can then construct a low-dimensional topological representation that minimizes the differences between the two representations.
+Uniform Manifold Approximation and Projection (UMAP) is a dimension reduction technique designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}.
 With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds.
@@
 
-A common practice in dimensional reduction is to first apply PCA to reduce the number of dimensions before performing nonlinear dimensional reduction.
-In addition to being computationally advantageous, this discards statistical noise that can confound nonlinear approaches: Population structure arising from $n$ isolated randomly-mating demes can be described by the leading $n-1$ PCs, with the following PCs describing stochastic variation in relatedness~\cite{eigen2006}.
+A common practice in dimensional reduction is to first apply PCA.
+In addition to being computationally advantageous, this discards noise that can confound nonlinear approaches: population structure arising from $n$ isolated randomly-mating demes can be described by the leading $n-1$ PCs, with the following PCs describing stochastic variation in relatedness~\cite{eigen2006}.
 Selecting the leading PCs therefore has potential to extract meaningful population structure while filtering out stochastic noise.
-
-We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}.
-
-\begin{figure}
-\centering
-   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
-   \caption{UMAP on the first 10 principal components of HRS data.
-Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
-This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
-    \label{fig:umap_hrs_admix}
-\end{figure}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.6\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
-    \end{subfigure}
-    \caption{Applying UMAP to the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
-Coloring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
+We explore different strategies to pre-process the data and investigate discrete and continuous population structure patterns present in large datasets of human genotypes: the 1KGP, the Health and Retirement Study (HRS)\cite{juster1995overview}, and the UK BioBank (UKBB)\cite{sudlow2015uk}, and compare UMAP's performance to t-SNE.
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.5\columnwidth}
+    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
+    \caption{Subset of Hispanic population of the HRS}
+    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
+    \end{subfigure}\hfill
+        \begin{subfigure}{0.5\columnwidth}
+    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
+    \caption{Subset of the Asian population of the UKBB}
+        \label{fig:umap_ukbb_indian}
+    \end{subfigure}
+    \caption{Applying UMAP to subets of data can reveal deep population structure.
+(a) UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
+Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
 When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster.
@@
     PUR, Puerto Rican;
-    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions.}
-    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
+    TSI, Tuscani in Italy.
+Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions.
+(b) UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
+Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
+    \label{fig:subpops}
 \end{figure*}
@@
 Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters.
-Using either on the top principal components leads to more distinct population clusters and less defined continental structure (see figure \ref{fig:supp_megamontage_pc10_50} for tSNE).
-Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; further components gradually approach results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
-
-Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we also find several population clusters that reflect shared ancestries.
+Using either on the top principal components leads to more distinct population clusters and less defined continental structure.
+Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
+Patterns are also evident when using UMAP on the last components (\ref{fig:supp_1kgp_3350}).
+
+Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we find several population clusters that reflect shared ancestries.
 British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry.
@@
 African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster.
-The East Asian super-population forms three sub-populations split by geography: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population.
+Three East Asian groups appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population.
 Looser geographical groupings include Colombians and Peruvians, and the Esan and Yoruba populations of Nigeria; both groupings appear as connected sub-clusters.
-The South Asian super-population also forms a loose grouping.
+The South Asian populations also form a loose grouping.
 
@@
 One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone.
-Only two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan.
-This clustering is robust to, e.g., the choice of the number of PCs considered (see figure ~\ref{fig:supp_megamontage_pc10_50}).
-Analysis by 23andMe determined the distinct Gujarati cluster mostly have the last name Patel and suggested their grouping is the result of endogamy~\cite{23andme}.
-
-Finally, contrasting UMAP and t-SNE, we find that UMAP preserves more of the global structure of the data than t-SNE, and is more robust to choices of data pre-processing (figure~\ref{fig:supp_megamontage_pc10_50}).
-
-\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct population groups across the world, which makes the data particularly easy to cluster.
+Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan.
+This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}).
+Analysis of the cluster by 23andMe suggested the Gujarati grouping is the result of endogamy~\cite{23andme}.
+
+\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form iclusters.
 Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas.
-
 For example, the HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}.
@@
 We have scaled these three proportions to values between 0 and 255, to color individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
-Using the first 10 principal components and UMAP, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.
+Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.
 
@@
 To investigate possible ancestries related to populations in the 1KGP we took two approaches.
-In the first we generated PC axes and a UMAP embedding using UMAP and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}).
+In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}).
 In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}).
@@
 
-\subsubsection*{Regional patterns in the American Hispanic population} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
+\subsubsection*{Deep population structure in the American Hispanic subpopulation} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
 One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
@@
 \begin{subfigure}{.5\columnwidth}
-\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.pdf}%
+\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
 \caption{Principal components 1 and 2}%
@@
 
-The largest body in the figure consists of the White British and Irish populations.
-The Irish population concentrates in a portion of this group, but many individuals are also scattered throughout the British-identified population.
+The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations.
+The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population.
 Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters.
-Chinese individuals form a cluster, within appears to be a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well.
+Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well.
 The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group".
 The patchwork of genetic neighbourhoods is connected by trails of admixed individuals.
-These trails come together in a nexus of individuals with a variety of ethnicities; many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group".
-Although their ethnicities are unknown to us, given their proximity to African, South Asian, and White individuals, possible candidates for these groups are North African, Middle Eastern, and West Asian backgrounds.
-Additionally, there are many individuals whose ethnicity is White but neither British nor Irish (AWB) forming clusters distinct from the British and Irish cluster.
-
-Figure~\ref{fig:umap_ukbb_geo_osgb} presents the projection in figure~\ref{fig:umap_ukbb_pc10} colored in by  geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
-UMAP coordinates within the "White British" cluster broadly map to geographic coordinates, as has been observed in Europe-wide data\cite{novembre2008europe}.
-Most admixture lines connect to the South East corner of this cluster, corresponding to the position of the city of London and reflecting its high migrant population.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
-    \includegraphics[width=.49\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
-    \caption{The UKBB projected onto two dimensions using UMAP with each individual colored by their geographical coordinates of residence.
+These trails come together in a nexus of individuals with a variety of ethnicities.
+Many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group"; some of these were identified using country of birth in figure~\ref{fig:supp_ukbb_cob} and appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
+
+Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection in figure~\ref{fig:umap_ukbb_pc10} coloured in by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
+Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}.
+Most admixture lines have coordinates in the south east of the UK, corresponding to the city of London and reflecting its high migrant population.
+The detailed shape of extended clusters is not stable as we vary the number of PCs included.
+Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.
+The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.
+
+\begin{figure}
+    \centering
+    \begin{subfigure}{\columnwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
+    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
+    \caption{Colouring the projection by geographic coordinates.}
+    \label{fig:umap_ukbb_geo_osgb}
+    \end{subfigure}
+     \begin{subfigure}{\columnwidth}
+         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
+    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
+   \caption{Colouring the projection by phenotypic data}
+    \label{fig:umap_height_female}
+      \end{subfigure}
+      \caption{Using different colourings with the projection in figure~\ref{fig:umap_ukbb_pc10}, relationships between population structure, geography, and phenotypes become clear.
+(Top) Each individual is coloured by their geographical coordinates of residence.
 Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain.
@@
 (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}).
-Individuals with missing geographic data are not shown.
-To prevent outlying individuals from washing out the color scheme, northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km.
-To protect participant privacy, data has been randomized as explained in the materials and methods section.}
-    \label{fig:umap_ukbb_geo_osgb}
-\end{figure}
-
-The detailed shape of extended clusters is not stable as we vary the number of PCs included.
-Figure \ref{fig:supp_montage_ukbb_eth} shows a UMAP plot using the top 20 PCs from the UKBB.
-The shape of the "White British Cluster" is notably different, and we observe finer patterns of geographic variation, yet the qualitative observations made above are maintained.
-As an alternate visualization of diversity's correlation with geography, we performed a 3D UMAP projection and converted the normalized UMAP values into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}).
+Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km.
+(Bottom) Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right).
+Individuals with missing data were excluded.
+To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
+\end{figure}
+
+As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}).
 The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.
-
-Similarly to UMAP, t-SNE applied to the UKBB data both displays diversity within the "White British" population and identifies clusters among other groups.
-However, it has three drawbacks: it is much slower, requiring 2.26 hours for its first thousand iterations alone on 10 principal components against UMAP's 14 minutes; it fails to find a global optimum, which results in a scattering of individuals and groups that are not stable across independent runs; and it does not identify continuity between different continental groups resulting from admixture (figure~\ref{fig:supp_ukbb_tsne}).
-
-\subsubsection*{Patterns in phenotype related to population structure} Our involvement with the UK biobank data is through a project on autoimmune disease and asthma.
-More than in geographic coordinates, we are interested in whether genetic population structure correlates with phenotypes and covariates of interest.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
-    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
-    \caption{The UKBB projected onto two dimensions using UMAP (as in~\ref{fig:umap_ukbb_pc10}), with females colored by age-adjusted difference from mean population height (left) and leukocyte counts (right).
-To protect participant privacy, data has been randomized as explained in the materials and methods section.}
-    \label{fig:umap_height_female}
-\end{figure}
-
-Covariates such as height (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures  (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure.
-Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and several unidentified ethnic groups have noticeably lower-than-average heights.
-More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t-test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
-Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.
-
-Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
-Notably, there appears to be a juncture in the admixture continuum, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes.
-This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds.
-Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; using available country of birth data, most of the unidentified ethnic group were born in Japan.
-These results merit further investigation and underscore the exploratory value of UMAP --- these  populations are largely unidentified and there is no straightforward way to separate them within the data otherwise.
 
@@
 
+Similarly to UMAP, t-SNE applied to the UKBB data both displays diversity within the "White British" population and identifies clusters among other groups.
+However, it has three drawbacks: it is much slower, requiring 2.26 hours for its first thousand iterations alone on 10 principal components against UMAP's 14 minutes; it fails to find a global optimum, which results in a scattering of individuals and groups that are not stable across independent runs; and it does not identify continuity between different continental groups resulting from admixture (figure~\ref{fig:supp_ukbb_tsne}).
+
+\subsubsection*{Patterns in phenotype related to population structure} 
+Covariates such as height and leukocyte count (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures  (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure.
+Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and others have noticeably lower-than-average heights.
+More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t-test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
+Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.
+
+Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
+Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes.
+This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds.
+Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
+Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; using available country of birth data, most of the unidentified ethnic group were born in Japan.
+These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).
+
 \subsubsection*{Comparing t-SNE and UMAP}
-Assessing the relative merits of data visualization methods is notoriously difficult, because different users might give different importance to different features of the data.
-Methods differ in their choice of a similarity metric between high- and low-dimensional representation, which are hard to assess,
- but also in their strategy to optimize the metrics, which are amenable to direct comparison.
-This is especially true with UMAP and t-SNE which, despite different mathematical motivations, have computationally similar similarity matrices.
-Given that t-SNE generated a smattering of clusters for the minority populations when applied to the UKBB, we hypothesized that this was due to a convergence issue rather than a property of the global optimum of the t-SNE metric.
-
-To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data).
+To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data).
 The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).
-When applied to the UKbiobank, we find (perhaps subjectively) that UMAP provides a more compelling visualization of the data compared to both version of t-SNE.
-We also find that the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum.
-Qualitatively, the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity between clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
+When applied to the UKBB, UMAP provides a more compelling visualization of the data compared to both versions of t-SNE.
+The hybrid approach also produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum.
+The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but again began arbitrarily creating clusters and breaking the projection's continuous structure (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
 An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files.
-Given the many desirable properties of UMAP, such as computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.
+Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
 
@@
 Understanding population structure is important to identify confounders in medical genomics and studies of anthropology and human evolution.
-PCA of genomic data reflects genealogical and geographic data, but visualization in large datasets still requires scanning through a large number of pairwise plots.
+PCA reflects genealogical and geographic data, but visualization requires scanning through a large number of pairwise plots.
 UMAP condenses these components and comprehensively illustrates information --- phenotypic, geographic, and ancestral --- contained within genotypes on fine-scale levels and within the context of a global population structure.
@@
 
-Examinations of clustering in the three datasets provided many intriguing clusters that would otherwise have been difficult to identify.
+The three datasets provide many intriguing clusters that would otherwise have been difficult to identify.
 In particular, several areas from figure~\ref{fig:umap_ukbb_pc10}, highlighted in figure~\ref{fig:supp_ukbb_zoom}, show multiple unidentified groups related to each of the East Asian and South Asian super-populations, as well as to either or both of African or admixed populations.
@@
 
-Importantly, using UMAP is straightforward and fast.
+Using UMAP is straightforward and fast.
 Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results).
@@
 
-%added new stuff here on cautions and how to appropriately use UMAP
 Like most non-linear methods, UMAP lacks direct interpretability.
@@
 
-% what I'd like to highlight: UMAP finds local similarities in populations. even in the HRS
-% hispanic population we see a cluster appear
-% basically you can find groups that seem to be uniquely closely related within a cluster
-% you can't really say "how different" they are from others though
-UMAP works well at highlighting groups of individuals who share more ancestry within their group, as evidenced by the Hispanic individuals in the HRS and the many 1KGP populations.
+UMAP works well at highlighting groups of individuals who share more ancestry, as evidenced by the Hispanic individuals in the HRS and the many 1KGP populations.
 When clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters.
@@
 In figure~\ref{fig:supp_megamontage_pc10_50}, population clusters bounce around in each projection, again highlighting that the positions of disconnected clusters relative to one another are not meaningful.
-For this reason, UMAP coordinates should not be used, for example, as covariates in a GWAS or calculating distances between populations.
-
-% with the UKBB, the geography colouring is only possible with UMAP
-% tSNE would just break apart into tiny islands and scatter the people everywhere else.
+For this reason, UMAP coordinates should not be used as covariates in a GWAS or calculating distances between populations.
+
 UMAP excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
@@
 t-SNE can generate seemingly arbitrary clusters, as in figures \ref{fig:supp_tsne_umap_compare_ukbb} and \ref{fig:supp_ukbb_tsne}, where small islands of ethnic groups are scattered across the projection.
-These are undesirable when looking at colourings by phenotypes, geography, or admixture, where we are interested in spotting gradients across populations.
+These are undesirable when looking at colourings by phenotypes, geography, or admixture, where we are interested in gradients across populations.
 
@@
 
-There are downsides to using nonlinear approaches to visualize the data.
 Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for populations with larger sample sizes compared to PCA.
-This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult.
-Nonlinearity also complicates the interpretation of results.
-Distances in UMAP or tSNE space should not be used as a proxy for genetic distance.
+This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the variance within a population.
 We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artifact of the dimensional reduction strategy rather than a meaningful feature of the data.
@@
 
-\subsection*{Conclusion}
 With these caveats in mind, a priori data visualization plays a central role in quality control, hypothesis generation, and confounder identification for a wide range of genomic applications.
-Nonlinear approaches, despite their limitations, become increasingly useful as the size of datasets increases: We have shown that UMAP, in particular, reveals a wide range of features that would not be apparent using linear maps.
+Nonlinear approaches, despite their limitations, become increasingly useful as the size of datasets increases.
+UMAP, in particular, reveals a wide range of features that would not be apparent using linear maps.
 Given its ease of use, breadth of results, and low computational cost, we propose that UMAP should become a default companion to PCA in large genomic cohorts.
@@
     \begin{subfigure}{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{images/megamontage_eth_ukbb.pdf}
-    \end{subfigure}
-    \caption{UMAP on UKBB data, colored by self-identified ethnic background. Images are labelled by the number of components included.}
-    \label{fig:supp_montage_ukbb_eth}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{images/montage_ukbb_ns_pm.pdf}
-    \end{subfigure}
-    \caption{UMAP on UKBB data, colored by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates.
-Images are labelled by the number of components included.
-Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_montage_ukbb_ns}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_ukbb_ew_pm.pdf}
-    \end{subfigure}
-    \caption{UMAP on UKBB data, colored by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates.
-Images are labelled by the number of components included.
-Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_montage_ukbb_ew}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_combined_resized.pdf}
-    \end{subfigure}
-    \caption{Zoomed in areas of figure~\ref{fig:umap_ukbb_pc10}.
-Sections (i) and (ii) respectively focus on the African and Asian superpopulations, and section (iii) focuses on an area with individuals from many ethnic backgrounds.
-Noticeable clusters of unidentified ethnic backgrounds appear and are labelled "OEG" "(Other Ethnic Group)".}
-    \label{fig:supp_ukbb_zoom}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_cob_resized.jpeg}
-    \end{subfigure}
-    \caption{Using country of birth data, some of the larger unidentified groups from figure~\ref{fig:supp_ukbb_zoom} were identified as being born mostly in Japan, the Philippines, North Africa, the Middle East, and Central and South America.
-The large cluster of "Any other Asian Background" were mostly born in Sri Lanka.}
-    \label{fig:supp_ukbb_cob}
-\end{figure*}
-
-%%%%% UKBB phenotype plots
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417039_basophill_count_pct5_f.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (female). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_basophill_f}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417039_basophill_count_pct5_m.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (male). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_basophill_m}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417720_eosinophill_count_pct5_f.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_eosinophill_f}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417720_eosinophill_count_pct5_m.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_eosinophill_m}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416305_3063_0_0_pct1_f.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_fev_f}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416305_3063_0_0_pct1_m.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_fev_m}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by height (female). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_height_f}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_m.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by height (male). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_height_m}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_leukocyte_f}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_m.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_leukocyte_m}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714165614_neutrophill_count_pct5_f.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_neutrophill_f}
-\end{figure}
-
-\begin{figure}
-    \centering
-    %\begin{subfigure}{0.8\textwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714165614_neutrophill_count_pct5_m.pdf}
-    %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
-    \label{fig:supp_ukbb_neutrophill_m}
-\end{figure}
-
-%%%%% UKBB boxplots below
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_basophill_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_basophill_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_basophill_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_basophill_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_eosinophill_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_eosinophill_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_eosinophill_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_eosinophill_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_fev_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_fev_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_height_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_height_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_leukocyte_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_leukocyte_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_leukocyte_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_leukocyte_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_neutrophil_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_neutrophill_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_neutrophil_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_neutrophill_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_fev1_height_afr_permuted.pdf}
-    \end{subfigure}
-    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) colored by self-identified ethnic background (left, from figure~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right).
-An arrow points to an area where the FEV1 distribution appears to change, corresponding to where the clusters contain more people with self-identified mixed backgrounds.}
-    \label{fig:supp_comparison_fev_afr}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.pdf}
-    \end{subfigure}
-    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) colored by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
-The OEG cluster next to the Chinese cluster is colored differently, suggesting this population may have different FEV1 characteristics.
-A cluster of OEG/other white individuals is more blue, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
-    \label{fig:supp_comparison_fev_chi_eur}
+    \includegraphics[width=0.95\textwidth]{images/1KGP_UMAP_PCS100_PCE3450_NC2_NN15_MD05_201944185843.jpeg}
+    \end{subfigure}
+    \caption{UMAP applied the last 3350 principal components of the 1KGP.}
+    \label{fig:supp_1kgp_3350}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
+    \end{subfigure}
+    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, colored by estimated admixture proportions.}
+    \label{fig:supp_umap_hrs_hisp_admix}
 \end{figure*}
@@
 
+\begin{figure}
+\centering
+   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
+   \caption{UMAP on the first 10 principal components of HRS data.
+Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
+This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
+    \label{fig:umap_hrs_admix}
+\end{figure}
+
 \begin{figure*}
@@
     \caption{UMAP on the top 10 principal components of the HRS dataset, colored by Census Bureau birth region.
+Each colour represents one of the 10 birth regions.
 There is no obvious pattern in the clusters of majority "White Not Hispanic" individuals.}
     \label{fig:supp_hrs_born}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
+    \end{subfigure}
+    \caption{UMAP projection of the top 10 principal components of the combined HRS and 1KGP datasets.
+One cluster (in the box) does not group with any of the 1KGP populations.
+A cluster of Finnish (FIN) individuals consistently appears in the "White Not Hispanic" (WNH) group.
+Groups of Central and South American populations from the 1KGP (CLM, Colombian; MXL, Mexican; PEL, Peruvian; PUR, Puerto Rican) form nearby or within the HRS Hispanic cluster (HIS).
+Iberian individuals (IBS) cluster near the Hispanic population.
+Toscani individuals (TSI) form some small clusters and sometimes appear near the Iberian and Hispanic populations.
+Individuals with British/Scottish (GBR) or Northern/Western European ancestry (CEU) are scattered throughout the WNH clusters.
+Individuals with African ancestry from the 1KGP group with Black Americans from the HRS (AFR).
+Similar population groupings occur with South Asian (SAS) and East Asian (EAS) individuals.}
+    \label{fig:supp_hrs_1000g}
 \end{figure*}
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
-    \end{subfigure}
-    \caption{UMAP projection of the top 10 principal components of the combined HRS and 1KGP datasets.
-One cluster (in the box) does not group with any of the 1KGP populations.
-A cluster of Finnish (FIN) individuals consistently appears in the "White Not Hispanic" (WNH) group.
-Groups of Central and South American populations from the 1KGP (CLM, Colombian; MXL, Mexican; PEL, Peruvian; PUR, Puerto Rican) form nearby or within the HRS Hispanic cluster (HIS).
-Iberian individuals (IBS) cluster near the Hispanic population.
-Toscani individuals (TSI) form some small clusters and sometimes appear near the Iberian and Hispanic populations.
-Individuals with British/Scottish (GBR) or Northern/Western European ancestry (CEU) are scattered throughout the WNH clusters.
-Individuals with African ancestry from the 1KGP group with Black Americans from the HRS (AFR).
-Similar population groupings occur with South Asian (SAS) and East Asian (EAS) individuals.}
-    \label{fig:supp_hrs_1000g}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
     \includegraphics[width=\textwidth]{images/HRS_PCGRID_8.pdf}
@@
     \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
-    \end{subfigure}
-    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, colored by estimated admixture proportions.}
-    \label{fig:supp_umap_hrs_hisp_admix}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/admixture_plot_highlight_mountain_copy.pdf}
-    \end{subfigure}
-    \caption{Admixture plot of Hispanic individuals in the HRS. Those born in the Mountain census region fall between the white lines (indices 48 to 184)}
-    \label{fig:supp_hrs_hisp_admix}
+    \begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_cob_resized.jpeg}
+    \end{subfigure}
+    \caption{Using country of birth data, some of the larger unidentified groups from figure~\ref{fig:supp_ukbb_zoom} were identified as being born mostly in Japan, the Philippines, North Africa, the Middle East, and Central and South America.
+The large cluster of "Any other Asian Background" were mostly born in Sri Lanka.}
+    \label{fig:supp_ukbb_cob}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.95\textwidth}
+    \includegraphics[width=0.95\textwidth]{images/megamontage_eth_ukbb.pdf}
+    \end{subfigure}
+    \caption{UMAP on UKBB data, colored by self-identified ethnic background. Images are labelled by the number of components included.}
+    \label{fig:supp_montage_ukbb_eth}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.95\textwidth}
+    \includegraphics[width=0.95\textwidth]{images/montage_ukbb_ns_pm.pdf}
+    \end{subfigure}
+    \caption{UMAP on UKBB data, colored by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates.
+Images are labelled by the number of components included.
+Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_montage_ukbb_ns}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/montage_ukbb_ew_pm.pdf}
+    \end{subfigure}
+    \caption{UMAP on UKBB data, colored by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates.
+Images are labelled by the number of components included.
+Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_montage_ukbb_ew}
 \end{figure*}
@@
 
+%%%%% UKBB phenotype plots
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417039_basophill_count_pct5_f.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_basophill_f}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417039_basophill_count_pct5_m.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_basophill_m}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417720_eosinophill_count_pct5_f.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_eosinophill_f}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871417720_eosinophill_count_pct5_m.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_eosinophill_m}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416305_3063_0_0_pct1_f.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_fev_f}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416305_3063_0_0_pct1_m.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_fev_m}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by height (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_height_f}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_m.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by height (male). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_height_m}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_leukocyte_f}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_m.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_leukocyte_m}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714165614_neutrophill_count_pct5_f.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_neutrophill_f}
+\end{figure}
+
+\begin{figure}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714165614_neutrophill_count_pct5_m.pdf}
+    %\end{subfigure}
+    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_neutrophill_m}
+\end{figure}
+
+%%%%% UKBB phenotypes end
+
+\begin{figure*}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/ukbb_pcs_2019410184041_Height_res_pct1_f.jpeg}
+    %\end{subfigure}
+    \caption{Principal components 1 and 2 from the UKBB, coloured by age-adjusted residual height (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_pca_height_res_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/ukbb_pcs_201941019249_3063_0_0_pct1_f.jpeg}
+    %\end{subfigure}
+    \caption{Principal components 1 and 2 from the UKBB, coloured by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_pca_fev_f}
+\end{figure*}
+
+%\begin{figure*}
+%    \centering
+%    \includegraphics[width=0.4\columnwidth]{images/ukbb_pcs_201941019249_3063_0_0_pct1_f.jpeg}
+%    \caption{Test. Data has been randomized as explained in the materials and methods section.}
+%    \label{fig:supp_ukbb_tsne_fev_f}
+%\end{figure*}
+
+\clearpage
+\begin{figure*}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_TSNE_10PCs_DefaultPerplexity_2019410184913_Height_res_pct1_f.jpeg}
+    %\end{subfigure}
+    \caption{t-SNE on the first 10 principal components from the UKBB, coloured by age-adjusted residual height (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_tsne_height_res_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    %\begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=0.4\columnwidth]{images/UKBB_TSNE_10PCs_DefaultPerplexity_2019410191255_3063_0_0_pct1_f.jpeg}
+    %\end{subfigure}
+    \caption{t-SNE on the first 10 principal components from the UKBB, coloured by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_ukbb_tsne_fev_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.8\textwidth}
+    \includegraphics[width=\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_combined_resized.pdf}
+    \end{subfigure}
+    \caption{Zoomed in areas of figure~\ref{fig:umap_ukbb_pc10}.
+Sections (i) and (ii) respectively focus on the African and Asian superpopulations, and section (iii) focuses on an area with individuals from many ethnic backgrounds.
+Noticeable clusters of unidentified ethnic backgrounds appear and are labelled "OEG" "(Other Ethnic Group)".}
+    \label{fig:supp_ukbb_zoom}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/montage_fev1_height_afr_permuted.pdf}
+    \end{subfigure}
+    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) colored by self-identified ethnic background (left, from figure~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right).
+An arrow points to an area where the FEV1 distribution appears to change, corresponding to where the clusters contain more people with self-identified mixed backgrounds.}
+    \label{fig:supp_comparison_fev_afr}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.pdf}
+    \end{subfigure}
+    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) colored by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
+The OEG cluster next to the Chinese cluster is colored differently, suggesting this population may have different FEV1 characteristics.
+A cluster of OEG/other white individuals is more blue, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
+    \label{fig:supp_comparison_fev_chi_eur}
+\end{figure*}
+
 \begin{figure}[!htb]
@@
     \label{fig:supp_tsne_umap_compare_hrs}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_ukbb.jpeg}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
+Done on the UKBB dataset with 20000 iterations.
+The UMAP-initialized graph has been shifted by 230 iterations to approximate the 230 epochs UMAP uses for large datasets ($n>10,000$).}
+    \label{fig:supp_tsne_umap_compare_ukbb_graph}
 \end{figure}
@@
 
-\begin{figure}[!htb]
-    \centering
-    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_ukbb.jpeg}
-    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
-Done on the UKBB dataset with 20000 iterations.
-The UMAP-initialized graph has been shifted by 230 iterations to approximate the 230 epochs UMAP uses for large datasets ($n>10,000$).}
-    \label{fig:supp_tsne_umap_compare_ukbb_graph}
-\end{figure}
+%%%%% UKBB boxplots below
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_basophill_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_basophill_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_basophill_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_basophill_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_eosinophill_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_eosinophill_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_eosinophill_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_eosinophill_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_fev_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_fev_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_height_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_height_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_leukocyte_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_leukocyte_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_leukocyte_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_leukocyte_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_neutrophil_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_neutrophill_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_neutrophil_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values.
+Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_neutrophill_m}
+\end{figure*}
+
+%%%%% boxplots end
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/admixture_plot_highlight_mountain_copy.pdf}
+    \end{subfigure}
+    \caption{Admixture plot of Hispanic individuals in the HRS. Those born in the Mountain census region fall between the white lines (indices 48 to 184)}
+    \label{fig:supp_hrs_hisp_admix}
+\end{figure*}
 

### 2019-04-10 ms/main.tex
reorient some images

@@
     \begin{subfigure}{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{images/megamontage_eth_ukbb.pdf}
+    \includegraphics[width=0.95\textwidth]{images/default_clean_size10_alpha60_flip.jpeg}
     \end{subfigure}
@@
     \begin{subfigure}{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{images/montage_ukbb_ns_pm.pdf}
+    \includegraphics[width=0.95\textwidth]{images/default_clean_size10_alpha60_ns.jpeg}
     \end{subfigure}
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_ukbb_ew_pm.pdf}
+    \includegraphics[width=\textwidth]{images/default_clean_size10_alpha60_ew.jpeg}
     \end{subfigure}

### 2019-04-11 ms/main.tex
tidying up some writing

@@
     TSI, Tuscani in Italy.
-Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS colored by estimated admixture proportions.
+Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.
 (b) UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
@@
 Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters.
-Using either on the top principal components leads to more distinct population clusters and less defined continental structure.
+Using either on the top principal components leads to distinct population clusters and less defined continental structure.
 Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
@@
 
-Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), we find several population clusters that reflect shared ancestries.
+Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries.
 British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry.
@@
 
-\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form iclusters.
+\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters.
 Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas.
-For example, the HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}.
+The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}.
 We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g.
@@
 Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
-We have scaled these three proportions to values between 0 and 255, to color individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
 Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.
@@
 
-\subsubsection*{Deep population structure in the American Hispanic subpopulation} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
+\subsubsection*{Regional patterns in the Hispanic subpopulation} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
 One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
@@
 \end{subfigure}%
-\caption{The UKBB projected onto two dimensions, colored by self-reported ethnic background.
+\caption{The UKBB coloured by self-reported ethnic background.
 (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations.
@@
 Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain.
-The left image colors individuals by their north-south ("northing") coordinates, and the right image colors them by their east-west ("easting") coordinates.
+The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates.
 Adding more components creates finer clusters.
@@
     \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
-    \caption{A map of Great Britain colored by a 3D UMAP projection.
-Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in color.
+    \caption{A map of Great Britain coloured by a 3D UMAP projection.
+Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in colour.
 Patterns in genetic similarity are visible in Scotland, South England, the East and West Midlands, and major urban centres.
-A flattened 2D view of of the 3D projection used for coloring is presented in the top right.
+A flattened 2D view of of the 3D projection used for colouring is presented in the top right.
 To protect participant privacy, data has been randomized as explained in the materials and methods.}
@@
 \subsection*{Discussion}
-Understanding population structure is important to identify confounders in medical genomics and studies of anthropology and human evolution.
-PCA reflects genealogical and geographic data, but visualization requires scanning through a large number of pairwise plots.
-UMAP condenses these components and comprehensively illustrates information --- phenotypic, geographic, and ancestral --- contained within genotypes on fine-scale levels and within the context of a global population structure.
-In large datasets where the number of significant PCs is large, the resulting representation has important advantages over PCA alone and provides a superior visualization to t-SNE.
-
-The three datasets provide many intriguing clusters that would otherwise have been difficult to identify.
-In particular, several areas from figure~\ref{fig:umap_ukbb_pc10}, highlighted in figure~\ref{fig:supp_ukbb_zoom}, show multiple unidentified groups related to each of the East Asian and South Asian super-populations, as well as to either or both of African or admixed populations.
-Additionally, the Hispanic population of the HRS contains a geographically-restricted cluster that could not have been identified from pairwise examinations of principal components.
-The 1KGP --- frequently used in medical and population studies --- contained splits in the Gujarati and Punjabi population samples that were not visible PCA or Admixture analysis alone (although a split among Gujarati is arguably visible in the Admixture analysis with K=12 in \cite{10002015global}).
-
-Application to the UKBB underscores the strength of UMAP in large cohorts.
-We see clear, fine relationships between genotype and phenotype and geography, and this is presented in a visualization that accounts for natural genetic clustering.
-Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with phenotypes such as height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.
-
-Using UMAP is straightforward and fast.
-Most of the plots presented in this article were generated directly from the PCA data using UMAP with default parameters, except that we set the "minimum distance" parameter to 0.5 which made fine features on UMAP more visible (results with default parameter 0.1 provided qualitatively similar results).
-Given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
+UMAP comprehensively illustrates information --- phenotypic, geographic, and ancestral --- contained within genotypes on fine-scale levels and within the context of a global population structure.
+Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
+It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
+By patching together many local genetic neighbourhoods, the visualizations have continuous flow and a semblance of a global structure, unlike arbitrary clusters generated by t-SNE as in figures \ref{fig:supp_tsne_umap_compare_ukbb} and \ref{fig:supp_ukbb_tsne}.
+These are undesirable when we are interested in gradients across populations.
+
+Using UMAP reveals intriguing clusters that would otherwise have been very difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS.
+The 1KGP --- frequently used in medical and population studies --- contained splits in the Gujarati and Punjabi population samples that were not visible through PCA or Admixture analysis alone (although a split among Gujarati is arguably visible in the Admixture analysis with K=12 in \cite{10002015global}).
+Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering.
+Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.
+
+UMAP works well at highlighting groups of individuals who share more ancestry; when clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters.
+The 1KGP dataset was deliberately crafted to come from geographically separated populations that are homogeneous internally.
+Thus, most of its populations form isolated clusters, with the exception of those that most closely share ancestries.
 
 Like most non-linear methods, UMAP lacks direct interpretability.
-The dimensions in its embeddings have no inherent meaning, unlike PCA where the axes correspond to the proportion of variance explained.
-Approaches like PCA will keep more distantly related populations further from each other, but in choosing which dimensions to examine, we are forced to flatten dimensions and clump together groups such as Central and South Americans with South Asians as in figure~\ref{fig:1000g_compare}.
-UMAP is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are very distant from each other cannot be considered to be more different from each other.
-Since the low dimensional embedding is formed by patching together many neighbourhoods there is a semblance of a global structure -- one that is superior to methods like t-SNE; however, the developers of UMAP caution that researchers interested primarily in global structure may wish to consider other methods.
-
-UMAP works well at highlighting groups of individuals who share more ancestry, as evidenced by the Hispanic individuals in the HRS and the many 1KGP populations.
-When clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters.
-The 1KGP dataset was deliberately crafted to come from geographically separated populations that are relatively homogeneous internally.
-For this reason most of its populations form isolated clusters, with the exception of those that most closely share ancestries (Iberians and Toscani, Han Chinese and Southern Han Chinese).
-In figure~\ref{fig:supp_megamontage_pc10_50}, population clusters bounce around in each projection, again highlighting that the positions of disconnected clusters relative to one another are not meaningful.
-For this reason, UMAP coordinates should not be used as covariates in a GWAS or calculating distances between populations.
-
-UMAP excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
-Unlike t-SNE, it patches together many local neighbourhoods, resulting in a relatively continuous flow to the data.
-t-SNE can generate seemingly arbitrary clusters, as in figures \ref{fig:supp_tsne_umap_compare_ukbb} and \ref{fig:supp_ukbb_tsne}, where small islands of ethnic groups are scattered across the projection.
-These are undesirable when looking at colourings by phenotypes, geography, or admixture, where we are interested in gradients across populations.
-
-% Note: the HRS data didn't really "converge" as nicely as 1KGP data with its subcontinental structure. PCs may serve a dual role to filter data?
-We suggest making several runs of UMAP and to highlight consistent qualitative results.
-Tuning parameters will generally not make a difference beyond making some patterns easier to see.
+It is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are distant from each other cannot be considered to be more different from each other.
+Figure~\ref{fig:supp_megamontage_pc10_50} demonstrates that disconnected population clusters can bounce around in each projection.
+For this reason, UMAP coordinates should not be used as GWAS covariates or for calculating distances between populations.
 The number of principal components used will have an effect on the visualization depending on the variable of interest.
-In each dataset, the groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
-Geographical coordinates in the UKBB continued to concentrate even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
-The total number of components to use is limited primarily by computational power.
-
-Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for populations with larger sample sizes compared to PCA.
-This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the variance within a population.
-We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artifact of the dimensional reduction strategy rather than a meaningful feature of the data.
+Across datasets, groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
+Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
+The main limitation on how many principal components to use is computational power.
+
+Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for larger populations than PCA.
+This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB.
+We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.
 Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.
@@
 UMAP developers described "sensible" values for nearest neighbours as between 5 and 50 and minimum distance between 0.5 and 0.001.
+Tuning these parameters will not change qualitative results much but may make patterns easier to identify.
+Increasing the number of neighbours will increase the computational load.
 
@@
 
-For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the color value from a randomly selected neighbour.}
+For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}
 
@@
     \end{subfigure}
-    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, colored by estimated admixture proportions.}
+    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions.}
     \label{fig:supp_umap_hrs_hisp_admix}
@@
     \caption{UMAP applied to the first 10 principal components of HRS data.
-Points colored by self-identified race, Hispanic status, and Mexican-American status.
+Points coloured by self-identified race, Hispanic status, and Mexican-American status.
 The cluster on the left is mostly people who identify as neither Black nor White and were born outside the contiguous United States or in the Pacific census region.
@@
    \caption{UMAP on the first 10 principal components of HRS data.
-Coloring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
-This projection colored by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
+colouring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
+This projection coloured by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
     \label{fig:umap_hrs_admix}
@@
     \end{subfigure}
-    \caption{UMAP on the top 10 principal components of the HRS dataset, colored by Census Bureau birth region.
+    \caption{UMAP on the top 10 principal components of the HRS dataset, coloured by Census Bureau birth region.
 Each colour represents one of the 10 birth regions.
@@
     \end{subfigure}
-    \caption{Pairwise plots of the first 8 principal components of the Hispanic subset of the HRS. Those born in the Mountain region are colored green.}
+    \caption{Pairwise plots of the first 8 principal components of the Hispanic subset of the HRS. Those born in the Mountain region are coloured green.}
     \label{fig:supp_hrs_hisp_grid}
@@
     \end{subfigure}
-    \caption{UMAP on UKBB data, colored by self-identified ethnic background. Images are labelled by the number of components included.}
+    \caption{UMAP on UKBB data, coloured by self-identified ethnic background. Images are labelled by the number of components included.}
     \label{fig:supp_montage_ukbb_eth}
@@
     \end{subfigure}
-    \caption{UMAP on UKBB data, colored by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates.
+    \caption{UMAP on UKBB data, coloured by northing values, with more blue representing more northern coordinates and more red representing more southern coordinates.
 Images are labelled by the number of components included.
@@
     \end{subfigure}
-    \caption{UMAP on UKBB data, colored by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates.
+    \caption{UMAP on UKBB data, coloured by easting values, with more yellow representing more eastern coordinates and more pink representing more western coordinates.
 Images are labelled by the number of components included.
@@
     \includegraphics[width=0.95\columnwidth]{images/UKBB_TSNE_10PCs_DefaultPerplexity_eth.pdf}
-    \caption{t-SNE applied to the top 10 principal components of the UKBB, colored by ethnic background.
+    \caption{t-SNE applied to the top 10 principal components of the UKBB, coloured by ethnic background.
 The unbalanced populations resulted in many individuals and populations being orphaned along the periphery of the main cluster.}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by basophil count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_basophill_f}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by basophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by basophil count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_basophill_m}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by eosinophil count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_eosinophill_f}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by eosinophil count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_eosinophill_m}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by FEV1 (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_fev_f}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by FEV1 (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_fev_m}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by height (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by height (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_height_f}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by height (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by height (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_height_m}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by leukocyte count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_leukocyte_f}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by leukocyte count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_leukocyte_m}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by neutrophil count (female). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_neutrophill_f}
@@
     %\end{subfigure}
-    \caption{UMAP on the top 10 principal components of the UKBB colored by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
+    \caption{UMAP on the top 10 principal components of the UKBB coloured by neutrophil count (male). Data has been randomized as explained in the materials and methods section.}
     \label{fig:supp_ukbb_neutrophill_m}
@@
     \end{subfigure}
-    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) colored by self-identified ethnic background (left, from figure~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right).
+    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) coloured by self-identified ethnic background (left, from figure~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right).
 An arrow points to an area where the FEV1 distribution appears to change, corresponding to where the clusters contain more people with self-identified mixed backgrounds.}
@@
     \end{subfigure}
-    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) colored by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
-The OEG cluster next to the Chinese cluster is colored differently, suggesting this population may have different FEV1 characteristics.
+    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
+The OEG cluster next to the Chinese cluster is coloured differently, suggesting this population may have different FEV1 characteristics.
 A cluster of OEG/other white individuals is more blue, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}

### 2019-04-11 ms/main.tex
move some stuff to matmethods, small fixes

@@
 Here we apply UMAP to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data.
-We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry population-scale datasets.
+We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
 \end{abstract}
@@
 With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds.
-The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.
+The end result is a patchwork of low-dimensional representations of neighbourhoods that groups genetically similar individuals together on a local scale while better preserving long-range topological connections to more distantly related individuals.
 The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.
@@
 \subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters.
-Most medical cohorts comprise larger numbers of individuals sampled across extended geographical areas.
-The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}.
-We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g.
-White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau.
-Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
-We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas.
+The HRS contains genotype data of 12,454 Americans from a variety of backgrounds.
 Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.
@@
 
-\subsubsection*{Regional patterns in the Hispanic subpopulation} In contrast to the WNH individuals, applying UMAP to self-identified Hispanic individuals reveals clear groupings related to birth region.
-One separate cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
+\subsubsection*{Regional patterns in the Hispanic subpopulation} Applying UMAP to self-identified Hispanic individuals in the HRS reveals clear groupings related to birth region.
+One cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
 This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6.
@@
 
-\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} The UKBB provides genotype data on 488,377 individuals along with self-identified ethnic background in a hierarchical tree-structured dictionary.
-Participants provided ethnic background on two occasions.
-We used the initial ethnicity after finding minimal differences between the two.
-The dataset is majority White (88.3\% British, 2.6\% Irish, 3.4\% other), with large populations identifying as Black (1.6\% either African, Caribbean, or other), Asian (1.9\% either Indian, Pakistani, Bangladeshi, or other), Chinese (0.3\%), an other ethnic group (0.8\%), mixed ethnicity (0.6\%), or an unavailable response (0.5\%).
-
-UMAP on the top 10 principal components reveals both continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): The patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes.
+\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
+The UKBB is a trove of data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds.
+UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes.
 The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of large and multi-ethnic datasets.
@@
 The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group".
-The patchwork of genetic neighbourhoods is connected by trails of admixed individuals.
-These trails come together in a nexus of individuals with a variety of ethnicities.
-Many claim mixed ancestry, and there are clear groups of individuals who belong to an "other ethnic group"; some of these were identified using country of birth in figure~\ref{fig:supp_ukbb_cob} and appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
+The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities.
+Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group".
+Using data on countries of birth, we identified many groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
 
@@
 
-Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for larger populations than PCA.
+Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for larger sample sizes than PCA.
 This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB.
@@
 We used genotype data from 3,450 individuals from the 1KGP project using Affy 6.0 genotyping\cite{10002015global}.
+
+The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}.
+We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g.
+White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau.
+Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+
+The UKBB provides genotype data on 488,377 individuals along with self-identified ethnic background in a hierarchical tree-structured dictionary.
+Participants provided ethnic background on two occasions.
+We used the initial ethnicity after finding minimal differences between the two.
+The dataset is majority White (88.3\% British, 2.6\% Irish, 3.4\% other), with large populations identifying as Black (1.6\% either African, Caribbean, or other), Asian (1.9\% either Indian, Pakistani, Bangladeshi, or other), Chinese (0.3\%), an other ethnic group (0.8\%), mixed ethnicity (0.6\%), or an unavailable response (0.5\%).
 

### 2019-04-11 ms/cover_letter.tex
add cover letter

--- 2018-12-06 ms/cover_letter.tex
@@
 \\\\
-We are pleased to submit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics.
-Using a newly developed dimensional reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in massive and heterogeneous genomic datasets.
-In contrast to existing methods, this approach is able to represent both discrete and continuous population structure, as well as admixture.
+We are pleased to submit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in Nature Genetics.
+Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are expected to proliferate.
+In contrast to existing methods, this approach is able to represent both discrete and continuous population structure and the distributions of ancestries, admixture, and phenotypes within the context of such structure.
+We expect this approach will become standard in genetics --- since the release of our preprint, it has been used to study ancestral diversity in the Genome Aggregation Database (gnomAD)\footnote{Karczewski, Konrad J., et al.
+"Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes." BioRxiv (2019): 531210.} and the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
+"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and integrated as a preprocessing step in machine learning approaches\footnote{Tonkin-Hill, Gerry, et al.
+"Fast Hierarchical Bayesian Analysis of Population Structure." BioRxiv (2018): 454355.}.
 \\\\
 By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts.
-For example, we identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region.
-These individuals are likely descendants of the Hispano population.
-We also identify multiple associations between genetics, geography, and phenotypes in the UK biobank and population structure within 1000 Genomes Project populations.
+We identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region --- likely descendants of the Hispano population --- and  multiple associations between genetics, geography, and phenotypes in the UK biobank as well as population structure within 1000 Genomes Project populations, which led to a follow-up study by 23andMe\footnote{23andMe.
+"23andMe Tests New Ancestry Breakdown in Central and South Asia".
+https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/ (2019-04-03)}.
+%\\\\
+%In other -omics fields, such as single-cell expression data, nonlinear nonparametric approaches such as t-SNE have become standard tools for data exploration.
+Principal component analysis has remained the standard for dimension reduction in human genetics, partly because nonlinear alternatives such as t-SNE don't behave well when applied to large-scale genetic data.
+We show that UMAP outperforms t-SNE computationally, qualitatively and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
 \\\\
-This work has broad applicability.
-Our approach has already been implemented as a visualization tool by leading cohorts such as GnomAD\footnote{Fancioli, Laurent.
-``gnomAD v2.1''.
-https://macarthurlab.org/2018/10/17/gnomad-v2-1/ (2018-10-17).}, and used as a preprocessing step for machine learning approaches\footnote{Tonkin-Hill, Gerry, et al.
-``Fast Hierarchical Bayesian Analysis of Population Structure.'' bioRxiv (2018): 454355.}.
-Our preprint is also in the 99th percentile of all-time bioRxiv preprints for the Altmetric impact statistic.
-Because of its extreme computational tractability and its ability to efficiently summarize genomic information, we expect that our approach will become a new standard tool in data analysis for large genomic studies.
-\\\\
-The manuscript presents new methodology and applications that reveal fine-scale genomic and phenotypic variation in large-scale cohorts.
-We believe that this will be of interest to a broad variety of disciplines including population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
+We show that UMAP outperforms t-SNE --- the current state-of-the-art --- computationally, qualitatively, and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
+This will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for Nature Genetics.
 \\

### 2019-04-18 ms/main.tex
add some more maps. add citations, discussions

@@
 This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}).
-Analysis of the cluster by 23andMe suggested the Gujarati grouping is the result of endogamy~\cite{23andme}.
+Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}, and analysis of the cluster by 23andMe suggests it is the result of endogamy\cite{23andme}.
 
@@
 The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
-    \caption{A map of Great Britain coloured by a 3D UMAP projection.
+Using data about country of birth, we also generalized this approach to colour a world map in figure~\ref{fig:umap_world_map}.
+
+%\begin{figure}
+%    \centering
+%    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
+%    \caption{A map of Great Britain coloured by a 3D UMAP projection.
 Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in colour.
@@
 To protect participant privacy, data has been randomized as explained in the materials and methods.}
+%    \label{fig:umap_ukbb_geo_map}
+%\end{figure}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}{\columnwidth}
+    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
+    \caption{Each point is an individual placed based on where they live.
+Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
     \label{fig:umap_ukbb_geo_map}
+    \end{subfigure}
+     \begin{subfigure}{\columnwidth}
+         \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
+   \caption{Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
+Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
+Vestiges of colonialism and related migrations are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States.
+Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
+    \label{fig:umap_world_map}
+      \end{subfigure}
+      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.
+Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates.
+Individuals who are closer to each other in the projection will be closer in colour in the maps.
+More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.}
 \end{figure}
@@
 Tuning these parameters will not change qualitative results much but may make patterns easier to identify.
-Increasing the number of neighbours will increase the computational load.
+Increasing the number of neighbours will increase the computational load, and a smaller minimum distance can break the connectivity between clusters, though the same individuals will continue to group together.
 
 UMAP and t-SNE projections were carried out on an iMac with a 3.5GhZ Intel Core i7 processor, 32 GB 1600 MHz DDR3 of RAM, and an NVIDIA GeForce GTX 775M 2048 MB graphics card.
+
+Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate.
+For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country.
+Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States.
+This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations.
+For this reason, differences between Ireland/the UK and other countries appear exaggerated.
+Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing.
+We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}.
 
@@
 Additionally, spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km.
-
 For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}
@@
 \subsubsection*{Acknowledgements}
-We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the dataset.
-We also thank Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access, and Selin Jessa for introducing us to UMAP.
+We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the datasets.
+We also thank Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access; David Poznik, Liz Babalola, and Adam Auton from 23andMe for discussing findings in the 1KGP; and Selin Jessa for introducing us to UMAP.
 This research has been conducted using the UK Biobank Resource under Application Number 6728.
@@
 
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/meanAsiaMap.jpg}
+    \end{subfigure}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia.}
+    \label{fig:supp_umap_ukbb_asia}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/meanCarribeanMap.jpg}
+    \end{subfigure}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean.}
+    \label{fig:supp_umap_ukbb_car}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/meanEuroMap.jpg}
+    \end{subfigure}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe.}
+    \label{fig:supp_umap_ukbb_eur}
+\end{figure*}
+
 \begin{figure}[!htb]
@@
 %%%%% boxplots end
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
+    \end{subfigure}
+    \caption{Alternate colouring of figre~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
+    \label{fig:supp_umap_ukbb_worldmap_alt}
+\end{figure*}
 

### 2019-04-23 ms/main.tex
add citation for hispano popn

@@
 Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding.
-The cluster possibly comprises the Hispano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations.
-A recent preprint discusses the Mountain Region Hispanics and provides a more detailed historical description \cite{Jordan333609}.
+The cluster possibly comprises the Hispano/Nuevomexicano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations.
+Such a grouping has been previously identified in AncestryDNA data using sequential clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.
 

### 2019-04-23 ms/main.tex
abstract

@@
 %The abstract is typically 100-150 words, unreferenced. 
-Improvements in sequencing have generated massive and diverse genomic datasets of populations.
-Dimension reduction is necessary to condense and analyze the data.
-Uniform manifold approximation and projection (UMAP) is a non-linear dimension reduction tool recently made available.
-Here we apply UMAP to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data.
+Human populations feature both discrete and continuous patterns of variation.
+Current analysis approaches struggle to jointly identify these patterns because of modelling assumptions, mathematical constraints, or numerical challenges.
+Here we apply uniform manifold approximation and projection (UMAP), a non-linear dimension reduction tool, to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data.
 We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.

### 2019-05-02 ms/main.tex
edits and comments throughout

@@
 Here we apply uniform manifold approximation and projection (UMAP), a non-linear dimension reduction tool, to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data.
-We demonstrate this approach outperforms other standard methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
+This approach outperforms other visualization methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
 \end{abstract}
@@
 
-A common practice in dimensional reduction is to first apply PCA.
+Nonlinear dimension reduction methods tend to be computationally intensive.
+A common practice to reduce this burden is to first apply PCA to data, and perform dimensional reduction on data projected to leading principal components (PCs).
 In addition to being computationally advantageous, this discards noise that can confound nonlinear approaches: population structure arising from $n$ isolated randomly-mating demes can be described by the leading $n-1$ PCs, with the following PCs describing stochastic variation in relatedness~\cite{eigen2006}.
@@
 Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters.
-Using either on the top principal components leads to distinct population clusters and less defined continental structure.
+Using either method on the top principal components leads to distinct population clusters and less defined continental structure.
 Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
@@
 This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}).
-Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}, and analysis of the cluster by 23andMe suggests it is the result of endogamy\cite{23andme}.
+Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}.
+Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym.
+[D.
+Poznik, 23andme, personal communication, and \cite{23andme}].
 
@@
 The cluster possibly comprises the Hispano/Nuevomexicano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations.
-Such a grouping has been previously identified in AncestryDNA data using sequential clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.
+Such a cluster has been previously identified in AncestryDNA data using sequential clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.
 
 \subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
-The UKBB is a trove of data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds.
+The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds.
 UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes.
-The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of large and multi-ethnic datasets.
+The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.
 
@@
 The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.
-Using data about country of birth, we also generalized this approach to colour a world map in figure~\ref{fig:umap_world_map}.
+Using data about country of birth, we performed a similar analysis of a world map~\ref{fig:umap_world_map}, revealing subtle regional variation around the world.
 
@@
 Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
-Vestiges of colonialism and related migrations are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
-Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States.
+Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history.
+Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States \sgcomment{rationale?}.
 Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
@@
 This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds.
-Boxplots and statistical testing suggest that relative to White British populations FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
+Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
 Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; using available country of birth data, most of the unidentified ethnic group were born in Japan.
@@
 When applied to the UKBB, UMAP provides a more compelling visualization of the data compared to both versions of t-SNE.
-The hybrid approach also produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum.
-The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but again began arbitrarily creating clusters and breaking the projection's continuous structure (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
+Interestingly, the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), suggesting that UMAP can help standard t-SNE implementation to reach the t-SNE objective.
+The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but lost much fine-scale resolution and the relationship among clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
 An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files.
@@
 \subsection*{Discussion}
-UMAP comprehensively illustrates information --- phenotypic, geographic, and ancestral --- contained within genotypes on fine-scale levels and within the context of a global population structure.
+UMAP comprehensively illustrates genotypic information at fine-scales and within the context of a global population structure.
 Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
 It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
-By patching together many local genetic neighbourhoods, the visualizations have continuous flow and a semblance of a global structure, unlike arbitrary clusters generated by t-SNE as in figures \ref{fig:supp_tsne_umap_compare_ukbb} and \ref{fig:supp_ukbb_tsne}.
-These are undesirable when we are interested in gradients across populations.
-
-Using UMAP reveals intriguing clusters that would otherwise have been very difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS.
-The 1KGP --- frequently used in medical and population studies --- contained splits in the Gujarati and Punjabi population samples that were not visible through PCA or Admixture analysis alone (although a split among Gujarati is arguably visible in the Admixture analysis with K=12 in \cite{10002015global}).
+By patching together many local genetic neighbourhoods, the visualizations preserve aspects of both local and global structure, unlike any existing method that we are familiar with.
+
+Using UMAP reveals intriguing clusters that would otherwise have been very difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}.
 Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering.
@@
 Thus, most of its populations form isolated clusters, with the exception of those that most closely share ancestries.
+\sgcomment{I don't understand what this paragraph is trying to say.
+You might be a bit too wordy here.
+}
 
 Like most non-linear methods, UMAP lacks direct interpretability.
-It is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are distant from each other cannot be considered to be more different from each other.
-Figure~\ref{fig:supp_megamontage_pc10_50} demonstrates that disconnected population clusters can bounce around in each projection.
-For this reason, UMAP coordinates should not be used as GWAS covariates or for calculating distances between populations.
-The number of principal components used will have an effect on the visualization depending on the variable of interest.
-Across datasets, groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
+It is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are distant from each other are not necessarily very different from each other.
+Figure~\ref{fig:supp_megamontage_pc10_50} demonstrates that disconnected population clusters can appear at different locations depending on parameter choices.
+For this reason, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
+
+We found that pre-processing the data with PCA allowed for considerable time savings.
+Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
 Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
-The main limitation on how many principal components to use is computational power.
-
-Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for larger sample sizes than PCA.
+
+Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for populations with larger sample sizes.
 This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB.

### 2019-05-02 ms/main.tex
shortened UMAP-tsne comparison

@@
 \subsubsection*{Comparing t-SNE and UMAP}
-To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data).
+
+Identifying the best dimension reduction technique is to an extent subjective.
+However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
+Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}.
+Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
+
+%To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data).
 The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).

### 2019-05-03 ms/main.tex
added titles to captions

@@
       \end{subfigure}
-      \caption{Using different colourings with the projection in figure~\ref{fig:umap_ukbb_pc10}, relationships between population structure, geography, and phenotypes become clear.
+      \caption{Relationships between population structure, geography, and phenotype heterogeneity.
 (Top) Each individual is coloured by their geographical coordinates of residence.
@@
          \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
-   \caption{Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
+   \caption{Geographic distribution of UMAP coordinates.
+Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
 Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
@@
 Individuals who are closer to each other in the projection will be closer in colour in the maps.
-More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.}
+More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.
+\sgcomment{Merge captions.
+Make sure to have overall title!
+}}
 \end{figure}
@@
 We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}.
+\sgcomment{Is this really a "material and method" comment?
+THe logical flow with the last comment is not clear to me.}
 
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia.}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
     \label{fig:supp_umap_ukbb_asia}
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean.}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
     \label{fig:supp_umap_ukbb_car}
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe.}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
     \label{fig:supp_umap_ukbb_eur}
@@
     \end{subfigure}
-    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_basophill_f}
@@
     \end{subfigure}
-    \caption{Boxplot of basophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Basophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_basophill_m}
@@
     \end{subfigure}
-    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Eeosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_eosinophill_f}
@@
     \end{subfigure}
-    \caption{Boxplot of eosinophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Eosinophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_eosinophill_m}
@@
     \end{subfigure}
-    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_fev_f}
@@
     \end{subfigure}
-    \caption{Boxplot of FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_fev_m}
@@
     \end{subfigure}
-    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_height_f}
@@
     \end{subfigure}
-    \caption{Boxplot of height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_height_m}
@@
     \end{subfigure}
-    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_leukocyte_f}
@@
     \end{subfigure}
-    \caption{Boxplot of leukocyte counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Leukocyte counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_leukocyte_m}
@@
     \end{subfigure}
-    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_neutrophill_f}
@@
     \end{subfigure}
-    \caption{Boxplot of neutrophil counts by sex and ethnic group, annotated with p-values.
-Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \caption{Neutrophil counts by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
     \label{fig:supp_box_neutrophill_m}
@@
     \end{subfigure}
-    \caption{Alternate colouring of figre~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
+    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
     \label{fig:supp_umap_ukbb_worldmap_alt}
@@
     \end{subfigure}
-    \caption{Admixture plot of Hispanic individuals in the HRS. Those born in the Mountain census region fall between the white lines (indices 48 to 184)}
+    \caption{Admixture plot of Hispanic individuals in the HRS. Individuals born in the Mountain census region fall between the white lines (indices 48 to 184)}
     \label{fig:supp_hrs_hisp_admix}

### 2019-05-03 ms/cover_letter.tex
reordered a bit

@@
 We are pleased to submit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in Nature Genetics.
-Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are expected to proliferate.
+Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are becoming the new frontier of genomics research.
 In contrast to existing methods, this approach is able to represent both discrete and continuous population structure and the distributions of ancestries, admixture, and phenotypes within the context of such structure.
-We expect this approach will become standard in genetics --- since the release of our preprint, it has been used to study ancestral diversity in the Genome Aggregation Database (gnomAD)\footnote{Karczewski, Konrad J., et al.
-"Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes." BioRxiv (2019): 531210.} and the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
-"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and integrated as a preprocessing step in machine learning approaches\footnote{Tonkin-Hill, Gerry, et al.
-"Fast Hierarchical Bayesian Analysis of Population Structure." BioRxiv (2018): 454355.}.
 \\\\
 By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts.
-We identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region --- likely descendants of the Hispano population --- and  multiple associations between genetics, geography, and phenotypes in the UK biobank as well as population structure within 1000 Genomes Project populations, which led to a follow-up study by 23andMe\footnote{23andMe.
-"23andMe Tests New Ancestry Breakdown in Central and South Asia".
-https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/ (2019-04-03)}.
+We identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region --- likely descendants of the Hispano population --- and  multiple associations between genetics, geography, and phenotypes in the UK biobank as well as population structure within 1000 Genomes Project populations.
 %\\\\
@@
 We show that UMAP outperforms t-SNE --- the current state-of-the-art --- computationally, qualitatively, and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
-This will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for Nature Genetics.
+Methods described in our preprint have already been used to study ancestral diversity in the Genome Aggregation Database (gnomAD)\footnote{Karczewski, Konrad J., et al.
+"Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes." BioRxiv (2019): 531210.} and the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
+"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and by 23andme.
+Indeed, 23andme was able to follow up on one of the population structure signals we identified in our preprint by recontacting their participants, and the resulting interpretation of the population structure in the Gujarati is now part of the current manuscript.
+
+We believe that this is compelling evidence that UMAP will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for Nature Genetics.
 \\

### 2019-05-27 ms/main.tex
clarify some stuff and better image captions

@@
 \newcommand{\adpcomment}[1]{{\textcolor{orange}{ADP: #1}}}
+
+\newcommand{\rulesep}{\unskip\ \vrule\ }
+
 %\usepackage{subfig}
@@
     \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
-    \caption{Four methods of dimension reduction of 1KGP genotype data with population labels (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry.
+    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry.
 Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians.
@@
     \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
-    \caption{Subset of Hispanic population of the HRS}
+    \caption{}
     \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
@@
     \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
-    \caption{Subset of the Asian population of the UKBB}
+    \caption{}
         \label{fig:umap_ukbb_indian}
     \end{subfigure}
-    \caption{Applying UMAP to subets of data can reveal deep population structure.
-(a) UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
+    \caption{\textbf{Applying UMAP to subets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
 Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
@@
 Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.
-(b) UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
+\textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
 Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
@@
 \includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
-\caption{Principal components 1 and 2}%
+\caption{}%
 \label{fig:pc_ukbb}%
@@
 \includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
-\caption{UMAP on first 10 principal components}%
+\caption{}%
 \label{fig:umap_ukbb_pc10}%
 \end{subfigure}%
-\caption{The UKBB coloured by self-reported ethnic background.
-(a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations.
-(b) UMAP on the first 10 principal components.
+\caption{\textbf{The UKBB coloured by self-reported ethnic background.} \textbf{(a)} The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations.
+\textbf{(b)} UMAP on the first 10 principal components.
 The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population.
@@
     \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
-    \caption{Colouring the projection by geographic coordinates.}
+    \caption{}
     \label{fig:umap_ukbb_geo_osgb}
@@
     \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
-   \caption{Colouring the projection by phenotypic data}
+   \caption{}
     \label{fig:umap_height_female}
       \end{subfigure}
-      \caption{Relationships between population structure, geography, and phenotype heterogeneity.
-(Top) Each individual is coloured by their geographical coordinates of residence.
+      \caption{\textbf{UMAP captures relationships between population structure, geography, and phenotype heterogeneity.} \textbf{(a)} Each individual is coloured by their geographical coordinates of residence.
 Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain.
@@
 Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km.
-(Bottom) Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right).
+\textbf{(b)} Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right).
 Individuals with missing data were excluded.
@@
 
-\begin{figure}
-    \centering
-    \begin{subfigure}{\columnwidth}
-    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
-    \caption{Each point is an individual placed based on where they live.
+%\begin{figure}[h!]
+%  \centering
+%  \begin{subfigure}[b]{0.4\linewidth}
+%    \includegraphics[width=\linewidth]{Figures/1KGP_UMAP_labels.jpeg}
+%  \end{subfigure}
+%  \rulesep
+%  \begin{subfigure}[b]{0.4\linewidth}
+%    \includegraphics[width=\linewidth]{Figures/1KGP_PC15_UMAP_labels.jpeg}
+%  \end{subfigure}
+%  \caption{UMAP embeddings of the 1KGP genotype data (left) and the top 15 principal components of the genotype data (right).
+Each point is coloured and labelled by its population.
+See Appendix~\ref{AppendixA} for label definitions.}
+%  \label{fig:1KGP}
+%\end{figure}
+
+%\begin{figure}
+%    \centering
+%     \textbf{Test}\par\medskip
+%    \begin{subfigure}{\columnwidth}
+%    \includegraphics[width=0.35\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
+%     %\includegraphics[width=0.55\columnwidth]{images/meanWorldMap.jpg}
+%    \caption{Each point is an individual placed based on where they live.
 Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
-    \label{fig:umap_ukbb_geo_map}
-    \end{subfigure}
-     \begin{subfigure}{\columnwidth}
-         \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
-   \caption{Geographic distribution of UMAP coordinates.
+%    \label{fig:umap_ukbb_geo_map}
+%    \end{subfigure}
+%          \rulesep
+%     \begin{subfigure}{\columnwidth}
+%         \includegraphics[width=0.45\columnwidth]{images/meanWorldMap.jpg}
+%   \caption{Geographic distribution of UMAP coordinates.
 Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
@@
 Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
-Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States \sgcomment{rationale?}.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States.
+\sgcomment{rationale?}.
 Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
-    \label{fig:umap_world_map}
-      \end{subfigure}
-      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.
+%    \label{fig:umap_world_map}
+%      \end{subfigure}
+%      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.
 Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates.
@@
 }}
+%\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.4\linewidth}
+    \includegraphics[width=\linewidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
+    \caption{}
+    \label{fig:umap_ukbb_geo_map}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.4\linewidth}
+    \includegraphics[width=\linewidth]{images/meanWorldMap.jpg}
+    \caption{}
+  \label{fig:umap_world_map}
+  \end{subfigure}
+  \caption{\textbf{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.} Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (a flattened projection is in the top right of \textbf{(a)}.
+Individuals who are closer to each other in the projection will be closer in colour in the maps.
+More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.
+\textbf{(a)} Each point is an individual placed based on where they live.
+Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.
+\textbf{(b)} Geographic distribution of UMAP coordinates.
+Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
+Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
+Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history.
+Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.
+\sgcomment{rationale?}.
+Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
 \end{figure}
@@
 
-Using UMAP reveals intriguing clusters that would otherwise have been very difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}.
+Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}\adpcomment{PCA on the Gujarati alone followed by FST.
+I guess the highlight is it was found without having to look at a sub-population specifically}.
 Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering.
@@
 
-UMAP works well at highlighting groups of individuals who share more ancestry; when clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters.
-The 1KGP dataset was deliberately crafted to come from geographically separated populations that are homogeneous internally.
-Thus, most of its populations form isolated clusters, with the exception of those that most closely share ancestries.
-\sgcomment{I don't understand what this paragraph is trying to say.
-You might be a bit too wordy here.
-}
+\adpcomment{(scrapped a paragraph here)}
 
@@
 \sgcomment{Is this really a "material and method" comment?
-THe logical flow with the last comment is not clear to me.}
+THe logical flow with the last comment is not clear to me.} \adpcomment{This was moved to trim some text from the main body, I can just remove it as it's redundant with info from 2 paragraphs earlier.}
 
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia.
+Geographic distribution of UMAP coordinates.
+Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
+Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
+Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history.
+Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.}
     \label{fig:supp_umap_ukbb_asia}
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean.
+Geographic distribution of UMAP coordinates.
+Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
+Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
+Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history.
+Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.}
     \label{fig:supp_umap_ukbb_car}
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
+    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe.
+Geographic distribution of UMAP coordinates.
+Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
+Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
+Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history.
+Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
+Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.
+}
     \label{fig:supp_umap_ukbb_eur}

### 2019-05-29 ms/main.tex
some edits

@@
 The clarity of the interconnected clusters varies by parameterization, but they consistently form a large, roughly connected group.
+\sgcomment{Last sentence is a bit vague.
+Remove?}
 
@@
 Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group".
-Using data on countries of birth, we identified many groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
-
-Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection in figure~\ref{fig:umap_ukbb_pc10} coloured in by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
+Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
+
+Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
 Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}.
-Most admixture lines have coordinates in the south east of the UK, corresponding to the city of London and reflecting its high migrant population.
-The detailed shape of extended clusters is not stable as we vary the number of PCs included.
+Most admixture lines have coordinates in the south east of the UK, where the city of London isa located, likely reflecting its high migrant population.
+The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved.
 Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.
-The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.
+%The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.
 
@@
 White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau.
-Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian/Native American, and European populations using RFMIX \cite{Maples:2013fia}.
+Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
 We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.

### 2019-05-29 ms/main.tex
more edits

@@
 Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; using available country of birth data, most of the unidentified ethnic group were born in Japan.
+Figure~\ref{fig:supp_comparison_fev_chi_eur} also shows an important difference in FEV1 between self-identified Chinese and a nearby cluster enriched in individuals born in Japan.
+\sgcomment{boxblot}.
+
 These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).
@@
 \subsection*{Discussion}
-UMAP comprehensively illustrates genotypic information at fine-scales and within the context of a global population structure.
+UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure.
 Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
@@
 
-Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}\adpcomment{PCA on the Gujarati alone followed by FST.
-I guess the highlight is it was found without having to look at a sub-population specifically}.
-Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering.
+Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through global PCA or Admixture analysis alone.
+
+%Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering.
 Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.
 
-\adpcomment{(scrapped a paragraph here)}
+Using UMAP on the UKBB data revealed a range of patterns of covariation between geography, phenotypes and genotypes, with traits such as height showing continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.
+
+We found that pre-processing the data with PCA allowed for considerable time savings.
+Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
+Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
+In our experience, we find it useful to include as many PCs as are available.
 
@@
 For this reason, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
-
-We found that pre-processing the data with PCA allowed for considerable time savings.
-Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
-Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
 

### 2019-05-29 ms/main.tex
minor text tweaks

--- 2019-05-29 ms/main.tex
@@
     \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
-    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry.
+    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian, and European continental ancestry.
 Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians.
@@
 
-The HRS forms several large groupings and clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}).
+The HRS forms several large clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}).
 Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics.
 The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}).
-The clarity of the interconnected clusters varies by parameterization, but they consistently form a large, roughly connected group.
-\sgcomment{Last sentence is a bit vague.
-Remove?}
-
-To investigate possible ancestries related to populations in the 1KGP we took two approaches.
+Projecting 1KGP populations onto the embedding reveals substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}).
+One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.
+
+%To investigate possible ancestries related to populations in the 1KGP we took two approaches.
 In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}).
@@
 \end{subfigure}%
-\caption{\textbf{The UKBB coloured by self-reported ethnic background.} \textbf{(a)} The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations.
+\caption{\textbf{The UKBB coloured by self-reported ethnic background.} \textbf{(a)} The first two principal components, showing the usual triangle with vertices corresponding to African, Asian, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations.
 \textbf{(b)} UMAP on the first 10 principal components.
@@
 
-The largest body in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations.
-The Irish population concentrates in one part of this group, but many individuals are also scattered throughout the British-identified population.
+The largest cluster in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations.
+The Irish population forms a sub-cluster, but many individuals are also scattered throughout the British-identifying population.
 Individuals identifying as Black African and Black Caribbean partially overlap, but admixed individuals form distinct trails leading to Asian and European clusters.
-Chinese individuals form a cluster, within a broader East Asian super-population; Indian, Pakistani, and Bangladeshi populations form a closely bound group as well.
-The East Asian and South Asian super-populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group".
+Chinese individuals form a cluster, within a broader East Asian population; Indian, Pakistani, and Bangladeshi populations form a closely bound cluster as well.
+The East Asian and South Asian populations each have large clusters of individuals who identify as having an "other Asian background" or belonging to an "other ethnic group".
 The patchwork of genetic neighbourhoods is connected by trails of admixed individuals, which converge at a nexus of individuals with a variety of ethnicities.
 Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group".
-Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino groups being projected near Chinese groups.
+Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino clusters being projected near Chinese clusters.
 
@@
 Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}.
-Most admixture lines have coordinates in the south east of the UK, where the city of London isa located, likely reflecting its high migrant population.
+Most admixture lines have coordinates in the south east of the UK, where the city of London is located, likely reflecting its high migrant population.
 The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved.

### 2019-06-03 ms/main.tex
fix some text, add more FEV1 stuff for East Asia

@@
 Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Figure~\ref{fig:supp_comparison_fev_chi_eur} also shows an important difference in FEV1 between self-identified Chinese and a nearby cluster enriched in individuals born in Japan.
-\sgcomment{boxblot}.
+Figure~\ref{fig:supp_comparison_fev_chi_eur} also shows an important difference in FEV1 between self-identified Chinese and a nearby cluster enriched in individuals born in Japan\ref{fig:supp_fev_ridgeplots}.
 
@@
 
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/east_asia_ridgeplots_fev.png}
+    \end{subfigure}
+    \caption{Plots of the distributions of FEV1 by sex for East Asian populations.
+Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from figure~\ref{fig:supp_comparison_fev_chi_eur}.
+The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.
+Asterisks indicate significant difference from the Japanese population, using Welch's unpaired t-test with a Bonferroni correction for 3 groups.
+The dashed lines are the means of the distributions and Japanese populations have consistently higher means, though the test between Chinese and Japanese males fell short of the significance threshold.}
+    \label{fig:supp_fev_ridgeplots}
+\end{figure*}
+
 \begin{figure}[!htb]

### 2019-06-03 ms/main.tex
more cleaning

@@
 
+\adpcomment{trim this pgh?}
 Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority.
@@
 
-\subsubsection*{The genetic continuum of admixed populations} The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters.
+\subsubsection*{The genetic continuum of admixed populations}
+The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters.
 Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas.
@@
 Projecting 1KGP populations onto the embedding reveals substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}).
-One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.
-
-%To investigate possible ancestries related to populations in the 1KGP we took two approaches.
-In the first we generated PC axes and a UMAP embedding using HRS and 1KGP data together (figure~\ref{fig:supp_hrs_1000g}).
-In the second we used the PC axes and UMAP embedding generated in figure~\ref{fig:umap_hrs_admix} and projected 1KGP data onto it (figure~\ref{fig:supp_hrs_1kgp_projected}).
-Both approaches reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population.
 One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.
@@
 
-%\begin{figure}
-%    \centering
-%    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
-%    \caption{A map of Great Britain coloured by a 3D UMAP projection.
-Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in colour.
-Patterns in genetic similarity are visible in Scotland, South England, the East and West Midlands, and major urban centres.
-A flattened 2D view of of the 3D projection used for colouring is presented in the top right.
-To protect participant privacy, data has been randomized as explained in the materials and methods.}
-%    \label{fig:umap_ukbb_geo_map}
-%\end{figure}
-
-%\begin{figure}[h!]
-%  \centering
-%  \begin{subfigure}[b]{0.4\linewidth}
-%    \includegraphics[width=\linewidth]{Figures/1KGP_UMAP_labels.jpeg}
-%  \end{subfigure}
-%  \rulesep
-%  \begin{subfigure}[b]{0.4\linewidth}
-%    \includegraphics[width=\linewidth]{Figures/1KGP_PC15_UMAP_labels.jpeg}
-%  \end{subfigure}
-%  \caption{UMAP embeddings of the 1KGP genotype data (left) and the top 15 principal components of the genotype data (right).
-Each point is coloured and labelled by its population.
-See Appendix~\ref{AppendixA} for label definitions.}
-%  \label{fig:1KGP}
-%\end{figure}
-
-%\begin{figure}
-%    \centering
-%     \textbf{Test}\par\medskip
-%    \begin{subfigure}{\columnwidth}
-%    \includegraphics[width=0.35\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
-%     %\includegraphics[width=0.55\columnwidth]{images/meanWorldMap.jpg}
-%    \caption{Each point is an individual placed based on where they live.
-Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
-%    \label{fig:umap_ukbb_geo_map}
-%    \end{subfigure}
-%          \rulesep
-%     \begin{subfigure}{\columnwidth}
-%         \includegraphics[width=0.45\columnwidth]{images/meanWorldMap.jpg}
-%   \caption{Geographic distribution of UMAP coordinates.
-Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there.
-Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America.
-Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history.
-Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania.
-Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States.
-\sgcomment{rationale?}.
-Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
-%    \label{fig:umap_world_map}
-%      \end{subfigure}
-%      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.
-Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates.
-Individuals who are closer to each other in the projection will be closer in colour in the maps.
-More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.
-\sgcomment{Merge captions.
-Make sure to have overall title!
-}}
-%\end{figure}
-
 \begin{figure}[h!]
@@
 Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.
-\sgcomment{rationale?}.
 Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
@@
 Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Figure~\ref{fig:supp_comparison_fev_chi_eur} also shows an important difference in FEV1 between self-identified Chinese and a nearby cluster enriched in individuals born in Japan\ref{fig:supp_fev_ridgeplots}.
-
-These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).
+Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
+An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$).
+Distributions of FEV1 from Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.
+
+%These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).
 
@@
 However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
-Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}.
+Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
 Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
@@
 Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through global PCA or Admixture analysis alone.
-
-%Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering.
-Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.
-
-Using UMAP on the UKBB data revealed a range of patterns of covariation between geography, phenotypes and genotypes, with traits such as height showing continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.
+With the UKBB data we revealed a range of patterns of covariation between geography, phenotypes and genotypes.
+Traits such as height showed continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showed sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.
 
@@
 Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
-In our experience, we find it useful to include as many PCs as are available.
+In our experience, we find it useful to include as many PCs as are available and computationally feasible.
 
@@
 
-Both UMAP and t-SNE are sensitive to sample size, and spend more visual real estate for populations with larger sample sizes.
+UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes.
 This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB.
@@
 This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations.
-For this reason, differences between Ireland/the UK and other countries appear exaggerated.
-Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing.
-We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}.
-\sgcomment{Is this really a "material and method" comment?
-THe logical flow with the last comment is not clear to me.} \adpcomment{This was moved to trim some text from the main body, I can just remove it as it's redundant with info from 2 paragraphs earlier.}
 
@@
 %%%%% boxplots end
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
-    \end{subfigure}
-    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
-    \label{fig:supp_umap_ukbb_worldmap_alt}
-\end{figure*}
+%\begin{figure*}
+%    \centering
+%    \begin{subfigure}{\textwidth}
+%    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
+%    \end{subfigure}
+%    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
+%    \label{fig:supp_umap_ukbb_worldmap_alt}
+%\end{figure*}
 

### 2019-06-03 ms/main.tex
moar tweaks

@@
 British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry.
-Toscani and Iberian individuals form a group reflecting their Mediterranean heritage.
+Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage.
 African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster.
-Three East Asian groups appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population.
-Looser geographical groupings include Colombians and Peruvians, and the Esan and Yoruba populations of Nigeria; both groupings appear as connected sub-clusters.
-The South Asian populations also form a loose grouping.
-
-\adpcomment{trim this pgh?}
-Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority.
+Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population.
+Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.
+
+\adpcomment{changed this a bit. or maybe remove most of it?}
+A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone.
+Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan.
+This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}).
+Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}.
+Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym.
+[D.
+Poznik, 23andme, personal communication, and \cite{23andme}].
+
+%of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority.
 One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone.
@@
 
-\subsubsection*{The genetic continuum of admixed populations}
+%The genetic continuum of admixed populations
+\adpcomment{cleaned up, changed subtitle}
+\subsubsection*{Admixed individuals fall along a genetic continuum}
 The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters.

### 2019-06-05 ms/main.tex
couple of small edits

@@
 Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations.
-If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ typically larger than $100,000$.
+If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions.
 Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.
@@
 We found that pre-processing the data with PCA allowed for considerable time savings.
-Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components.
+Groupings on ethnicity formed slowly as PCs were added until reaching a stable number around 10 to 15 PCs.
 Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.

### 2019-06-11 ms/main.tex
minor cleanup and text changes

--- 2019-06-03 ms/main.tex
@@
     \caption{\textbf{Applying UMAP to subets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
-Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
+Colouring the points by birthplace shows they were born almost entirely in the Mountain region (in green) of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
 When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster.
@@
 \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
-Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
+Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.
+Colours represent self-identified ethnic background, with those identifying as Indian coloured purple.
+}
     \label{fig:subpops}
@@
 
-\adpcomment{changed this a bit. or maybe remove most of it?}
 A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone.
@@
 
-%of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority.
-One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone.
-Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan.
-This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}).
-Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}.
-Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym.
-[D.
-Poznik, 23andme, personal communication, and \cite{23andme}].
-
-%The genetic continuum of admixed populations
-\adpcomment{cleaned up, changed subtitle}
 \subsubsection*{Admixed individuals fall along a genetic continuum}
@@
 The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}).
-Projecting 1KGP populations onto the embedding reveals substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}).
+By generating the PC axes and UMAP embedding for the HRS data in figure~\ref{fig:supp_umap_hrs_eth}, and projecting the 1KGP data onto it, we reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}).
 One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.
@@
 Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.
-%The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.
 
@@
 
-%These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).
-
 \subsubsection*{Comparing t-SNE and UMAP}
@@
 Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
-Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
-
-%To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data).
-The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}).
-When applied to the UKBB, UMAP provides a more compelling visualization of the data compared to both versions of t-SNE.
-Interestingly, the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), suggesting that UMAP can help standard t-SNE implementation to reach the t-SNE objective.
-The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but lost much fine-scale resolution and the relationship among clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
-An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files.
 Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
@@
 We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+To project 1KGP data on HRS embeddings, as in figures~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} and \ref{fig:supp_hrs_1kgp_projected}, we created the PC axes and UMAP embedding for the HRS data and then projected the 1KGP data onto them.
 
@@
     \end{subfigure}
-    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions.}
+    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions.
+Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.}
     \label{fig:supp_umap_hrs_hisp_admix}
@@
 colouring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
-This projection coloured by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
+This projection coloured by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.
+Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.}
     \label{fig:umap_hrs_admix}
@@
     \includegraphics[width=0.95\columnwidth]{images/1KGP_tsne_umap.jpeg}
-    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP. t-SNE used 5000 iterations.}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP.
+t-SNE used 5000 iterations.
+Initializing t-SNE with UMAP breaks the continuous structure of the projection and instead forms many small clusters.}
     \label{fig:supp_tsne_umap_compare_1kgp}

### 2019-06-12 ms/main.tex
tweaks

@@
     \centering
-    \begin{subfigure}{0.5\columnwidth}
+    \begin{subfigure}{0.49\columnwidth}
     \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
@@
     \end{subfigure}\hfill
-        \begin{subfigure}{0.5\columnwidth}
+    \unskip\ \vrule\
+        \begin{subfigure}{0.49\columnwidth}
     \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
@@
     \end{subfigure}
-    \caption{\textbf{Applying UMAP to subets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
+    \caption{\textbf{Applying UMAP to subsets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
 Colouring the points by birthplace shows they were born almost entirely in the Mountain region (in green) of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).

### 2019-06-18 ms/main.tex
Add some stuff on PCs and variation

@@
     \end{subfigure}
-    \caption{UMAP applied to the first few hundred principal components of the 1KGP data.
+    \caption{UMAP applied to the first few hundred principal components of the 1KGP data with the amount of variance explained in parentheses.
 As more components are added, the figure begins to resemble that of UMAP carried out on the full genotype dataset.}
@@
     \end{subfigure}
-    \caption{UMAP applied the last 3350 principal components of the 1KGP.}
+    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation.}
     \label{fig:supp_1kgp_3350}
@@
 
+\clearpage
+\begin{table}[htb!]
+\centering
+ \begin{tabular}{||c c||} 
+ \hline
+ Number of PCs & Variance explained  \\ [0.5ex] 
+ \hline\hline
+ 2 & 13.6\% \\
+\hline
+3 & 14.7\% \\
+\hline
+4 & 15.5\% \\
+\hline
+5 & 15.7\% \\
+\hline
+6 & 15.8\% \\
+\hline
+7 & 15.9\% \\
+\hline
+8 & 16.0\% \\
+\hline
+9 & 16.1\% \\
+\hline
+10 & 16.2\% \\
+\hline
+11 & 16.2\% \\
+\hline
+12 & 16.3\% \\
+\hline
+13 & 16.4\% \\
+\hline
+14 & 16.5\% \\
+\hline
+15 & 16.5\% \\
+\hline
+30 & 17.5\% \\
+\hline
+50 & 18.6\% \\
+\hline
+100 & 21.3\% \\
+\hline
+200 & 26.4\% \\
+\hline
+300 & 31.1\% \\
+\hline
+400 & 35.5\% \\
+\hline
+500 & 39.7\% \\
+\hline
+600 & 43.7\% \\
+\hline
+700 & 47.4\% \\
+\hline
+800 & 50.8\% \\
+\hline
+900 & 54.0\% \\
+\hline
+1000 & 57.0\% \\
+\hline
+1100 & 60.0\% \\
+\hline
+1200 & 62.8\% \\
+\hline
+1300 & 65.4\% \\
+\hline
+1400 & 68.0\% \\
+\hline
+1500 & 70.5\% \\
+\hline
+1600 & 73.0\% \\
+\hline
+1700 & 75.3\% \\
+\hline
+1800 & 77.5\% \\
+\hline
+1900 & 79.7\% \\
+\hline
+2000 & 81.8\% \\
+\hline
+2100 & 83.8\% \\
+\hline
+2200 & 85.8\% \\
+\hline
+2300 & 87.6\% \\
+\hline
+2400 & 89.4\% \\
+\hline
+2500 & 91.1\% \\
+\hline
+2600 & 92.7\% \\
+\hline
+2700 & 94.2\% \\
+\hline
+2800 & 95.3\% \\
+\hline
+2900 & 96.3\% \\
+\hline
+3000 & 97.1\% \\
+\hline
+3100 & 97.8\% \\
+\hline
+3200 & 98.5\% \\
+\hline
+3300 & 99.2\% \\
+\hline
+3400 & 99.7\% \\[1ex] 
+ \hline
+ \end{tabular}
+ \caption{Variance explained in the 1KGP data by the number of principal components used.}
+ \label{table:1KGP_var_exp}
+\end{table}
+
 \end{document}

### 2019-06-18 ms/main.tex
clean up HRS Hispanic projections

@@
     \begin{subfigure}{0.49\columnwidth}
-    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
+    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.jpeg}
     \caption{}
@@
     \end{subfigure}
-    \caption{\textbf{Applying UMAP to subsets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted).
+    \caption{\textbf{Applying UMAP to subsets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster.
 Colouring the points by birthplace shows they were born almost entirely in the Mountain region (in green) of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana).
@@
     TSI, Tuscani in Italy.
-Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions.
+Figures~\ref{fig:supp_umap_hrs_hisp_admix} and \ref{fig:supp_umap_hrs_hisp_birth} present the same projection of individuals from the HRS coloured by estimated admixture proportions census region of birth, respectively.
 \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
@@
     \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp_birthonly.jpeg}
+    \end{subfigure}
+    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured region of birth.}
+    \label{fig:supp_umap_hrs_hisp_birth}
+\end{figure*}
+
+\begin{figure*}
+    \centering
     \begin{subfigure}{0.95\textwidth}

### 2019-06-18 ms/main.tex
add alt colour for HRS

@@
 Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
-We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.}
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+An alternate colouring is provided in figure~\ref{fig:umap_hrs_admix_alt}}
     \label{fig:umap_hrs_admix}
+\end{figure}
+
+\begin{figure}
+\centering
+   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix132.jpeg}
+   \caption{An alternate colouring of \ref{fig:umap_hrs_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
+    \label{fig:umap_hrs_admix_alt}
 \end{figure}

### 2019-06-18 ms/main.tex
add other alt HRS colouring

@@
 Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
-We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.}
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+An alternate colouring is provided in figure~\ref{fig:supp_umap_hrs_hisp_admix_alt}.}
     \label{fig:supp_umap_hrs_hisp_admix}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
+    \end{subfigure}
+    \caption{An alternate colouring of figure~\ref{fig:supp_umap_hrs_hisp_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
+    \label{fig:supp_umap_hrs_hisp_admix_alt}
 \end{figure*}

### 2019-06-19 ms/main.tex
add stuff on ukbb_asian_pops

@@
         \begin{subfigure}{0.49\columnwidth}
-    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
+    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256_ethcob.jpeg}
     \caption{}
@@
 \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
-Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.
-Colours represent self-identified ethnic background, with those identifying as Indian coloured purple.
+Indian individuals born in Kenya (in purple) form one such cluster.
+A version coloured by self-identified ethnicity is presented in figure~\ref{fig:supp_umap_ukbb_asian_eth}.
 }
@@
     \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.8\textwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256.jpeg}
+    \end{subfigure}
+    \caption{UMAP of the first 8 principal components of the Asian population in the UKBB coloured by self-identified ethnicity. This is an alternate colouring of figure~\ref{fig:umap_ukbb_indian}}
+    \label{fig:supp_umap_ukbb_asian_eth}
+\end{figure*}
+
+\begin{figure*}
+    \centering
     \begin{subfigure}{0.95\textwidth}

### 2019-06-20 ms/main.tex
tidying, add some stuff on families

@@
 Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
-Patterns are also evident when using UMAP on the last components (\ref{fig:supp_1kgp_3350}).
+Patterns are evident even when using UMAP on the last components (figure~\ref{fig:supp_1kgp_3350}).
 
@@
 Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.
+Within population clusters, family members were projected near each other but did not form disjoint clusters unless UMAP was parameterized to use a minimal number of neighbours (figure~\ref{fig:supp_1kgp_families}).
 
@@
 Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan.
-This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}).
+This clustering effect is robust to the number of components considered (figure~\ref{fig:supp_megamontage_pc10_50}).
 Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}.
@@
 
-Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through global PCA or Admixture analysis alone.
+Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1KGP that were not visible through global PCA or Admixture analysis alone.
 With the UKBB data we revealed a range of patterns of covariation between geography, phenotypes and genotypes.
@@
     \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.52\textwidth]{images/1KGP_UMAP_PC15_families.png}
+    \end{subfigure}
+    \caption{UMAP applied to the first 15 principal components of the 1KGP, with the number of neighbours set to 5 (top) and 15 (bottom).
+Six members of one Southern Han Chinese family are highlighted: HG00656 (grandfather), HG00657 (grandmother), HG00658 (uncle, mother's brother), HG00701 (mother), HG00702 (father), HG00703 (child).
+When using UMAP with five neighbours, the father (in blue) is projected to the cluster of the Southern Han Chinese population while the rest of the family members (in red) form their own disjoint cluster.
+Using 15 neighbours, the family still clusters together, but as part of the Southern Han Chinese population.}
+    \label{fig:supp_1kgp_families}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
     \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}

### 2019-06-26 ms/main.tex
some more FEV1 stuff plus distance from London

@@
 Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}.
-Most admixture lines have coordinates in the south east of the UK, where the city of London is located, likely reflecting its high migrant population.
+Colouring the projection by distance from London reflects the high migration levels to the city and surrounding area; individuals with more admixed backgrounds tend to have coordinates in the south east of the UK, as visible in figure~\ref{fig:supp_london_distance}.
 The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved.
@@
 Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$).
-Distributions of FEV1 from Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.
+An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller.
+Distributions of FEV1 by sex from selected Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.
 
@@
 Because this process is done independently for each visualization, a given point shown on the figure will copy values from different randomly selected individuals.
-Additionally, spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km.
+Projections coloured by participants' spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km.
+Projections coloured by participants' distance from London have random noise added (normally distributed about 0 with a standard deviation of 5km) before binning to the nearest km.
 For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}
@@
     \centering
+    \begin{subfigure}{0.9\textwidth}
+    \includegraphics[width=0.9\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_london_permuted_10nn_sd10_base1_maxdist200_2019625172227_resize.jpeg}
+    \end{subfigure}
+    \caption{UMAP on UKBB data, coloured by distance from London, with red representing those living closer to London and blue representing those living farther from London.
+A 200km radius extends roughly to Cardiff, and a 100km radius extends roughly to cities such as Leicester and Bath, and contains cities such as Oxford, Cambridge, and Peterborough.
+Data has been randomized as explained in the materials and methods section.}
+    \label{fig:supp_london_distance}
+\end{figure*}
+
+\begin{figure*}
+    \centering
     \begin{subfigure}{0.95\textwidth}
@@
     \centering
+    \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=0.5\textwidth]{images/east_asia_population_colours.png}
+    \end{subfigure}
+    \caption{Individuals from the zoomed in section in figure~\ref{fig:supp_comparison_fev_chi_eur} used in statistical testing, coloured the same as in figure~\ref{fig:supp_fev_ridgeplots}.
+Brown, blue, and green represent those born in the Philippines, Malaysia, and Japan; pink represents those who self-identify as Chinese.
+The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.}
+    \label{fig:supp_fev_test_pops}
+\end{figure*}
+
+\begin{figure*}
+    \centering
     \begin{subfigure}{\textwidth}
@@
     \end{subfigure}
-    \caption{Plots of the distributions of FEV1 by sex for East Asian populations.
+    \caption{Plots of the distributions of residual FEV1 by sex for East Asian populations, after adjusting for height, age, age$^2$, and sex through linear regression.
 Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from figure~\ref{fig:supp_comparison_fev_chi_eur}.
@@
 Asterisks indicate significant difference from the Japanese population, using Welch's unpaired t-test with a Bonferroni correction for 3 groups.
-The dashed lines are the means of the distributions and Japanese populations have consistently higher means, though the test between Chinese and Japanese males fell short of the significance threshold.}
+The dashed lines are the means of the distributions, and Japanese populations have consistently higher means.}
     \label{fig:supp_fev_ridgeplots}

### 2019-06-27 ms/main.tex
tweak text

@@
 Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller.
-Distributions of FEV1 by sex from selected Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.
+An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
 

### 2019-07-03 ms/main.tex
tweak text

@@
 Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
+An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
 

### 2019-07-03 ms/main.tex
ciiiites

@@
 The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds.
+Comparing UMAP to PCA with this data, the former captures local and fine-scale structure well while the latter identifies major axes of variation in figure~\ref{fig:fig_ukbb}.
 UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes.
@@
 Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
-Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}.
+Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe and British-wide data\cite{novembre2008europe,leslie2015fine}.
 Colouring the projection by distance from London reflects the high migration levels to the city and surrounding area; individuals with more admixed backgrounds tend to have coordinates in the south east of the UK, as visible in figure~\ref{fig:supp_london_distance}.

### 2019-07-03 ms/main.tex
edits throughout

@@
 Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe and British-wide data\cite{novembre2008europe,leslie2015fine}.
-Colouring the projection by distance from London reflects the high migration levels to the city and surrounding area; individuals with more admixed backgrounds tend to have coordinates in the south east of the UK, as visible in figure~\ref{fig:supp_london_distance}.
+Figure~\ref{fig:umap_ukbb_pc10} shows that the admixed individuals have UMAP coordinates next to White British individuals residing in the South East of the UK, where London is located (see also figure~\ref{fig:supp_london_distance}, where individuals are coloured by distance from London).
+This likely reflects the high migration levels to the city and surrounding area: the UMAP projection attempted to preserve both the genetic similarity among admixed individuals and the relatedness with White British individuals in cosmopolitan areas.
+
  The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved.

### 2019-07-04 ms/main.tex
edits up to line 85

@@
 Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed.
-Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference}.
+Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference} \sgcomment{This is a lone reference for a broad topic...
+Are there more recent reviews perhaps?
+Else add a couple of recent references eg by the novembre lab?}.
 
 Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations.
-If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions.
+If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of copies of the derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions.
 Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.
@@
 The amount of information encoded in the lower-variance principal components increases with sample size, so researchers typically examine multiple two-dimensional projections to explore data.
-Some features may be hidden by the projections or hard to interpret.
-To display more of the high-dimensional features of the data in a two dimensional figure, we can use non-linear transformations that seek to preserve the local structure of the data.
-A popular method for visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}.
+In this process, finer features of the data may be hidden by the projections or hard to interpret.
+To display finer features of the data in a two dimensional figure, we can use non-linear transformations that emphasize the local structure of the data.
+A popular method for such visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}.
 t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}; using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}.
@@
 Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
-Patterns are evident even when using UMAP on the last components (figure~\ref{fig:supp_1kgp_3350}).
-
-Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries.
+To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs).
+Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.
+\sgcomment{Note figure S4 lacks a color code -- at least point out that it uses the same coloring as X}
+
+Focusing on UMAP with the leading 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries.
 British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry.

### 2019-07-04 ms/main.tex
edits up to line 159

@@
 Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.
-Within population clusters, family members were projected near each other but did not form disjoint clusters unless UMAP was parameterized to use a minimal number of neighbours (figure~\ref{fig:supp_1kgp_families}).
+
+Within population clusters, family members were projected near each other within broader population groups.
+When UMAP was parameterized to use only 5 nearest neighbours, however, families often formed distinct clusters (figure~\ref{fig:supp_1kgp_families}): Using few neighbors to build a manifold emphasizes closer relatedness.
 
@@
 The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds.
-Comparing UMAP to PCA with this data, the former captures local and fine-scale structure well while the latter identifies major axes of variation in figure~\ref{fig:fig_ukbb}.
-UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies continuous structure within the British population as well as admixture gradients despite the very unbalanced population sizes.
+Figure~\ref{fig:fig_ukbb} compares UMAP to PCA apply to UKBB.
+As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture.
+UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations.
 The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

### 2019-07-04 ms/main.tex
fix some images, add more stuff on FEV1, add citation on AFR/EUR FEV1 study

--- 2019-07-03 ms/main.tex
@@
 Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
+Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
+Though studies in several populations have shown an inverse relationship between African ancestry and FEV1 levels compared to European populations\cite{ortega2015effect}, we are not aware of such results for specific Asian ancestries.
+We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines (the largest populations in the UKBB by country of birth) as well as the self-identified Chinese population (mutually exclusive groups, shown in figure~\ref{fig:supp_fev_test_pops}).
+After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$).
+By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
+For comparison, the adjusted difference between self-identified African and British individuals is 0.762 (p-value $2.2\times 10^{-16}$).
 

### 2019-07-05 ms/main.tex
Revert "fix some images, add more stuff on FEV1, add citation on AFR/EUR FEV1 study"
This reverts commit 9f088e6fd106a716a22a0fbcb61aa0dfc3226ec4.

--- 2019-07-04 ms/main.tex
@@
 As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}).
-The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.
+The geographical patterns outside major urban centers are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.
 Using data about country of birth, we performed a similar analysis of a world map~\ref{fig:umap_world_map}, revealing subtle regional variation around the world.
@@
 
-\subsubsection*{Patterns in phenotype related to population structure} 
+\subsubsection*{Identifying patterns in phenotype variation related to genetic population structure}
+
 Covariates such as height and leukocyte count (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure.
 Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and others have noticeably lower-than-average heights.
-More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t-test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
+More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals.
+To quantify and statistically test these qualitative observations, we performed an unpaired two sample t-test of self-identified White Irish and White British individuals and found British males taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
 Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.
@@
 This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds.
-Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations.
-Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
+Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations \sgcomment{point to box plots again?}.
+
+Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
+An unpaired two-sample t-test shows those born in Japan \sgcomment{if test is not about those born in Japan, needs to be clarified!} have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (see figure~\ref{fig:supp_fev_ridgeplots} for distributions).
+For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.
+
+Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups.
+This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.
 
 \subsubsection*{Comparing t-SNE and UMAP}
-
-Identifying the best dimension reduction technique is to an extent subjective.
-However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
+\sgcomment{tweaked this a bit -- would be good to include in reply to reviewer}
+Identifying the best dimension reduction technique is challenging, both because the `best' representation depends on context, and because convergence issues may mean that a good theoretical model for dimensional reduction might perform poorly because of challenges in numerical optimization.
+To assess whether the relatively poor performance of t-SNE could be due to convergence rather than a flawed model, we used UMAP to preprocess the UKBB data and provide a starting point to a standard t-SNE implementation.
+This led to representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
 Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
@@
 UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure.
-Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
+It is easy to use and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
 It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
@@
 
-Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1KGP that were not visible through global PCA or Admixture analysis alone.
-With the UKBB data we revealed a range of patterns of covariation between geography, phenotypes and genotypes.
+Using UMAP reveals  clusters that would have been difficult to identify via pairwise PCA plots or Admixture analysis, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1KGP.
+More importantly, UMAP helps reveal patterns of covariation between geography, phenotypes and genotypes.
 Traits such as height showed continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showed sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.
 
-We found that pre-processing the data with PCA allowed for considerable time savings.
+We found that pre-processing the data with PCA allowed for time savings, but identifying an optimal number of PCs to use is challenging.
 Groupings on ethnicity formed slowly as PCs were added until reaching a stable number around 10 to 15 PCs.
 Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
-In our experience, we find it useful to include as many PCs as are available and computationally feasible.
-
+
+\subsubsection*{Caveats}
+The existence of a (small) number of adjustable parameter to UMAP, including PC cutoff and number of neighbors, opens the door to tweaking.
+Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices.
+In the absence of clear theoretical rationales, we generally recommend to use as many PCs as are available and computationally feasible, and to use the default number of neighbours (15).
+We sometimes found that lower number of PCs led to a simpler shape that facilitated discussion (e.g., figure \ref{fig:fig_ukbb}).
+We recommend reporting on a range of parameter values and following up on observations with statistical testing.
 Like most non-linear methods, UMAP lacks direct interpretability.
-It is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are distant from each other are not necessarily very different from each other.
-Figure~\ref{fig:supp_megamontage_pc10_50} demonstrates that disconnected population clusters can appear at different locations depending on parameter choices.
-For this reason, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
+It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data.
+For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
 
 UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes.
-This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB.
+This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB.
 We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.

### 2019-07-05 ms/main.tex
add some text on FEV1 and ciations

@@
 This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds.
-Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations \sgcomment{point to box plots again?}.
-
-Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-An unpaired two-sample t-test shows those born in Japan \sgcomment{if test is not about those born in Japan, needs to be clarified!} have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (see figure~\ref{fig:supp_fev_ridgeplots} for distributions).
-For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.
-
-Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups.
+Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations (figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m}).
+
+\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage.}
+Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
+Baseline FEV1 levels are determined via equations that include ancestry or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}; we are not aware of any results comparing populations within Asia.
+We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines (the largest populations in the UKBB by country of birth) as well as the self-identified Chinese population (mutually exclusive groups, shown in figure~\ref{fig:supp_fev_test_pops}).
+After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$).
+By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
+For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).
+
+%For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.
+
+%Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups.
 This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.

### 2019-07-05 ms/main.tex
more FEV1 explanation, add some citations to intro, clarify colour scheme in S4

@@
 
-%An Analysis is a new analysis of existing data (typically large genomic, transcriptomic or proteomic data sets from arrays or other high-throughput platforms) or new data obtained in a comparative analysis of technologies that lead to novel and arresting conclusions of importance to a broad audience.
-The main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words.
-The abstract is typically 100-150 words, unreferenced.
-Analyses have no more than 6 display items (figures and/or tables).
-An introduction (without heading) is followed by sections headed Results, Discussion and online Methods.
-The Results and online Methods should be divided by topical subheadings; the Discussion does not contain subheadings.
-As a guideline, Analysis allow up to 50 references.
-
 \begin{abstract}
@@
 Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed.
-Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference} \sgcomment{This is a lone reference for a broad topic...
-Are there more recent reviews perhaps?
-Else add a couple of recent references eg by the novembre lab?}.
+Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference, novembre2016recent, spence2018inference}.\adpcomment{added two citations}
 
@@
 Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.
-\sgcomment{Note figure S4 lacks a color code -- at least point out that it uses the same coloring as X}
 
@@
 
-\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage.}
-Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan.
-Baseline FEV1 levels are determined via equations that include ancestry or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}; we are not aware of any results comparing populations within Asia.
-We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines (the largest populations in the UKBB by country of birth) as well as the self-identified Chinese population (mutually exclusive groups, shown in figure~\ref{fig:supp_fev_test_pops}).
+\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
+Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations.
+Baseline FEV1 levels are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}.
+We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population.
+These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}.
+Using this approach facilitates identifying large phenotypic differences among genetically related groups, which is useful when population labels are unavailable or correlate poorly with genetic ancestry.
 After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$).
@@
 
-%For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.
-
-%Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups.
-This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.
-
 \subsubsection*{Comparing t-SNE and UMAP}
-\sgcomment{tweaked this a bit -- would be good to include in reply to reviewer}
+%\sgcomment{tweaked this a bit -- would be good to include in reply to reviewer}
 Identifying the best dimension reduction technique is challenging, both because the `best' representation depends on context, and because convergence issues may mean that a good theoretical model for dimensional reduction might perform poorly because of challenges in numerical optimization.
@@
     \end{subfigure}
-    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation.}
+    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:subpops}.}
     \label{fig:supp_1kgp_3350}

### 2019-07-08 ms/main.tex
type in figure reference
add text on caveats

@@
 \subsubsection*{Caveats}
-The existence of a (small) number of adjustable parameter to UMAP, including PC cutoff and number of neighbors, opens the door to tweaking.
+The existence of adjustable parameter, including PC cutoff, minimum distance, and number of neighbors, opens the door to tweaking.
 Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices.
-In the absence of clear theoretical rationales, we generally recommend to use as many PCs as are available and computationally feasible, and to use the default number of neighbours (15).
-We sometimes found that lower number of PCs led to a simpler shape that facilitated discussion (e.g., figure \ref{fig:fig_ukbb}).
+Using a minimal number of neighbours (e.g.
+5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact and losing visual detail, as well as possibly becoming disjoint.
+We sometimes found that a lower number of PCs led to a simpler shape that facilitated discussion (e.g.
+figure \ref{fig:fig_ukbb}).
+In the absence of clear theoretical rationales, we suggest to use as many PCs as are available and computationally feasible, and beginning with a minimum distance of 0.1 to 0.5 and the default number of neighbours (15).
 We recommend reporting on a range of parameter values and following up on observations with statistical testing.
+
 Like most non-linear methods, UMAP lacks direct interpretability.
@@
 For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
-
 UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes.
@@
     \end{subfigure}
-    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:subpops}.}
+    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:1000g_compare}.}
     \label{fig:supp_1kgp_3350}

### 2019-07-08 ms/cover_letter.tex
cover letter draft

@@
 \\\\
-We are pleased to submit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in Nature Genetics.
-Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are becoming the new frontier of genomics research.
+We are pleased to resubmit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics.
+Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are the new frontier of genomics research.
 In contrast to existing methods, this approach is able to represent both discrete and continuous population structure and the distributions of ancestries, admixture, and phenotypes within the context of such structure.
@@
 We identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region --- likely descendants of the Hispano population --- and  multiple associations between genetics, geography, and phenotypes in the UK biobank as well as population structure within 1000 Genomes Project populations.
-%\\\\
-%In other -omics fields, such as single-cell expression data, nonlinear nonparametric approaches such as t-SNE have become standard tools for data exploration.
-Principal component analysis has remained the standard for dimension reduction in human genetics, partly because nonlinear alternatives such as t-SNE don't behave well when applied to large-scale genetic data.
-We show that UMAP outperforms t-SNE computationally, qualitatively and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
 \\\\
@@
 "Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes." BioRxiv (2019): 531210.} and the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
-"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and by 23andme.
-Indeed, 23andme was able to follow up on one of the population structure signals we identified in our preprint by recontacting their participants, and the resulting interpretation of the population structure in the Gujarati is now part of the current manuscript.
-
-We believe that this is compelling evidence that UMAP will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for Nature Genetics.
+"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and by 23andMe.
+Indeed, 23andMe was able to follow up on one of the population structure signals we identified in our preprint by recontacting their participants, and the resulting interpretation of the population structure in the Gujarati is now part of the current manuscript.
+\\\\
+Since our initial submission, we have thoroughly edited the manuscript and addressed the comments provided by reviewers.
+We have added practical discussion on the use of UMAP, caveats to its uses and interpretation , and follow-up on findings suggesting differences in the distribution of FEV1 levels among East Asian individuals in the UK biobank.
+\\\\
+We believe that this is compelling evidence that UMAP will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
 \\

### 2019-07-08 ms/main.tex
small edits

@@
 Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed.
-Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference, novembre2016recent, spence2018inference}.\adpcomment{added two citations}
+Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference}] \sgcomment{Could we put the original PCA paper and structure/admixture, chromopainter?}.
 
@@
 
-The amount of information encoded in the lower-variance principal components increases with sample size, so researchers typically examine multiple two-dimensional projections to explore data.
+The amount of information encoded in the highest-variance PCs increases slowly with sample size, so researchers typically examine multiple two-dimensional projections to lower-variance PCs to explore data.
 In this process, finer features of the data may be hidden by the projections or hard to interpret.
@@
 A popular method for such visualization is t-distributed stochastic neighbour embedding (t-SNE)\cite{maaten2008visualizing}.
-t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}; using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}.
+t-SNE has been used before to visualize SNPs\cite{platzer2013visualization}.
+Using data from the 1000 Genomes Project (1KGP)\cite{10002015global}, it groups individuals corresponding roughly to their continent of origin, with smaller ethnic sub-groups visible within the larger continental clusters\cite{li2017tsne}.
 However, t-SNE struggles with very large datasets, when a large number of locally optimal configurations make convergence to a globally satisfying solution difficult.
@@
 The end result is a patchwork of low-dimensional representations of neighbourhoods that groups genetically similar individuals together on a local scale while better preserving long-range topological connections to more distantly related individuals.
-The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.
+The method has been successfully applied to single-cell RNA sequencing datasets \cite{umap2018singlecell}.
 

### 2019-07-08 ms/main.tex
suggested edit to FEV1

@@
 This clustering effect is robust to the number of components considered (figure~\ref{fig:supp_megamontage_pc10_50}).
-Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}.
-Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym.
+Differentiation in the 1KGP Gujarati population has been previously identified through a PCA restricted to the Gujarati \cite{reich2009india}.
+Following a preprint version of the present article, 23andme released a statement \cite{23andme} arguing that one of the two clusters could be traced, via 23andMe participant recontact to individuals from a group in Western India with shared ancestry and patronym.
 [D.
@@
 \subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
-The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds.
-Figure~\ref{fig:fig_ukbb} compares UMAP to PCA apply to UKBB.
+The UKBB contains data on 488,377 individuals including genotypes, phenotypic measures and self-identified ethnic backgrounds.
+Figure~\ref{fig:fig_ukbb} compares UMAP to PCA applied to UKBB.
 As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture.
@@
 Baseline FEV1 levels are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}.
-We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population.
+To focus on individuals of Asian ancestry (rather than, e.g., individuals born in Japan but who have European ancestry), we first selected the individuals whose UMAP coordinates were near the Chinese cluster.
+We then focused on individuals born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population.
 These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}.

### 2019-07-08 ms/main.tex

@@
 
-Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
+In clinical settings, baseline Forced Expiratory Volume in 1 second (FEV1) are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}.
+Figures~\ref{fig:supp_ukbb_fev_f} and \ref{fig:supp_ukbb_fev_m} show strong correlations with genetic clustering: certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
+
  Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes.
@@
 
-\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
+%\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
 Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations.
-Baseline FEV1 levels are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}.
 To focus on individuals of Asian ancestry (rather than, e.g., individuals born in Japan but who have European ancestry), we first selected the individuals whose UMAP coordinates were near the Chinese cluster.
@@
 These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}.
-Using this approach facilitates identifying large phenotypic differences among genetically related groups, which is useful when population labels are unavailable or correlate poorly with genetic ancestry.
 After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$).
@@
 Using a minimal number of neighbours (e.g.
-5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact and losing visual detail, as well as possibly becoming disjoint.
-We sometimes found that a lower number of PCs led to a simpler shape that facilitated discussion (e.g.
+5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail.
+However, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally performs well across datasets.
+
+In the absence of clear theoretical rationales, we suggest to use as many PCs as are available and computationally feasible, even though we sometimes found that a lower number of PCs led to a simpler shape that facilitated discussion (e.g.
 figure \ref{fig:fig_ukbb}).
-In the absence of clear theoretical rationales, we suggest to use as many PCs as are available and computationally feasible, and beginning with a minimum distance of 0.1 to 0.5 and the default number of neighbours (15).
-We recommend reporting on a range of parameter values and following up on observations with statistical testing.
+Overall, we recommend reporting on a range of parameter values and following up on observations with statistical testing.
 
@@
 We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.
-Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.
+Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.
 
@@
     \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
-The OEG cluster next to the Chinese cluster is coloured differently, suggesting this population may have different FEV1 characteristics.
-A cluster of OEG/other white individuals is more blue, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
+The OEG cluster next to the Chinese cluster appears redder on the middle panel despite similar ages (right panel), suggesting higher FEV1 in that population.
+A cluster of OEG/other white individuals is bluer relative to the rest of the British or white population, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
     \label{fig:supp_comparison_fev_chi_eur}

### 2019-07-08 ms/cover_letter.tex
more citations to intro, edit caveats a bit, rm arrows from images

@@
 We are pleased to resubmit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics.
-Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are the new frontier of genomics research.
-In contrast to existing methods, this approach is able to represent both discrete and continuous population structure and the distributions of ancestries, admixture, and phenotypes within the context of such structure.
+Since our previous submission we have thoroughly edited the manuscript to address reviewer comments in detail, to provide guidance on usage and interpretation of UMAP, and to provide rigorous follow up on findings.
 \\\\
-By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts.
-We identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region --- likely descendants of the Hispano population --- and  multiple associations between genetics, geography, and phenotypes in the UK biobank as well as population structure within 1000 Genomes Project populations.
+Among the findings are differences between East Asian populations in the distributions of FEV1, a clinically important measurement used in studies of lung function, which have not been identified to date to the best of our knowledge.
+We have also identified more detailed population structure patterns using the country of birth for individuals in the UK biobank and demonstrated how to use UMAP to identify patterns in ancestry on a global scale.
 \\\\
-We show that UMAP outperforms t-SNE --- the current state-of-the-art --- computationally, qualitatively, and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
-Methods described in our preprint have already been used to study ancestral diversity in the Genome Aggregation Database (gnomAD)\footnote{Karczewski, Konrad J., et al.
-"Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes." BioRxiv (2019): 531210.} and the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
-"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and by 23andMe.
-Indeed, 23andMe was able to follow up on one of the population structure signals we identified in our preprint by recontacting their participants, and the resulting interpretation of the population structure in the Gujarati is now part of the current manuscript.
-\\\\
-Since our initial submission, we have thoroughly edited the manuscript and addressed the comments provided by reviewers.
-We have added practical discussion on the use of UMAP, caveats to its uses and interpretation , and follow-up on findings suggesting differences in the distribution of FEV1 levels among East Asian individuals in the UK biobank.
-\\\\
-We believe that this is compelling evidence that UMAP will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
+The methods described in our preprint continue to be used to study ancestral diversity, such as in the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
+"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.} and by 23andMe\footnote{23andMe.
+https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/.
+``23andMe Tests New Ancestry Breakdown in Central and South Asia'' (2019-04-03)}, whose interpretation of the population structure in the Gujarati is now part of the current manuscript.
 \\

### 2019-07-08 ms/main.tex
more citations to intro, edit caveats a bit, rm arrows from images

@@
 Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed.
-Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference}] \sgcomment{Could we put the original PCA paper and structure/admixture, chromopainter?}.
+Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference, eigen2006, Hellenthal747}].
 
@@
 \subsubsection*{Caveats}
-The existence of adjustable parameter, including PC cutoff, minimum distance, and number of neighbors, opens the door to tweaking.
-Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices.
+The existence of adjustable parameters, including PC cutoff, minimum distance, and number of neighbours, opens the door to tweaking.
 Using a minimal number of neighbours (e.g.
 5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail.
-However, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally performs well across datasets.
+We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally perform well across datasets.
 
@@
 It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data.
+Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in figure~\ref{fig:supp_megamontage_pc10_50}.
 For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.pdf}
+    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.png}
     \end{subfigure}
     \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
-The OEG cluster next to the Chinese cluster appears redder on the middle panel despite similar ages (right panel), suggesting higher FEV1 in that population.
-A cluster of OEG/other white individuals is bluer relative to the rest of the British or white population, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
+The OEG cluster next to the Chinese cluster appears redder on the middle panel, suggesting higher levels of FEV1.}
     \label{fig:supp_comparison_fev_chi_eur}

### 2019-07-08 ms/cover_letter.tex
minor edits, added gnomAD

@@
 \\\\
-Among the findings are differences between East Asian populations in the distributions of FEV1, a clinically important measurement used in studies of lung function, which have not been identified to date to the best of our knowledge.
+Among the findings that are now discussed in more detail, we found differences between East Asian populations in the distributions of FEV1, a clinically important measurement of lung function, which have not been identified to date to the best of our knowledge.
 We have also identified more detailed population structure patterns using the country of birth for individuals in the UK biobank and demonstrated how to use UMAP to identify patterns in ancestry on a global scale.
 \\\\
-The methods described in our preprint continue to be used to study ancestral diversity, such as in the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
-"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.} and by 23andMe\footnote{23andMe.
+The methods described in our preprint continue to be used by leading groups including the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
+"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, GnomAD \footnote{}, and 23andMe\footnote{23andMe.
 https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/.
-``23andMe Tests New Ancestry Breakdown in Central and South Asia'' (2019-04-03)}, whose interpretation of the population structure in the Gujarati is now part of the current manuscript.
+``23andMe Tests New Ancestry Breakdown in Central and South Asia'' (2019-04-03)}.
+In fact, a 23andme follow-up on a result from this manuscript provided a definite explanation for a pattern of population structure among the Gujarati; this follow-up observation is now part of the manuscript.
+
 \\

### 2019-07-08 ms/main.tex
add colour legends

@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
+    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.jpeg}
     \end{subfigure}
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
+    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
     \end{subfigure}
@@
 \centering
-   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
+   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.jpeg}
    \caption{UMAP on the first 10 principal components of HRS data.

### 2019-07-09 ms/cover_letter.tex
c/e and citation

@@
 The methods described in our preprint continue to be used by leading groups including the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
-"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, GnomAD \footnote{}, and 23andMe\footnote{23andMe.
+``Population histories of the United States revealed through fine-scale migration and haplotype analysis.'' BioRxiv (2019): 577411.}, gnomAD\footnote{Karczewski, Konrad J., et al.
+``Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes.'' BioRxiv (2019): 531210.}, and 23andMe\footnote{23andMe.
 https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/.
 ``23andMe Tests New Ancestry Breakdown in Central and South Asia'' (2019-04-03)}.
-In fact, a 23andme follow-up on a result from this manuscript provided a definite explanation for a pattern of population structure among the Gujarati; this follow-up observation is now part of the manuscript.
-
-\\
-\\
+In fact, a 23andMe follow-up on a result from this manuscript provided a definite explanation for a pattern of population structure among the Gujarati; this follow-up observation is now part of the manuscript.
+\\\\
 Sincerely,

### 2019-07-09 ms/main.tex
figure legends

@@
 We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the datasets.
-We also thank Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access; David Poznik, Liz Babalola, and Adam Auton from 23andMe for discussing findings in the 1KGP; and Selin Jessa for introducing us to UMAP.
+We also thank Audrey Grant, Ryan Hernandez, Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, Melissa Spear, and Dara Torgerson for useful discussions about science, programming, and data access; David Poznik, Liz Babalola, and Adam Auton from 23andMe for discussing findings in the 1KGP; and Selin Jessa for introducing us to UMAP.
 This research has been conducted using the UK Biobank Resource under Application Number 6728.
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.75\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_born.pdf}
+    \includegraphics[width=0.9\textwidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_born.jpeg}
     \end{subfigure}

### 2019-07-09 ms/cover_letter.tex
revert cover letter

--- 2019-07-08 ms/cover_letter.tex
@@
 We are pleased to resubmit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics.
-Since our previous submission we have thoroughly edited the manuscript to address reviewer comments in detail, to provide guidance on usage and interpretation of UMAP, and to provide rigorous follow up on findings.
+Since our previous submission we have thoroughly edited the manuscript to address reviewer and preprint reader comments in detail, to provide guidance on usage and interpretation of UMAP, and to provide rigorous follow up on some findings.
 \\\\
@@
 The methods described in our preprint continue to be used by leading groups including the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al.
-"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, GnomAD \footnote{}, and 23andMe\footnote{23andMe.
+"Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411}, GnomAD \footnote{Karczewski, Konrad J., et al., BioRxiv (2019): 531210 }, and 23andMe\footnote{23andMe.
 https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/.
 ``23andMe Tests New Ancestry Breakdown in Central and South Asia'' (2019-04-03)}.
-In fact, a 23andme follow-up on a result from this manuscript provided a definite explanation for a pattern of population structure among the Gujarati; this follow-up observation is now part of the manuscript.
-
 \\
+We therefore believe that the manuscript will be of interest to the wide readership of PLoS Genetics and hope that you will find this revision suitable for publication.
 \\

### 2019-07-13 ms/main.tex
small edit

@@
 UMAP, in particular, reveals a wide range of features that would not be apparent using linear maps.
-Given its ease of use, breadth of results, and low computational cost, we propose that UMAP should become a default companion to PCA in large genomic cohorts.
+Given its ease of use, broad applicability, and low computational cost, we propose that UMAP should become a default companion to PCA in large genomic cohorts.
 

### 2019-07-15 ms/main.tex
c/e for ms submission

@@
 \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+\subsection*{Author summary}
+Population structure is a phenomenon where there are systematic differences in allele frequencies between populations, resulting in some populations sharing more DNA than expected.
+It is the result of non-random mating over generations, usually from geographic isolation, founder effects from migrations, sociocultural factors, or a combination of these and other factors.
+This structure informs us of human history, and can have a significant impact on studies of medical genetics.
+Visualization is an effective approach to study population structure.
+We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction.
+We show that using UMAP works especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries.
+The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions.
+
+\subsection*{Introduction}
 Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies.

### 2019-07-15 ms/main.tex
author summary

--- 2019-07-13 ms/main.tex
@@
 \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+\subsection*{Author summary}
+The demographic history of human populations features evolving geographic and social barriers to mating.
+Over time, these barriers have led to varying levels of genetic relatedness among individuals.
+This \emph{population structure} is informative about human history, and can have a significant impact on studies of medical genetics.
+Because population structure depends on myriad demographic forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure.
+We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction.
+Using UMAP is computationally efficient and can identify fine-scale population structure in large population datasets.
+We find it particularly useful to reveal phenotypic variation among genetically related populations, and argue that it is a useful complement  to principal component analysis in primary data visualization.
+
+\subsection*{Introduction}
 Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies.

### 2019-07-15 ms/main.tex
rewrite author summary

@@
 \subsection*{Author summary}
-The demographic history of human populations features evolving geographic and social barriers to mating.
+How about: The demographic history of human populations features evolving geographic and social barriers to mating.
 Over time, these barriers have led to varying levels of genetic relatedness among individuals.
@@
 Using UMAP is computationally efficient and can identify fine-scale population structure in large population datasets.
-We find it particularly useful to reveal phenotypic variation among genetically related populations, and argue that it is a useful complement  to principal component analysis in primary data visualization.
+We find it particularly useful to reveal phenotypic variation among genetically related populations, and recommend it is a complement  to principal component analysis in primary data visualization.
 

### 2019-08-16 HRS_plotting.Rmd
manuscript changes, add some R scripts used in analysis and dataviz

---
title: "HRS plotting"
author: 
  - name: Alex Diaz-Papkovich
affiliation:
  - &cruk Quantitative Life Sciences, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
  df_print: paged
code_folding: show
toc: yes
toc_float: 
  collapsed: false
smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## HRS admixture plots

teeeeeext

## Including Plots

```{r}
library(tidyverse)
library(ggplot2)
library(Ternary)
library(dplyr)
```

# Import the HRS auxiliary data and projection to use.
```{r}
# Auxiliary data
hrs_aux_data <- read.csv("/Volumes/Stockage/alex/hrs/aux/HRS_aux.csv")

# Projection
fname <- "HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD0.5_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116"
proj_dir <- "/Volumes/Stockage/alex/gt-dimred/np1/projections"

hrs_1kgp_proj <- read.table(paste(proj_dir, fname, sep="/"))

# This is HRS+1KGP data; first 12,454 are HRS
hrs_proj <- hrs_1kgp_proj[1:12454,]

# Merge the tables
hrs_full_data <- cbind(hrs_aux_data, hrs_proj)

# RGB colours from admixture
hrs_full_data$rgb_vals <- rgb(hrs_full_data$ADMIX1*255,
                              hrs_full_data$ADMIX2*255,
                              hrs_full_data$ADMIX3*255,
                              maxColorValue = 255)

```

```{r}
# legends
require(RColorBrewer)
#pcol=brewer.pal(9, "Set1")

AncCol=c("#FF0000","#00FF00","#0000FF")

AncRGB=col2rgb(AncCol)/255
#pdf("UMAP_ternary_plot_legend.pdf",height=2.375,width=2.375,pointsize=10)
par(mar=c(2.1,2.1,0.1,0.1))
TernaryPlot(alab=expression(''%->% "African ancestry (%)" %->% ''),blab=expression(''%->%"European ancestry (%)" %->% ''),clab=expression(''%<-% "Asian/Native American ancestry (%)" %<-% ''),lab.cex=1,axis.cex = 0.8)
dx=0.01;
cnt=0
bk=0;
for(xi in seq(0,1,length=round(1/dx))){
  for(yi in seq(0,1,length=round(1/dx))){
    for(zi in seq(0,1,length=round(1/dx))){
      if(abs(1-(xi+yi+zi)) < 1e-8){
        ired = xi*AncRGB[1,1] + yi*AncRGB[1,2] + zi*AncRGB[1,3]
        igreen = xi*AncRGB[2,1] + yi*AncRGB[2,2] + zi*AncRGB[2,3]
        iblue = xi*AncRGB[3,1] + yi*AncRGB[3,2] + zi*AncRGB[3,3]
        AddToTernary(points,c(xi,yi,zi),col=rgb(ired,igreen,iblue,1),pch=20)
        cnt=cnt+1
      }
    }
  }
}

# legends
require(RColorBrewer)
#pcol=brewer.pal(9, "Set1")

AncCol=c("#FF0000","#0000FF","#00FF00")

AncRGB=col2rgb(AncCol)/255
#pdf("UMAP_ternary_plot_legend.pdf",height=2.375,width=2.375,pointsize=10)
par(mar=c(2.1,2.1,0.1,0.1))
TernaryPlot(alab=expression(''%->% "African ancestry (%)" %->% ''),blab=expression(''%->%"European ancestry (%)" %->% ''),clab=expression(''%<-% "Asian/Native American ancestry (%)" %<-% ''),lab.cex=1,axis.cex = 0.8)
dx=0.01;
cnt=0
bk=0;
for(xi in seq(0,1,length=round(1/dx))){
  for(yi in seq(0,1,length=round(1/dx))){
    for(zi in seq(0,1,length=round(1/dx))){
      if(abs(1-(xi+yi+zi)) < 1e-8){
        ired = xi*AncRGB[1,1] + yi*AncRGB[1,2] + zi*AncRGB[1,3]
        igreen = xi*AncRGB[2,1] + yi*AncRGB[2,2] + zi*AncRGB[2,3]
        iblue = xi*AncRGB[3,1] + yi*AncRGB[3,2] + zi*AncRGB[3,3]
        AddToTernary(points,c(xi,yi,zi),col=rgb(ired,igreen,iblue,1),pch=20)
        cnt=cnt+1
      }
    }
  }
}
```

```{r}
out_dir <- "/Volumes/Stockage/alex/hrs_1000G/images/for_export/alt_colours"
fname <- "HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_birth"

p_hrs_birth <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=BirthRegionName)) +
  geom_point(alpha=0.6, size=0.1) +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position="bottom",
        legend.title=element_blank()) + 
  guides(colour = guide_legend(override.aes = list(size=4, alpha = 1)))

p_hrs_birth

ggsave(paste(out_dir, "/", fname, ".jpeg", sep=""), height=h, width=w, units=u)
```

```{r}
p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX1*255,
                                                       hrs_full_data$ADMIX2*255,
                                                       hrs_full_data$ADMIX3*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs
```

# Alternate colourings
```{r}

out_dir <- "/Volumes/Stockage/alex/hrs_1000G/images/for_export/alt_colours"
fname <- "HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix"

# dimensions
h <- 120.4
w <- 186.0

# units
u <- "mm"

p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX1*255,
                                                       hrs_full_data$ADMIX2*255,
                                                       hrs_full_data$ADMIX3*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "123.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX1*255,
                                                       hrs_full_data$ADMIX3*255,
                                                       hrs_full_data$ADMIX2*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "132.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX2*255,
                                                       hrs_full_data$ADMIX1*255,
                                                       hrs_full_data$ADMIX3*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "213.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX2*255,
                                                       hrs_full_data$ADMIX3*255,
                                                       hrs_full_data$ADMIX1*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "231.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX3*255,
                                                       hrs_full_data$ADMIX1*255,
                                                       hrs_full_data$ADMIX2*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "312.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_full_data, aes(x=V1, y=V2, col=rgb(hrs_full_data$ADMIX3*255,
                                                       hrs_full_data$ADMIX2*255,
                                                       hrs_full_data$ADMIX1*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=0.1) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "321.jpeg", sep=""), height=h, width=w, units=u)
```

# Same, but for Hispanic data
```{r}

# Auxiliary data
#hrs_aux_data <- read.csv("/Volumes/Stockage/alex/hrs/aux/HRS_aux.csv")

# Projection
fname <- "HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD0.5_pca_hrshisp_added1kgp_2018115153245"
proj_dir <- "/Volumes/Stockage/alex/gt-dimred/np1/projections"

hrs_hisp_1kgp_proj <- read.table(paste(proj_dir, fname, sep="/"))

# Only the first 12,454 are HRS.
# Note that we did the projection with all the HRS (so we can also project the other HRS onto Hisps)
hrs_hisp_1kgp_proj <- hrs_hisp_1kgp_proj[1:12454,]

# Merge the tables
hrs_hisp_data <- cbind(hrs_aux_data, hrs_hisp_1kgp_proj)

# Reduce to just Hispanic individuals
hrs_hisp_data <- subset(hrs_hisp_data, HispanicStatus=="Hispanic")

# RGB colours from admixture
hrs_hisp_data$rgb_vals <- rgb(hrs_hisp_data$ADMIX1*255,
                              hrs_hisp_data$ADMIX2*255,
                              hrs_hisp_data$ADMIX3*255,
                              maxColorValue = 255)

```

# Create beautiful plots!
```{r}
out_dir <- "/Volumes/Stockage/alex/hrs_1000G/images/for_export/alt_colours"
fname <- "HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix"

# dimensions
h <- 154.6
w <- 169.2

# units
u <- "mm"

# point size
s <- 0.5

p_hrs <- ggplot(hrs_hisp_data, aes(x=V1, y=V2, col=rgb(hrs_hisp_data$ADMIX1*255,
                                                       hrs_hisp_data$ADMIX2*255,
                                                       hrs_hisp_data$ADMIX3*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=s) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "123_hisp.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_hisp_data, aes(x=V1, y=V2, col=rgb(hrs_hisp_data$ADMIX1*255,
                                                       hrs_hisp_data$ADMIX3*255,
                                                       hrs_hisp_data$ADMIX2*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=s) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "132_hisp.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_hisp_data, aes(x=V1, y=V2, col=rgb(hrs_hisp_data$ADMIX2*255,
                                                       hrs_hisp_data$ADMIX1*255,
                                                       hrs_hisp_data$ADMIX3*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=s) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "213_hisp.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_hisp_data, aes(x=V1, y=V2, col=rgb(hrs_hisp_data$ADMIX2*255,
                                                       hrs_hisp_data$ADMIX3*255,
                                                       hrs_hisp_data$ADMIX1*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=s) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "231_hisp.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_hisp_data, aes(x=V1, y=V2, col=rgb(hrs_hisp_data$ADMIX3*255,
                                                       hrs_hisp_data$ADMIX1*255,
                                                       hrs_hisp_data$ADMIX2*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=s) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "312_hisp.jpeg", sep=""), height=h, width=w, units=u)

p_hrs <- ggplot(hrs_hisp_data, aes(x=V1, y=V2, col=rgb(hrs_hisp_data$ADMIX3*255,
                                                       hrs_hisp_data$ADMIX2*255,
                                                       hrs_hisp_data$ADMIX1*255,
                                                       maxColorValue = 255))) +
  geom_point(alpha=0.6, size=s) +
  scale_color_identity() +
  theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p_hrs

ggsave(paste(out_dir, "/", fname, "321_hisp.jpeg", sep=""), height=h, width=w, units=u)
```

```{r}
library('Ternary')
require(RColorBrewer)
#pcol=brewer.pal(9, "Set1")

AncCol=c("#4DAF4A","#E41A1C","#377EB8")

AncRGB=col2rgb(AncCol)/255
#pdf("UMAP_ternary_plot_legend.pdf",height=2.375,width=2.375,pointsize=10)
par(mar=c(2.1,2.1,0.1,0.1))
TernaryPlot(alab=expression(''%->% "African ancestry (%)" %->% ''),blab=expression(''%->%"European ancestry (%)" %->% ''),clab=expression(''%<-% "Native American ancestry (%)" %<-% ''),lab.cex=0.6,axis.cex = 0.5)
dx=0.01;
cnt=0
bk=0;
for(xi in seq(0,1,length=round(1/dx))){
  for(yi in seq(0,1,length=round(1/dx))){
    for(zi in seq(0,1,length=round(1/dx))){
      if(abs(1-(xi+yi+zi)) < 1e-8){
        ired = xi*AncRGB[1,1] + yi*AncRGB[1,2] + zi*AncRGB[1,3]
        igreen = xi*AncRGB[2,1] + yi*AncRGB[2,2] + zi*AncRGB[2,3]
        iblue = xi*AncRGB[3,1] + yi*AncRGB[3,2] + zi*AncRGB[3,3]
        AddToTernary(points,c(xi,yi,zi),col=rgb(ired,igreen,iblue,1),pch=20)
        cnt=cnt+1
      }
    }
  }
}
```

```{r}
tern_data <- select(hrs_full_data,ADMIX1*255,ADMIX2*255,ADMIX3*255)

p_tern <- ggtern() +
  geom_point(data=data.frame(x=255,y=255,z=255), aes(x,y,z))

p_tern
```

### 2019-08-16 UKBB_plotting.Rmd
manuscript changes, add some R scripts used in analysis and dataviz

---
title: "London alpha plotting"
author: 
  - name: Alex Diaz-Papkovich
affiliation:
  - &cruk Quantitative Life Sciences, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
  df_print: paged
code_folding: show
toc: yes
toc_float: 
  collapsed: false
smooth_scroll: false
---
## Plotting UKBB stuff
Python's plotting is getting on my nerves. I want to make a scatterplot where I specify each point's colour and alpha values.

# Import libraries
```{r}
library(ggplot2)
library(tidyverse)
```

# Import the data
```{r}
data_dir <- "/Volumes/Stockage/alex/ukbb_other/data/london_alphas.csv"

ukbb_plotting_data <- read.csv(data_dir)
ukbb_plotting_data <- ukbb_plotting_data[!(ukbb_plotting_data$eth_txt %in% c('British', 'Irish', 'Any other white background')),]

out_dir <- "/Volumes/Stockage/alex/ukbb_images/london_alpha"
fname <- "UKBB_UMAP_PC10_NN15_MD05_2018328174511_within50km"
 h <- 15
 w <- 15

p_scatter <- ggplot(ukbb_plotting_data, aes(x=x_coords, y=y_coords)) +
  geom_point(colour=ukbb_plotting_data$colour, alpha=ukbb_plotting_data$London_alpha, stroke=0) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank())
p_scatter

ggsave(paste(out_dir, paste(fname,".jpeg",sep=""), sep="/"), device = "jpeg", height = h, width = w, units = "in")
```

# What about those outside of 50k from London?
```{r}
data_dir <- "/Volumes/Stockage/alex/ukbb_other/data/london_alphas.csv"

ukbb_plotting_data <- read.csv(data_dir)
ukbb_plotting_data <- ukbb_plotting_data[!(ukbb_plotting_data$eth_txt %in% c('British', 'Irish', 'Any other white background')),]

ukbb_plotting_data$Outside_London <- ifelse(ukbb_plotting_data$London_alpha==0,1,0)

out_dir <- "/Volumes/Stockage/alex/ukbb_images/london_alpha"
fname <- "UKBB_UMAP_PC10_NN15_MD05_2018328174511_outside_50km"
 h <- 15
 w <- 15

p_scatter <- ggplot(ukbb_plotting_data, aes(x=x_coords, y=y_coords)) +
  geom_point(colour=ukbb_plotting_data$colour, alpha=ukbb_plotting_data$Outside_London, stroke=0) +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank())
p_scatter

ggsave(paste(out_dir, paste(fname,".jpeg",sep=""), sep="/"), device = "jpeg", height = h, width = w, units = "in")
```

### 2019-08-16 ms/main.tex
manuscript changes, add some R scripts used in analysis and dataviz

@@
 Here we apply uniform manifold approximation and projection (UMAP), a non-linear dimension reduction tool, to three well-studied genotype datasets and discover overlooked subpopulations within the American Hispanic population, fine-scale relationships between geography, genotypes, and phenotypes in the UK population, and cryptic structure in the Thousand Genomes Project data.
-This approach outperforms other visualization methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
+This approach is well-suited to the influx of large and diverse data and opens new lines of inquiry in population-scale datasets.
+%This approach outperforms other visualization methods, is well-suited to the influx of large and diverse data, and opens new lines of inquiry in population-scale datasets.
 \end{abstract}
@@
 \subsection*{Author summary}
-How about: The demographic history of human populations features evolving geographic and social barriers to mating.
+The demographic history of human populations features varying geographic and social barriers to mating.
 Over time, these barriers have led to varying levels of genetic relatedness among individuals.
 This \emph{population structure} is informative about human history, and can have a significant impact on studies of medical genetics.
-Because population structure depends on myriad demographic forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure.
+Because population structure depends on myriad demographic and ecological forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure.
 We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction.
@@
 Here we will focus on nonparametric approaches to visualize relatedness patterns among individuals within populations.
-If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of copies of the derived alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions.
+If we consider unphased single nucleotide polymorphism (SNP) data, an individual genome can be represented as a sequence of integers corresponding to the number of copies of the alleles carried by the individual at each of the $L$ SNPs for which genotypes are available, with $L$ ranging from hundreds of thousands to hundreds of millions.
 Since each individual is represented as an $L$-dimensional vector, dimension reduction methods are needed to visualize the data.
@@
 Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together.
-Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}.
+Results using t-SNE on principal components are presented in fig~\ref{fig:supp_megamontage_pc2_9}.
 Axes in UMAP and t-SNE are arbitrary.
@@
     TSI, Tuscani in Italy.
-Figures~\ref{fig:supp_umap_hrs_hisp_admix} and \ref{fig:supp_umap_hrs_hisp_birth} present the same projection of individuals from the HRS coloured by estimated admixture proportions census region of birth, respectively.
+Figs~\ref{fig:supp_umap_hrs_hisp_admix} and \ref{fig:supp_umap_hrs_hisp_birth} present the same projection of individuals from the HRS coloured by estimated admixture proportions census region of birth, respectively.
 \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters.
 Indian individuals born in Kenya (in purple) form one such cluster.
-A version coloured by self-identified ethnicity is presented in figure~\ref{fig:supp_umap_ukbb_asian_eth}.
+A version coloured by self-identified ethnicity is presented in fig~\ref{fig:supp_umap_ukbb_asian_eth}.
 }
@@
 \subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}.
-Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing.
+Fig~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing.
 Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters.
 Using either method on the top principal components leads to distinct population clusters and less defined continental structure.
-Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
+Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figs~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}).
 To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs).
-Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.
-
-Focusing on UMAP with the leading 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries.
+Fig~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.
+
+Focusing on UMAP with the leading 15 principal components (fig~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries.
 British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry.
@@
 Within population clusters, family members were projected near each other within broader population groups.
-When UMAP was parameterized to use only 5 nearest neighbours, however, families often formed distinct clusters (figure~\ref{fig:supp_1kgp_families}): Using few neighbors to build a manifold emphasizes closer relatedness.
+When UMAP was parameterized to use only 5 nearest neighbours, however, families often formed distinct clusters (fig~\ref{fig:supp_1kgp_families}): Using few neighbors to build a manifold emphasizes closer relatedness.
 
@@
 Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan.
-This clustering effect is robust to the number of components considered (figure~\ref{fig:supp_megamontage_pc10_50}).
+This clustering effect is robust to the number of components considered (fig~\ref{fig:supp_megamontage_pc10_50}).
 Differentiation in the 1KGP Gujarati population has been previously identified through a PCA restricted to the Gujarati \cite{reich2009india}.
@@
 
-The HRS forms several large clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}).
+The HRS forms several large clusters, reflecting both ethnicity (fig~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (fig~\ref{fig:umap_hrs_admix}).
 Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics.
-The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}).
-By generating the PC axes and UMAP embedding for the HRS data in figure~\ref{fig:supp_umap_hrs_eth}, and projecting the 1KGP data onto it, we reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}).
+The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (fig~\ref{fig:supp_hrs_born}).
+By generating the PC axes and UMAP embedding for the HRS data in fig~\ref{fig:supp_umap_hrs_eth}, and projecting the 1KGP data onto it, we reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (fig~\ref{fig:supp_hrs_1kgp_projected}).
 One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.
 
-\subsubsection*{Regional patterns in the Hispanic subpopulation} Applying UMAP to self-identified Hispanic individuals in the HRS reveals clear groupings related to birth region.
-One cluster, highlighted in figure~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}, consists almost entirely of individuals born in the Mountain Region of the United States.
-This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in figure~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6.
-Even though continental admixture patterns do correlate with UMAP position (figure~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster.
+\subsubsection*{Regional patterns in the Hispanic subpopulation} Applying UMAP to self-identified Hispanic individuals in the HRS reveals clear groupings related to birth region in fig~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp}.
+The highlighted cluster consists almost entirely of individuals born in the Mountain Region of the United States.
+This cluster is not apparent when looking at a grid of pairwise plots of the first 8 principal components, provided in fig~\ref{fig:supp_hrs_hisp_grid}, as the signal is distributed along PCs 3, 4, and 6.
+Even though continental admixture patterns do correlate with UMAP position (fig~\ref{fig:supp_umap_hrs_hisp_admix}), these do not explain the Mountain Region cluster.
 Individuals from 1KGP populations do not appear in the cluster when projected to the UMAP embedding.
 The cluster possibly comprises the Hispano/Nuevomexicano population of the Southwest US, who have been present in the Mountain Region area long before the more recent immigrants from Latin America, and whose ancestry is expected to reflect both distinct Native ancestry and population-specific drift relative to other Hispanic populations.
-Such a cluster has been previously identified in AncestryDNA data using sequential clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.
+Such a cluster has been previously identified in AncestryDNA data using network-based clustering on identity-by-descent connections\cite{han2017clustering}; a recent preprint discusses the Mountain Region Hispanics with a more detailed historical description\cite{Jordan333609}.
 
@@
 The UKBB contains data on 488,377 individuals including genotypes, phenotypic measures and self-identified ethnic backgrounds.
-Figure~\ref{fig:fig_ukbb} compares UMAP to PCA applied to UKBB.
-As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture.
-UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations.
-The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.
+Fig~\ref{fig:fig_ukbb} compares UMAP to PCA applied to UKBB.
+As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer structure.
+UMAP on the top 10 principal components reveals continuous and discrete population structure (fig~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations.
+%The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.
+The result is a succinct illustration of the complex structure and population relationships in a large and multi-ethnic dataset.
 
@@
 
-The largest cluster in figure~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations.
+The largest cluster in fig~\ref{fig:umap_ukbb_pc10} consists of the White British and Irish populations.
 The Irish population forms a sub-cluster, but many individuals are also scattered throughout the British-identifying population.
@@
 Many claim mixed ancestry, and there are clusters of individuals who belong to an "other ethnic group".
-Using data on countries of birth, we identified many finer groups in figure~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino clusters being projected near Chinese clusters.
-
-Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
+Using data on countries of birth, we identified many finer groups in fig~\ref{fig:supp_ukbb_cob}, and confirmed they appeared in intuitive areas with, e.g., Japanese and Filipino clusters being projected near Chinese clusters.
+
+Fig~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from fig~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly.
 Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe and British-wide data\cite{novembre2008europe,leslie2015fine}.
-Figure~\ref{fig:umap_ukbb_pc10} shows that the admixed individuals have UMAP coordinates next to White British individuals residing in the South East of the UK, where London is located (see also figure~\ref{fig:supp_london_distance}, where individuals are coloured by distance from London).
+Fig~\ref{fig:umap_ukbb_pc10} shows that the admixed individuals have UMAP coordinates next to White British individuals residing in the South East of the UK, where London is located (see also fig~\ref{fig:supp_london_distance}, where individuals are coloured by distance from London).
 This likely reflects the high migration levels to the city and surrounding area: the UMAP projection attempted to preserve both the genetic similarity among admixed individuals and the relatedness with White British individuals in cosmopolitan areas.
@@
  The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved.
-Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.
+Figs~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.
 
@@
 Adding more components creates finer clusters.
-(figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}).
+(figs~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}).
 Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km.
@@
 
-As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}).
+As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (fig~\ref{fig:umap_ukbb_geo_map}).
 The geographical patterns outside major urban centers are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.
@@
 Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.
-Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
+Zoomed maps of East Asia, the Caribbean, and Europe are available in figs~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
 \end{figure}
@@
 Similarly to UMAP, t-SNE applied to the UKBB data both displays diversity within the "White British" population and identifies clusters among other groups.
-However, it has three drawbacks: it is much slower, requiring 2.26 hours for its first thousand iterations alone on 10 principal components against UMAP's 14 minutes; it fails to find a global optimum, which results in a scattering of individuals and groups that are not stable across independent runs; and it does not identify continuity between different continental groups resulting from admixture (figure~\ref{fig:supp_ukbb_tsne}).
+However, it has three drawbacks: it is much slower, requiring 2.26 hours for its first thousand iterations alone on 10 principal components against UMAP's 14 minutes; it fails to find a global optimum, which results in a scattering of individuals and groups that are not stable across independent runs; and it does not identify continuity between different continental groups resulting from admixture (fig~\ref{fig:supp_ukbb_tsne}).
 
@@
 
-Covariates such as height and leukocyte count (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure.
-Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and others have noticeably lower-than-average heights.
-More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals.
-To quantify and statistically test these qualitative observations, we performed an unpaired two sample t-test of self-identified White Irish and White British individuals and found British males taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
+Covariates such as height and leukocyte count (fig~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures (figs~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure.
+Several populations in fig~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and others have noticeably lower-than-average heights.
+More subtle patterns are also visible: the area of the projection in figs~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals.
+To quantify and statistically test these qualitative observations, we performed an unpaired two sample t-test of self-identified White Irish and White British individuals and found British males taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figs~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots).
 Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.
@@
 In clinical settings, baseline Forced Expiratory Volume in 1 second (FEV1) are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}.
-Figures~\ref{fig:supp_ukbb_fev_f} and \ref{fig:supp_ukbb_fev_m} show strong correlations with genetic clustering: certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
-
- Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes.
+Figs~\ref{fig:supp_ukbb_fev_f} and \ref{fig:supp_ukbb_fev_m} show strong correlations with genetic clustering: certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figs~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values).
+
+ Notably, there appears to be a juncture in the admixture population, highlighted in fig~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes.
 This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds.
-Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations (figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m}).
+Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations (figs~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m}).
 
 %\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
-Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations.
+Fig~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations.
 To focus on individuals of Asian ancestry (rather than, e.g., individuals born in Japan but who have European ancestry), we first selected the individuals whose UMAP coordinates were near the Chinese cluster.
 We then focused on individuals born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population.
-These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}.
+These four groups are mutually exclusive and are shown in fig~\ref{fig:supp_fev_test_pops}.
 After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$).
-By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
+By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in fig~\ref{fig:supp_fev_ridgeplots}).
 For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).
@@
 To assess whether the relatively poor performance of t-SNE could be due to convergence rather than a flawed model, we used UMAP to preprocess the UKBB data and provide a starting point to a standard t-SNE implementation.
-This led to representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
-Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).
+This led to representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (fig~\ref{fig:supp_tsne_umap_compare_ukbb_graph}).
+Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (fig~\ref{fig:supp_tsne_umap_compare_ukbb}).
 Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.
@@
 \subsection*{Discussion}
+By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure.
+Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences including in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
+Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.
 UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure.
@@
 It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
-By patching together many local genetic neighbourhoods, the visualizations preserve aspects of both local and global structure, unlike any existing method that we are familiar with.
 
@@
 Groupings on ethnicity formed slowly as PCs were added until reaching a stable number around 10 to 15 PCs.
-Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
+Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figs~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.
 
 \subsubsection*{Caveats}
-The existence of adjustable parameters, including PC cutoff, minimum distance, and number of neighbours, opens the door to tweaking.
+Changing the PC cutoff, minimum distance, and number of neighbours can change characteristics of the visualizations.
 Using a minimal number of neighbours (e.g.
-5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail.
+5) can result in the formation of disjoint clusters comprised of related family members (fig~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail.
 We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally perform well across datasets.
@@
 In the absence of clear theoretical rationales, we suggest to use as many PCs as are available and computationally feasible, even though we sometimes found that a lower number of PCs led to a simpler shape that facilitated discussion (e.g.
-figure \ref{fig:fig_ukbb}).
+fig~\ref{fig:fig_ukbb}).
 Overall, we recommend reporting on a range of parameter values and following up on observations with statistical testing.
@@
 It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data.
-Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in figure~\ref{fig:supp_megamontage_pc10_50}.
+Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in fig~\ref{fig:supp_megamontage_pc10_50}.
 For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
@@
 This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB.
-We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.
+We did not assign meaning to wiggles in UMAP figs, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.
 Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.
@@
 UMAP, in particular, reveals a wide range of features that would not be apparent using linear maps.
-Given its ease of use, broad applicability, and low computational cost, we propose that UMAP should become a default companion to PCA in large genomic cohorts.
+Given its ease of use, broad applicability, and low computational cost, we propose that UMAP should become a default companion to PCA and other population structure visualization and inference methods in large genomic cohorts.
 
@@
 We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
-To project 1KGP data on HRS embeddings, as in figures~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} and \ref{fig:supp_hrs_1kgp_projected}, we created the PC axes and UMAP embedding for the HRS data and then projected the 1KGP data onto them.
+To project 1KGP data on HRS embeddings, as in figs~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} and \ref{fig:supp_hrs_1kgp_projected}, we created the PC axes and UMAP embedding for the HRS data and then projected the 1KGP data onto them.
 
@@
 Both UMAP and t-SNE feature a number of adjustable parameters.
-Among the parameters that we varied, the number of PCs used in pre-processing of the data has the largest effect for both methods (see figures~\ref{fig:supp_megamontage_pc2_9} and \ref{fig:supp_megamontage_pc10_50}).
+Among the parameters that we varied, the number of PCs used in pre-processing of the data has the largest effect for both methods (see figs~\ref{fig:supp_megamontage_pc2_9} and \ref{fig:supp_megamontage_pc10_50}).
 
@@
 
-Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate.
+Colours for maps in figs~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate.
 For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country.
@@
 
-To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figures.
+To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figs.
 We rounded each attribute to an attribute-specific number of bins, and then permuted the data in the following way:
@@
 Projections coloured by participants' distance from London have random noise added (normally distributed about 0 with a standard deviation of 5km) before binning to the nearest km.
-For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}
+For each point in fig~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}
 
@@
 \subsubsection*{Availability of data and material} All data is publicly available to researchers.
+Code to carry out dimension reducation is available at \url{https://github.com/diazale/gt-dimred}.
+A demo version using freely available 1KGP data is available at \url{https://github.com/diazale/1KGP_dimred}.
 
@@
     \end{subfigure}
-    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:1000g_compare}.}
+    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in fig~\ref{fig:1000g_compare}.}
     \label{fig:supp_1kgp_3350}
@@
 When using UMAP with five neighbours, the father (in blue) is projected to the cluster of the Southern Han Chinese population while the rest of the family members (in red) form their own disjoint cluster.
-Using 15 neighbours, the family still clusters together, but as part of the Southern Han Chinese population.}
+Using 15 neighbours, the family still clusters together, but as part of the Southern Han Chinese population rather than a separate cluster.}
     \label{fig:supp_1kgp_families}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.jpeg}
-    \end{subfigure}
-    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions.
-Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
-We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
-An alternate colouring is provided in figure~\ref{fig:supp_umap_hrs_hisp_admix_alt}.}
-    \label{fig:supp_umap_hrs_hisp_admix}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
-    \end{subfigure}
-    \caption{An alternate colouring of figure~\ref{fig:supp_umap_hrs_hisp_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
-    \label{fig:supp_umap_hrs_hisp_admix_alt}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp_birthonly.jpeg}
-    \end{subfigure}
-    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured region of birth.}
-    \label{fig:supp_umap_hrs_hisp_birth}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.8\textwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256.jpeg}
-    \end{subfigure}
-    \caption{UMAP of the first 8 principal components of the Asian population in the UKBB coloured by self-identified ethnicity. This is an alternate colouring of figure~\ref{fig:umap_ukbb_indian}}
-    \label{fig:supp_umap_ukbb_asian_eth}
 \end{figure*}
@@
 colouring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population.
-This projection coloured by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.
+This projection coloured by self-identified race and Hispanic status is presented in fig~\ref{fig:supp_umap_hrs_eth}.
 Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
 We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
-An alternate colouring is provided in figure~\ref{fig:umap_hrs_admix_alt}}
+An alternate colouring is provided in fig~\ref{fig:umap_hrs_admix_alt}}
     \label{fig:umap_hrs_admix}
@@
 Populations with African ancestry (AFR) appear with Black individuals.
-East Asian (EAS) populations comprising Chinese, Kinh, and Japanese individuals cluster together with what appears in figure~\ref{fig:umap_hrs_admix} as a population of mostly Asian ancestry.
+East Asian (EAS) populations comprising Chinese, Kinh, and Japanese individuals cluster together with what appears in fig~\ref{fig:umap_hrs_admix} as a population of mostly Asian ancestry.
 South Asian (SAS) populations with Indian, Pakistani, and Sri Lankan ancestry cluster in a separate area.
@@
     \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.jpeg}
+    \end{subfigure}
+    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions.
+Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}.
+We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.
+An alternate colouring is provided in fig~\ref{fig:supp_umap_hrs_hisp_admix_alt}.}
+    \label{fig:supp_umap_hrs_hisp_admix}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
+    \end{subfigure}
+    \caption{An alternate colouring of fig~\ref{fig:supp_umap_hrs_hisp_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
+    \label{fig:supp_umap_hrs_hisp_admix_alt}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp_birthonly.jpeg}
+    \end{subfigure}
+    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured region of birth.}
+    \label{fig:supp_umap_hrs_hisp_birth}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.8\textwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256.jpeg}
+    \end{subfigure}
+    \caption{UMAP of the first 8 principal components of the Asian population in the UKBB coloured by self-identified ethnicity. This is an alternate colouring of fig~\ref{fig:umap_ukbb_indian}}
+    \label{fig:supp_umap_ukbb_asian_eth}
+\end{figure*}
+
+\begin{figure*}
+    \centering
     \begin{subfigure}{0.8\textwidth}
@@
     \end{subfigure}
-    \caption{Using country of birth data, some of the larger unidentified groups from figure~\ref{fig:supp_ukbb_zoom} were identified as being born mostly in Japan, the Philippines, North Africa, the Middle East, and Central and South America.
+    \caption{Using country of birth data, some of the larger unidentified groups from fig~\ref{fig:fig_ukbb} were identified as being born mostly in Japan, the Philippines, North Africa, the Middle East, and Central and South America.
 The large cluster of "Any other Asian Background" were mostly born in Sri Lanka.}
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia.
+    \caption{Fig~\ref{fig:umap_world_map}, zoomed in on Asia.
 Geographic distribution of UMAP coordinates.
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean.
+    \caption{Fig~\ref{fig:umap_world_map}, zoomed in on the Caribbean.
 Geographic distribution of UMAP coordinates.
@@
     \end{subfigure}
-    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe.
+    \caption{Fig~\ref{fig:umap_world_map}, zoomed in on Europe.
 Geographic distribution of UMAP coordinates.
@@
 
-%%%%% UKBB phenotypes end
+%%%%% UKBB-UMAP phenotypes end
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/montage_fev1_height_afr_permuted.pdf}
+    \end{subfigure}
+    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) coloured by self-identified ethnic background (left, from fig~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right).
+An arrow points to an area where the FEV1 distribution appears to change, corresponding to where the clusters contain more people with self-identified mixed backgrounds.}
+    \label{fig:supp_comparison_fev_afr}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.png}
+    \end{subfigure}
+    \caption{Zoomed in section of fig~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
+The OEG cluster next to the Chinese cluster appears redder on the middle panel, suggesting higher levels of FEV1.}
+    \label{fig:supp_comparison_fev_chi_eur}
+\end{figure*}
 
@@
     \end{subfigure}
-    \caption{Zoomed in areas of figure~\ref{fig:umap_ukbb_pc10}.
+    \caption{Zoomed in areas of fig~\ref{fig:umap_ukbb_pc10}.
 Sections (i) and (ii) respectively focus on the African and Asian superpopulations, and section (iii) focuses on an area with individuals from many ethnic backgrounds.
@@
     \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_fev1_height_afr_permuted.pdf}
-    \end{subfigure}
-    \caption{Individuals of Black African, Black Caribbean, and mixed backgrounds (primarily White and Black Caribbean/African) coloured by self-identified ethnic background (left, from figure~\ref{fig:umap_ukbb_pc10}), FEV1 (middle), and age-adjusted height (right).
-An arrow points to an area where the FEV1 distribution appears to change, corresponding to where the clusters contain more people with self-identified mixed backgrounds.}
-    \label{fig:supp_comparison_fev_afr}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.png}
-    \end{subfigure}
-    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right).
-The OEG cluster next to the Chinese cluster appears redder on the middle panel, suggesting higher levels of FEV1.}
-    \label{fig:supp_comparison_fev_chi_eur}
-\end{figure*}
-
-\begin{figure*}
-    \centering
     \begin{subfigure}{0.5\textwidth}
@@
     \end{subfigure}
-    \caption{Individuals from the zoomed in section in figure~\ref{fig:supp_comparison_fev_chi_eur} used in statistical testing, coloured the same as in figure~\ref{fig:supp_fev_ridgeplots}.
+    \caption{Individuals from the zoomed in section in fig~\ref{fig:supp_comparison_fev_chi_eur} used in statistical testing, coloured the same as in fig~\ref{fig:supp_fev_ridgeplots}.
 Brown, blue, and green represent those born in the Philippines, Malaysia, and Japan; pink represents those who self-identify as Chinese.
@@
     \caption{Plots of the distributions of residual FEV1 by sex for East Asian populations, after adjusting for height, age, age$^2$, and sex through linear regression.
-Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from figure~\ref{fig:supp_comparison_fev_chi_eur}.
+Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from fig~\ref{fig:supp_comparison_fev_chi_eur}.
 The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.
@@
 %%%%% boxplots end
-%\begin{figure*}
-%    \centering
-%    \begin{subfigure}{\textwidth}
-%    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
-%    \end{subfigure}
-%    \caption{Alternate colouring of figure~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
-%    \label{fig:supp_umap_ukbb_worldmap_alt}
-%\end{figure*}
 

### 2019-08-16 ukbb_asian_plotting.Rmd
manuscript changes, add some R scripts used in analysis and dataviz

---
title: "UKBB plotting Asian populations"
author: 
  - name: Alex Diaz-Papkovich
affiliation:
  - &cruk Quantitative Life Sciences, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
  df_print: paged
code_folding: show
toc: yes
toc_float: 
  collapsed: false
smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import libraries

```{r}
library(tidyverse)
library(ggplot2)
```

## Import the files
```{r}

data_path <- "/Volumes/Stockage/alex/ukbb_other/data/ukbb_asian_data.csv"
ukbb_asian_data <- read.csv(data_path)

```

## Recreate the original image from the jupyter notebook
```{r}

out_dir <- "/Volumes/Stockage/alex/ukbb_images/ggplot_for_export"
fname <- "Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256"
h <- 10
w <- 6.49
s <- 0.5
a <- 0.6

# Colour palette from other projection
#original_pal <- c("#fcc5c0","#3f007d","#f768a1","#e377c2","#49006a","#ae017e")

# New colour palette to better distinguish groups
colour_pal <- c("#fcc5c0","#c7c7c7","#ff7f0e","#e377c2","#49006a","#2ca02c")

p_asian <- ggplot(data = ukbb_asian_data, aes(x=x_coords, y=y_coords, colour=eth_txt)) +
  geom_point(size=s, alpha=a) +
  scale_color_manual(values=colour_pal) +
  theme(
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank(),
        legend.position="bottom", legend.title=element_blank()
        ) +
  guides(colour = guide_legend(override.aes = list(size=3)))
#+
#  guides(colour=FALSE)

p_asian

ggsave(paste(paste(out_dir,fname,sep="/"),".jpeg", sep=""), height=h, width=w, units="in")
```

# Tweak the dataset so that we can use ethnicity x COB
```{r}
ukbb_asian_data$eth_COB <- paste(ukbb_asian_data$eth_txt, "born in", ukbb_asian_data$COB)

eth_cob_counts <- as.data.frame(table(ukbb_asian_data$eth_COB))
eth_cob_counts[order(-eth_cob_counts$Freq),][1:10,]
```

We only want the top few.

```{r}
ukbb_asian_data$eth_COB_2 <- ifelse(ukbb_asian_data$eth_COB=="Indian born in Kenya",
                                    "Indian born in Kenya", "Other")
```

# Now do some more plotting with this new variable
```{r}
out_dir <- "/Volumes/Stockage/alex/ukbb_images/ggplot_for_export"
fname <- "Asthma_FullAsian_PC_QC_new_UMAP_PC5_NC2_NN15_MD05_201931173256_ethcob"
h <- 10
w <- 6.49
s <- 0.5
a <- 0.4

# Colour palette from other projection
colour_pal <- c("#49006a","#e8e8e8")

p_asian <- ggplot(data = ukbb_asian_data, aes(x=x_coords, y=y_coords, colour=eth_COB_2)) +
  geom_point(size=s, alpha=a) +
  scale_color_manual(values=colour_pal) +
  theme(axis.title.y=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank()) +
  guides(colour=FALSE) +
  annotate("text", x=5, y=-25,label="Indians born in Kenya",colour=colour_pal[1])

p_asian

ggsave(paste(paste(out_dir,fname,sep="/"),".jpeg", sep=""), height=h, width=w, units="in")
```

### 2019-08-29 scripts/general_umap_script.py
update UMAP script for users

@@ -1,159 +1,211 @@
 # Python script to carry out UMAP on PC data
 
+import argparse
+from argparse import RawTextHelpFormatter
 import numpy as np
 import logging
 import os
-import sklearn
 import sys
 import time
+import timeit
 
 import umap
 
-dset = input('Specify PC dataset (UKBB/HRS/HRS_HISP/HRS_WHITE/HRS_BLACK/1000G/HRS_1000G): ')
-pcs = [int(n) for n in input('Number of principal components (spaces between list elements, default 10): ').split()]
-nn = int(input('Specify number of neighbours (default 15): '))
-md = float(input('Specify minimum distance (default 0.1): '))
-nc = int(input('Specify number of components (usually 2): '))
+# Desired inputs with argparse
+# -in (filename)
+# -pc (# PCs)
+# -nn 
+# -md
+# -nc
+# -dist
+# -outdir 
 
-# PCA files
-ukbb_dir = '/Volumes/Stockage/alex/ukbb_projections'
-hrs_dir = '/Volumes/Stockage/alex/hrs/projections'
-tg_dir = '/Volumes/Stockage/alex/1000G/projections'
-hrs_1000g_dir = '/Volumes/Stockage/alex/hrs_1000G/projections'
+# define a str2bool function to intake the -head argument
+def str2bool(v):
+    if isinstance(v, bool):
+       return v
+    if v.lower() in ('yes', 'true', 't', 'y', '1'):
+        return True
+    elif v.lower() in ('no', 'false', 'f', 'n', '0'):
+        return False
+    else:
+        raise argparse.ArgumentTypeError('Boolean value expected.')
 
-ukbb_path = os.path.join(ukbb_dir,'ukbb_pca_only')
-hrs_path = os.path.join(hrs_dir,'hrs_200_pc')
-hrs_hisp_path = os.path.join(hrs_dir, 'HRS_PCA_hispanic.eigenvec')
-hrs_white_path = os.path.join(hrs_dir, 'HRS_PCA_white.eigenvec')
-hrs_black_path = os.path.join(hrs_dir, 'HRS_PCA_black.eigenvec')
-hrs_1000g_path = os.path.join(hrs_1000g_dir, 'merged_1000G_HRS_pca.eigenvec')
-tg_path = os.path.join(tg_dir,'pca_1000g_100')
+# example text for usage
+example_text = '''
[…]

### 2019-08-29 scripts/general_umap_script.py
add more documentation

@@ -33,7 +33,9 @@ def str2bool(v):
 
 # example text for usage
 example_text = '''
-Requires the following packages: argparse, numpy, logging, os, sys, time, timeit, umap
+Requires the following packages: argparse, numpy, logging, os, sys, time, timeit, umap.
+File names are generated automatically based on the input file name. Files are time-stamped
+to avoid being overwritten from new runs.
 
 EXAMPLE USAGE
 
@@ -71,7 +73,7 @@ parser.add_argument('-nc', type=int,
     help='Integer. Low dimensional components to project to (default 2D)')
 parser.add_argument('-met', type=str,
     default='euclidean',
-    help='String. Type of distance metric to use (default euclidean)')
+    help='String. Type of distance metric to use (default euclidean). Check UMAP documentation for full list.')
 parser.add_argument('-outdir', type=str,
     help='String. Output directory')
 parser.add_argument('-head', type=str2bool,

### 2019-08-29 ms/main.tex
last few minor changes

@@
 This \emph{population structure} is informative about human history, and can have a significant impact on studies of medical genetics.
-Because population structure depends on myriad demographic and ecological forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure.
+Because population structure depends on myriad demographic, ecological, and social forces, \emph{a priori} visualization is useful to identify subtle patterns of population structure.
 We use a dimension reduction method --- UMAP --- to visualize population structure in three genomic datasets and find previously unobserved patterns, revealing fine-scale population structure and illustrating differences between groups in traits such as white blood cell count, height, and FEV1, a measure used to detect airway obstruction.

### 2019-08-29 ms/main.tex
add some citations
minor tweaks to text, final citations

@@
 For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.
-UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes.
+UMAP is sensitive to sample sizes and spends more visual space on populations with larger sample sizes.
 This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB.
@@
 
-Scripts for all tests and plotting functions can be found on \url{https://github.com/diazale/gt-dimred}.
+Scripts for all tests and plotting functions can be found on \url{https://github.com/diazale/gt-dimred} with a command line script for UMAP available at \url{https://github.com/diazale/gt-dimred/scripts/general_umap_script.py}.
 A demo version using freely available 1KGP data is available at \url{https://github.com/diazale/1KGP_dimred}.
@@
 UMAP was performed using a Python implementation\cite{2018arXivUMAP}.
-Statistical testing was done in SciPy\cite{scipy} and StatsModels\cite{seabold2010statsmodels}.
+Statistical testing was done in SciPy\cite{scipy}, StatsModels\cite{seabold2010statsmodels}, and R\cite{r2013}.
+Visualizations were created with Matplotlib\cite{Hunter2007} and ggplot2\cite{ggplot2016}, and maps were created with rgdal\cite{rgdal2019}.
 

### 2019-08-29 ms/main.tex
minor changes, add some stuff on global structure

@@
 \subsection*{Discussion}
+%By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure.
+Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
+Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.
+UMAP, however, comprehensively illustrates genotypic information at fine scales and within the context of global population structure.
+It is easy to use and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
+It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
+
 By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure.
-Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences including in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
+Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian and t-distributed approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
+Building the global structure in t-SNE requires tuning the \textit{perplexity} parameter, which in practice will need to be different for varying cluster sizes\cite{wattenberg2016how}, while UMAP merges discrete simplicial sets into a coherent global structure.
 Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.
+
 UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure.
-It is easy to use and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
+It is easy to use and fast: given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions.
 It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure.
@@
 \subsubsection*{Caveats}
+In contrast to PCA, UMAP has more adjustable parameters.
 Changing the PC cutoff, minimum distance, and number of neighbours can change characteristics of the visualizations.
@@
 Among the parameters that we varied, the number of PCs used in pre-processing of the data has the largest effect for both methods (see figs~\ref{fig:supp_megamontage_pc2_9} and \ref{fig:supp_megamontage_pc10_50}).
+With UMAP, there are other parameters, such as the learning rate and the distance metric; these were left to the default values.
 

### 2019-09-05 ms/main.tex
remove unnecessary sentence
add some stuff on hierarchical clustering

@@
 Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian and t-distributed approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
-Building the global structure in t-SNE requires tuning the \textit{perplexity} parameter, which in practice will need to be different for varying cluster sizes\cite{wattenberg2016how}, while UMAP merges discrete simplicial sets into a coherent global structure.
 Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.
+Hierarchical clustering of networks has also been successful in detecting fine-scale population structure, though it is instead motivated by partitioning data into subnetworks to then find stable subclusters\cite{han2017clustering}; in practice, one could carry this out by detecting clusters in data already dimensionally reduced using UMAP\cite{howtoUMAP}.
 

### 2019-09-05 ms/main.tex
tweak wording on clustering

--- 2019-08-29 ms/main.tex
@@
 Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian and t-distributed approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
-Building the global structure in t-SNE requires tuning the \textit{perplexity} parameter, which in practice will need to be different for varying cluster sizes\cite{wattenberg2016how}, while UMAP merges discrete simplicial sets into a coherent global structure.
 Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.
+Hierarchical clustering of networks has also been successful in detecting fine-scale population structure, though it was instead motivated by partitioning IBD data into subnetworks to then find stable subclusters\cite{han2017clustering}; in practice, one could carry this out by detecting clusters in data that has been pre-processed using UMAP as laid out in \cite{howtoUMAP}.
 

### 2019-09-07 ms/main.tex
reword pgh

--- 2019-08-29 ms/main.tex
@@
 
-By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure.
+%By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure.
 Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian and t-distributed approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP.
-Building the global structure in t-SNE requires tuning the \textit{perplexity} parameter, which in practice will need to be different for varying cluster sizes\cite{wattenberg2016how}, while UMAP merges discrete simplicial sets into a coherent global structure.
 Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.
+Hierarchical clustering of networks has also been successful in detecting fine-scale population structure, though it was instead motivated by partitioning IBD data into subnetworks to then find stable subclusters\cite{han2017clustering}; in practice, one could carry this out by detecting clusters in data that has been pre-processed using UMAP as laid out in \cite{howtoUMAP}.
+
+Methods such as UMAP and t-SNE focus on preserving local distances to reveal fine-scale structure in populations, and in the process may preserve aspects of global structure as well.
+In contrast, PCA preserves long range distances but hides finer-scale details.
+Hierarchical clustering of networks has also successfully detected fine-scale population structure using identity-by-descent similarity by attempting to preserve relations between global networks and smaller local ones\cite{han2017clustering}.
+We speculate that the addition of weak constraints favouring the preservation of longer distances in UMAP-like approaches has the potential of preserving the desirable local properties while encouraging more intuitive positioning of clusters on a global scale.
 
@@
 Using a minimal number of neighbours (e.g.
-5) can result in the formation of disjoint clusters comprised of related family members (fig~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail.
-We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally perform well across datasets.
+5 rather than the default 15) can result in the formation of disjoint clusters comprised of related family members (fig~\ref{fig:supp_1kgp_families}), and using a low minimum distance (e.g.
+0.001 rather than the default 0.1) can result in clusters becoming more compact, losing visual detail.
+We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP generally perform well across datasets.
 
@@
 This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB.
-We did not assign meaning to wiggles in UMAP figs, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.
+We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data.
 Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

### 2019-09-07 ms/main.tex
add e.g.

@@
 In contrast, PCA preserves long range distances but hides finer-scale details.
-Hierarchical clustering of networks has also successfully detected fine-scale population structure using identity-by-descent similarity by attempting to preserve relations between global networks and smaller local ones\cite{han2017clustering}.
+Hierarchical clustering of networks has also successfully detected fine-scale population structure using identity-by-descent similarity by attempting to preserve relations between global networks and smaller local ones (e.g., \cite{han2017clustering}).
 We speculate that the addition of weak constraints favouring the preservation of longer distances in UMAP-like approaches has the potential of preserving the desirable local properties while encouraging more intuitive positioning of clusters on a global scale.

### 2019-09-11 ms/main.tex
reorder supp figures
fix image
more image fixes

@@
 \begin{subfigure}{.5\columnwidth}
-\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
+\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.jpeg}%
 \caption{}%
@@
     \begin{subfigure}{\columnwidth}
-    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
-    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
+    %\includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
+    %\includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
+    \includegraphics[width=0.8\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_geo_permuted_10nn_sd50000_base50000_201883118582.pdf}
     \caption{}
@@
      \begin{subfigure}{\columnwidth}
-         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
-    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
+         %\includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
+    %\includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
+    \includegraphics[width=.8\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_height_pct1_f_2018714161841_leukocyte_count_pct_5_f_201871416519.jpeg}
    \caption{}
@@
 
-\begin{figure}
-\centering
-   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix132.jpeg}
-   \caption{An alternate colouring of \ref{fig:umap_hrs_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
-    \label{fig:umap_hrs_admix_alt}
-\end{figure}
-
 \begin{figure*}
@@
     \label{fig:supp_hrs_born}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
-    \end{subfigure}
-    \caption{UMAP projection of the top 10 principal components of the combined HRS and 1KGP datasets.
-One cluster (in the box) does not group with any of the 1KGP populations.
-A cluster of Finnish (FIN) individuals consistently appears in the "White Not Hispanic" (WNH) group.
-Groups of Central and South American populations from the 1KGP (CLM, Colombian; MXL, Mexican; PEL, Peruvian; PUR, Puerto Rican) form nearby or within the HRS Hispanic cluster (HIS).
-Iberian individuals (IBS) cluster near the Hispanic population.
-Toscani individuals (TSI) form some small clusters and sometimes appear near the Iberian and Hispanic populations.
-Individuals with British/Scottish (GBR) or Northern/Western European ancestry (CEU) are scattered throughout the WNH clusters.
-Individuals with African ancestry from the 1KGP group with Black Americans from the HRS (AFR).
-Similar population groupings occur with South Asian (SAS) and East Asian (EAS) individuals.}
-    \label{fig:supp_hrs_1000g}
 \end{figure*}
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
-    \end{subfigure}
-    \caption{An alternate colouring of fig~\ref{fig:supp_umap_hrs_hisp_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
-    \label{fig:supp_umap_hrs_hisp_admix_alt}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
     \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp_birthonly.jpeg}
@@
     \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_height_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_height_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_fev_f}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
+    \end{subfigure}
+    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
+    \label{fig:supp_box_fev_m}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
     \includegraphics[width=\textwidth]{images/montage_fev1_height_afr_permuted.pdf}
@@
 \end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.5\textwidth}
+    \includegraphics[width=0.5\textwidth]{images/east_asia_population_colours.png}
+    \end{subfigure}
+    \caption{Individuals from the zoomed in section in fig~\ref{fig:supp_comparison_fev_chi_eur} used in statistical testing, coloured the same as in fig~\ref{fig:supp_fev_ridgeplots}.
+Brown, blue, and green represent those born in the Philippines, Malaysia, and Japan; pink represents those who self-identify as Chinese.
+The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.}
+    \label{fig:supp_fev_test_pops}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/east_asia_ridgeplots_fev.png}
+    \end{subfigure}
+    \caption{Plots of the distributions of residual FEV1 by sex for East Asian populations, after adjusting for height, age, age$^2$, and sex through linear regression.
+Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from fig~\ref{fig:supp_comparison_fev_chi_eur}.
+The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.
+Asterisks indicate significant difference from the Japanese population, using Welch's unpaired t-test with a Bonferroni correction for 3 groups.
+The dashed lines are the means of the distributions, and Japanese populations have consistently higher means.}
+    \label{fig:supp_fev_ridgeplots}
+\end{figure*}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_ukbb.jpeg}
+    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
+Done on the UKBB dataset with 20000 iterations.
+The UMAP-initialized graph has been shifted by 230 iterations to approximate the 230 epochs UMAP uses for large datasets ($n>10,000$).}
+    \label{fig:supp_tsne_umap_compare_ukbb_graph}
+\end{figure}
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.95\columnwidth]{images/ukbb_tsne_umap.jpeg}
+    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the UKBB. t-SNE used 20000 iterations}
+    \label{fig:supp_tsne_umap_compare_ukbb}
+\end{figure}
 
@@
 
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.5\textwidth}
-    \includegraphics[width=0.5\textwidth]{images/east_asia_population_colours.png}
-    \end{subfigure}
-    \caption{Individuals from the zoomed in section in fig~\ref{fig:supp_comparison_fev_chi_eur} used in statistical testing, coloured the same as in fig~\ref{fig:supp_fev_ridgeplots}.
-Brown, blue, and green represent those born in the Philippines, Malaysia, and Japan; pink represents those who self-identify as Chinese.
-The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.}
-    \label{fig:supp_fev_test_pops}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/east_asia_ridgeplots_fev.png}
-    \end{subfigure}
-    \caption{Plots of the distributions of residual FEV1 by sex for East Asian populations, after adjusting for height, age, age$^2$, and sex through linear regression.
-Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from fig~\ref{fig:supp_comparison_fev_chi_eur}.
-The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.
-Asterisks indicate significant difference from the Japanese population, using Welch's unpaired t-test with a Bonferroni correction for 3 groups.
-The dashed lines are the means of the distributions, and Japanese populations have consistently higher means.}
-    \label{fig:supp_fev_ridgeplots}
-\end{figure*}
-
 \begin{figure}[!htb]
@@
     \centering
-    \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_ukbb.jpeg}
-    \caption{Comparing the error terms of standard t-SNE versus t-SNE initialized with a UMAP embedding and no early exaggeration.
-Done on the UKBB dataset with 20000 iterations.
-The UMAP-initialized graph has been shifted by 230 iterations to approximate the 230 epochs UMAP uses for large datasets ($n>10,000$).}
-    \label{fig:supp_tsne_umap_compare_ukbb_graph}
-\end{figure}
-
-\begin{figure}[!htb]
-    \centering
-    \includegraphics[width=0.95\columnwidth]{images/ukbb_tsne_umap.jpeg}
-    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the UKBB. t-SNE used 20000 iterations}
-    \label{fig:supp_tsne_umap_compare_ukbb}
-\end{figure}
-
-\begin{figure}[!htb]
-    \centering
     \includegraphics[width=0.95\columnwidth]{images/tsne_umap_graph_1kgp.jpeg}
@@
     \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_fev_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_fev_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_fev_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{FEV1 by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_fev_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/female_height_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_height_f}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
-    \includegraphics[width=\textwidth]{images/male_height_boxplot_annotated.pdf}
-    \end{subfigure}
-    \caption{Height by sex and ethnic group, annotated with p-values. Asterisks indicate significant difference from the White British group with a Bonferroni correction for 12 groups.}
-    \label{fig:supp_box_height_m}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}{\textwidth}
     \includegraphics[width=\textwidth]{images/female_leukocyte_boxplot_annotated.pdf}
@@
 %%%%% boxplots end
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=\textwidth]{images/HRS_1000G_UMAP_PC10_NC2_NN15_MD05_2018627203416_label.pdf}
+    \end{subfigure}
+    \caption{UMAP projection of the top 10 principal components of the combined HRS and 1KGP datasets.
+One cluster (in the box) does not group with any of the 1KGP populations.
+A cluster of Finnish (FIN) individuals consistently appears in the "White Not Hispanic" (WNH) group.
+Groups of Central and South American populations from the 1KGP (CLM, Colombian; MXL, Mexican; PEL, Peruvian; PUR, Puerto Rican) form nearby or within the HRS Hispanic cluster (HIS).
+Iberian individuals (IBS) cluster near the Hispanic population.
+Toscani individuals (TSI) form some small clusters and sometimes appear near the Iberian and Hispanic populations.
+Individuals with British/Scottish (GBR) or Northern/Western European ancestry (CEU) are scattered throughout the WNH clusters.
+Individuals with African ancestry from the 1KGP group with Black Americans from the HRS (AFR).
+Similar population groupings occur with South Asian (SAS) and East Asian (EAS) individuals.}
+    \label{fig:supp_hrs_1000g}
+\end{figure*}
+
+\begin{figure}
+\centering
+   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix132.jpeg}
+   \caption{An alternate colouring of \ref{fig:umap_hrs_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
+    \label{fig:umap_hrs_admix_alt}
+\end{figure}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}{\textwidth}
+    \includegraphics[width=0.7\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix132_hisp.jpeg}
+    \end{subfigure}
+    \caption{An alternate colouring of fig~\ref{fig:supp_umap_hrs_hisp_admix}. Here red, green, and blue correspond to African, Asian/Native American, and European ancestry, respectively.}
+    \label{fig:supp_umap_hrs_hisp_admix_alt}
+\end{figure*}
 
